\section{Bootstrap}
\label{sec:bootstrap}

Initializing a GFM actor requires discovering the capability poset from
scratch. The poset and the actor's ability to reason about $\volL(G)$ are
coupled: the actor cannot evaluate actions without a poset, and the poset
is populated through the actor's interactions. This section describes a
four-phase bootstrap procedure that uses both communication and observation
channels.

\subsection{Phase 1: Listen and Observe}

The actor begins with an empty poset ($\C = \{\bot\}$) and low self-trust
$T_s$. It receives capability reports through the communication channel and
observes agent behavior through the observation channel. Each report or
observation that introduces a benchmarked capability adds a node to $\C$ with
weight $w(d) = \log_2(1 + \smax(d))$. All claims are initially weighted by
low trust $T_k$ (the cooling period from the foundational paper).

\subsection{Phase 2: Cross-Validate}

As reports accumulate, the actor compares the two channels. Agents whose
communicated capabilities match observed behavior build trust quickly
(consistent pattern from Section~\ref{sec:dual_channel}). Overclaiming agents
lose trust. Underclaiming agents trigger model expansion: the actor adds
observed capabilities that the agent did not report. The cross-validation phase
improves the quality of $\weff(d, k)$ and calibrates $T_k$ across the
population.

\subsection{Phase 3: Discover Structure}

The actor discovers subsumption relationships through two mechanisms:

\emph{Logical analysis.} If capability descriptions in $L$ have compositional
structure, the actor can infer subsumption from the descriptions themselves.
``Can run a marathon in under 4 hours'' logically implies ``can run a mile.''
The shared language $L$ must be expressive enough for this inference; formal
specification languages or capability ontologies support it directly.

\emph{Empirical observation.} If every agent observed to have capability $d_1$
also demonstrates $d_2$, the actor infers $d_2 \leq d_1$ (with confidence
proportional to the number of confirming observations and the diversity of
agents observed). This is inductive: a sufficiently large and diverse
observation set makes the inference reliable.

As subsumption edges are added, the actor recomputes independent weights
$\iw(d)$ via full M\"{o}bius inversion ($O(|\C|^3)$ for general posets,
$O(|\C| \cdot k)$ for trees; Proposition~\ref{prop:tractability}).

\subsection{Phase 4: Expand}

With a populated poset and calibrated trust, the actor begins proposing
actions and evaluating them via $\Deltahat(\pi)$
(Equation~\eqref{eq:weighted_estimator}). Self-trust $T_s$ increases as the
actor's predictions ($\Deltahat$) match observed outcomes ($\Delta \volL$
computed post-action). Both channels continue refining the poset throughout
the actor's operation: new capabilities are discovered, benchmark scores are
updated, and subsumption relationships are confirmed or revised.

\subsection{Convergence}

\begin{proposition}[Poset Growth]
\label{prop:bootstrap_convergence}
Under assumptions of positive communication frequency and non-zero observation
    coverage:
\begin{enumerate}
\item[\textbf{(a)}] The poset node set grows monotonically: $\C(t) \subseteq
    \C(t+1)$ for all $t$.
\item[\textbf{(b)}] Adding new capabilities to $G$ (without changing
    subsumption structure) is $\volL$-non-decreasing.
\item[\textbf{(c)}] Discovering subsumption edges can \emph{decrease}
    $\volL(G)$: if $d_1$ and $d_2$ were treated as unrelated and the edge
    $d_1 \leq d_2$ is later discovered, the downward closure $\dc G$ does not
    change but the M\"{o}bius-inverted weights redistribute, and $\volL(G)$
    may drop from $w(d_1) + w(d_2)$ to $w(d_2)$.
\item[\textbf{(d)}] Updating a benchmark to a finer scale ($\smax' > \smax$)
    increases $\volL$ if $d$ is maximal in $\dc G$, leaves it unchanged if $d$
    has exactly one ancestor in $\dc G$, and can \emph{decrease} $\volL$ if $d$
    is a shared prerequisite of multiple elements in $\dc G$.
\end{enumerate}
\end{proposition}

\begin{proof}
(a) Nodes and edges are never removed from $\C$.

(b) If $G(t+1) = G(t) \cup \{d\}$ for a new capability $d$ with no new
subsumption edges, then $\dc G(t+1) \supseteq \dc G(t)$ and all independent
weights are unchanged. By M3, $\volL(G(t+1)) \geq \volL(G(t))$.

(c) Counterexample: let $d_1$ = ``write a function'' ($w = 6.66$) and $d_2$ =
``build a web app'' ($w = 9.97$). Before discovering $d_1 \leq d_2$, the
capabilities are poset-disjoint: $\volL(\{d_1, d_2\}) = 6.66 + 9.97 =
16.63$. After adding the edge, M\"{o}bius inversion gives $\iw(d_1) = 6.66$,
$\iw(d_2) = 9.97 - 6.66 = 3.31$, and $\volL(\{d_1, d_2\}) = 6.66 + 3.31 =
9.97 = w(d_2)$. The volume decreased by $6.66$. This is epistemically correct:
the actor previously overcounted because it did not know that $d_1$'s
measurement content was already captured by $d_2$. Structural refinement
corrects the overcount.

(d) Increasing $\smax(d)$ increases $w(d)$. The effect on $\volL(G)$ depends on
$d$'s position in $\dc G$. Write $\volL(G) = \sum_{d'' \in \dc G} \sum_{d'
\leq d''} \mu(d', d'') \cdot w(d')$ and rearrange by $d'$: the coefficient of
$w(d)$ in $\volL(G)$ is $c(d) = \sum_{d'' \geq d,\, d'' \in \dc G} \mu(d,
d'')$. If $d$ is maximal in $\dc G$, $c(d) = \mu(d, d) = 1$, so $\volL$
increases. If $d$ has exactly one ancestor in $\dc G$, $c(d) = 1 + (-1) = 0$,
so $\volL$ is unchanged. If $d$ is a shared prerequisite with $m > 1$ ancestors
in $\dc G$, the M\"{o}bius sum can produce $c(d) < 0$: for example, in a
V-shaped poset $d < y, d < z$ with $G = \{y, z\}$, $c(d) = \mu(d, d) +
\mu(d, y) + \mu(d, z) = 1 - 1 - 1 = -1$, so increasing $w(d)$
\emph{decreases} $\volL(G)$ (Appendix~\ref{app:proof_bootstrap}).
\end{proof}

\paragraph{Interpretation.} Part (b) shows that discovering new capabilities
always increases $\volL$. Part (d) shows that benchmark refinement is
position-dependent: it increases $\volL$ for maximal capabilities, is neutral
for capabilities with a single ancestor, and can \emph{decrease} $\volL$ for
shared prerequisites of multiple elements in the closure. Part (c) shows that
discovering structure (subsumption) can also decrease $\volL$ by correcting
overcounting. The net trajectory of $\volL$ during bootstrap depends on the
balance between capability discovery (expanding) and structural refinement
(potentially contracting). In practice, early bootstrap is dominated by (b)
(rapid discovery of new capabilities), while later phases are dominated by (c)
(structural refinement that tightens the measure). The long-run $\volL$
converges toward the correct volume given the actor's full structural
knowledge, which may be lower than the naive estimate from the early
structureless phase.

\paragraph{Convergence bounds.} Let $N$ be the total number of capabilities in
the population (unknown to the actor), $E$ the total number of subsumption
edges, and suppose the actor discovers capabilities at rate $r_c$ and edges at
rate $r_e$ per time step. The poset volume trajectory can be bounded:

\emph{Upper bound (structureless).} Before any edges are discovered, each
capability is an atom with $\iw(d) = w(d)$. The structureless volume is
$\volL^{\max}(t) = \sum_{d \in \C(t)} w(d)$, which grows monotonically with
discovery. After $t$ steps with $\min(r_c t, N)$ capabilities discovered, this
is at most $\sum_{d \in \C} w(d)$.

\emph{Lower bound (full structure).} With all edges discovered, M\"{o}bius
inversion redistributes weight and the volume is
$\volL^*(G) = \sum_{d \in \dc G} \iw(d) \leq \volL^{\max}$. The maximum
contraction from structural discovery is bounded by:
\begin{equation}
\label{eq:max_contraction}
\volL^{\max} - \volL^* \leq \sum_{d' < d} |w(d')|
\end{equation}
summing over all comparable pairs, which is finite for finite posets.

\emph{Convergence trajectory.} The volume trajectory $\volL(t)$ rises during
early capability discovery (Phase~1--2), then tightens toward $\volL^*$ as
structural refinement progresses (Phase~3). The time to convergence is bounded
by $\tau_c = N/r_c + E/r_e$: the time to discover all capabilities plus the
time to discover all edges. Under adversarial conditions (a scorpion
withholding information about its capabilities or providing false subsumption
claims), $r_c$ and $r_e$ degrade, and convergence may stall for capabilities
and edges the adversary controls. The dual-channel design partially mitigates
this: the observation channel discovers capabilities and infers edges
independently of the communication channel, so a silent adversary slows
convergence but does not halt it as long as its capabilities are observable.

\paragraph{Binary default and deferred structure.} Every capability that
enters the poset through communication or observation immediately receives
the binary default weight $w(d) = 1$ bit ($\smax = 1$): the actor knows the
capability exists. The capability enters as an \emph{unstructured atom} with no
subsumption edges, contributing $\iw(d) = w(d) = 1$ to $\volL(G)$ immediately.

Subsumption edges are added in a separate step (Phase~3) only after the
non-negativity check passes: for a proposed edge set making $d$ a multi-cover
node, the actor verifies $\iw(d) \geq 0$ under M\"{o}bius inversion before
committing the edges. If the check fails (e.g., a diamond node with two
binary-default prerequisites gives $\iw = 1 - 1 - 1 < 0$), the edges are
deferred until the capability's benchmark is refined to a level where the check
passes. The capability continues contributing as a poset-disjoint atom in the
interim.

This sequencing ensures two properties simultaneously: (1) every discovered
capability contributes to $\volL(G)$ from the moment of discovery, and (2)
$\iw(d) \geq 0$ is maintained for all nodes at all times. Developing a finer
benchmark ($\smax > 1$) increases $w(d)$ beyond the binary default, eventually
enabling deferred edges to be added. The remaining zero-weight gap consists
only of capabilities the actor has not yet learned about through either
channel.

\subsection{Adversarial Bootstrap}
\label{sec:adversarial_bootstrap}

The foundational paper's scorpion taxonomy \citep{lasser2026gfm} classifies
agents that persistently contract $\vol(G)$. During bootstrap, scorpions
interact with each phase differently:

\paragraph{Phase 1 (Listen and Observe).} A scorpion can overclaim
capabilities to inflate the poset with nodes the population does not
actually possess. These nodes enter at the binary default ($w = 1$) and
inflate $\volL(G)$. \emph{Defense:} the cooling period assigns low $T_k$ to
new agents' claims, limiting the inflation to $1 \cdot T_k \approx 0$ bits
per overclaimed capability until trust builds. The observation channel
independently verifies or falsifies claims during Phase~2.

\paragraph{Phase 2 (Cross-Validate).} Overclaiming scorpions are caught by
the divergence signal ($\wcomm \gg \wobs$), driving $T_k$ downward. A
scorpion that conceals capabilities (underclaiming) does not harm the poset;
it merely slows discovery. The cross-validation phase is where the
dual-channel design provides its strongest adversarial defense.

\paragraph{Phase 3 (Discover Structure).} This is the most vulnerable phase.
A scorpion could assert false subsumption relationships through the
communication channel, causing the actor to add incorrect edges. Incorrect
edges redistribute independent weights through M\"{o}bius inversion and may
be difficult to undo (Section~\ref{sec:discussion}). \emph{Defense:} the
deterministic insertion policy (Section~\ref{sec:poset}) requires the
non-negativity check before committing edges, which catches some false claims
(a claimed subsumption between capabilities that would produce $\iw < 0$).
The empirical observation mechanism of Phase~3 provides independent
verification: if every observed agent with $d_1$ also demonstrates $d_2$, the
edge is confirmed regardless of what the scorpion claims. Edges inferred
solely from a single agent's communication and not confirmed by observation
should be treated with the same trust discount as the agent's other claims.

\paragraph{Phase 4 (Expand).} Once the poset is populated and the
optimization loop begins, scorpion detection proceeds through the standard
channels of the foundational paper (Proposition~2 of \citet{lasser2026gfm}):
contraction attribution and deception detection. The poset construction
does not introduce new scorpion vulnerabilities in this phase; it inherits the
foundational paper's detection capabilities and limitations.
