\section{Conclusion}
\label{sec:conclusion}

The capability poset provides the first tractable instantiation of
goal-frontier maximization. By constructing the capability space as a finite
weighted poset discovered through dual-channel sensing, the poset resolves
the structure-of-$\G$ problem that the foundational paper identified as its
most urgent open question. The poset measure $\volL$, defined via M\"{o}bius
inversion, satisfies the axioms that unlock all of the foundational paper's
structural results: self-balancing, scorpion detection, and cooperative
expansion hold on finite posets exactly as they do on abstract capability
spaces.

The construction also advances the estimator and introduces leverage. On the
poset, agent reports carry real-valued weighted changes rather than ternary
signs, resolving the foundational paper's magnitude-estimation limitation and
strengthening the sign-correctness conditions for mixed-sign capability
changes. Leverage, defined as the marginal contribution of a capability to
$\volL$ growth rate, acts as a ranking signal that guides action selection
toward capability-multiplying partnerships: cooperation between high-leverage
agents is disproportionately rewarded, creating a natural gradient toward
civilization-scale infrastructure (education, research, communication)
derived entirely from the $\volL$ objective.

The bootstrap procedure shows how a GFM actor can discover the poset
incrementally through communication and observation, with convergence bounds
and adversarial analysis for each phase. The binary default ensures every
discovered capability contributes immediately; deferred structure maintains
non-negativity at all times.

The poset is bounded by what can be communicated and observed. Every
discovered capability carries at least binary weight (1 bit); graded benchmarks
and leverage amplification refine this upward. The B-to-C tension partially self-corrects: the log weight function creates
diminishing returns on measurement precision, making maximal-position
binary-default capabilities the best targets for benchmark refinement. The
leverage ranking signal further directs investment toward high-leverage
binary-default capabilities (disproportionately the relational and
experiential ones). The residual gap narrows to capabilities whose value is
purely intrinsic: neither instrumental to $\volL$ growth nor expressible on any
graded scale.
