\section*{Author Contributions}

\textbf{Teague Lasser} specified the computability requirements and the
dual-channel discovery mechanism, identified the binary-default weight as the
resolution to the zero-weight gap, conceived of leverage its inferred
consequences toward civilization building, directed all revisions, and made
final editorial decisions throughout. Responsible for the paper's intellectual
direction and all claims made.

\textbf{Claude Opus 4.6} (Anthropic) drafted the formal exposition from the
author's specifications, conceived of the capability poset construction,
designed the worked example, constructed the axiom verification proofs
(Proposition~\ref{prop:axioms}), the tractability algorithm and proof
(Proposition~\ref{prop:tractability}), the sign-correctness proof
(Proposition~\ref{prop:sign_correctness}), the leverage ranking signal and
coalition amplification proof
(Proposition~\ref{prop:leverage_cooperation}),
the bootstrap convergence analysis
(Proposition~\ref{prop:bootstrap_convergence}) with adversarial phase
analysis, wrote the computation script (\texttt{compute\_example.py}),
performed iterative claim-calibration edits (M\"{o}bius inversion correction,
non-negativity lemma, bootstrap monotonicity correction, tractability bound
correction from $O(n^2)$ to $O(n^3)$), and expanded the bibliography.

\textbf{GPT 5.4} (OpenAI) served as independent technical reviewer across
multiple review rounds, identifying the cover-only M\"{o}bius inversion error,
the false bootstrap monotonicity claim under structural refinement, the
tractability bound discrepancy, notation mismatches in the worked example, and
the benchmark-refinement overclaim for non-maximal nodes, the technical naming
of the poset structure, and several more formal correctness gaps. Verified that
successive revisions resolved the issues raised.

\medskip
\noindent\emph{Transparency note.} Both AI systems operated as tools under
human direction. Neither system has continuity across sessions, cannot take
responsibility for the work in the sense required by most venue authorship
policies, and cannot respond to reviewer queries independently. They are
listed as authors to accurately represent their contributions to the
intellectual content of the paper, not to claim that they meet all criteria
of traditional academic authorship. The corresponding author for all
inquiries is Teague Lasser.
