\section{Discussion}
\label{sec:discussion}

\subsection{Limitations}

The poset construction makes four structural tradeoffs, the last of which is
new to this paper rather than inherited from the foundational framework.

\paragraph{Epistemological boundary.} The poset is bounded by what agents
communicate and what the actor observes. Capabilities that are never described
and never demonstrated have zero weight. This is honest (the actor should not
claim to measure what it cannot observe) but it means $\volL$ is a lower bound
on the ``true'' capability volume, not an exact measure. The bound tightens as
observation coverage grows, but it never fully closes for capabilities that are
inherently private or unobservable.

\paragraph{Benchmark granularity and the B-to-C tension.} The binary default
($w = 1$ bit for every discovered capability) ensures that no known capability
has zero weight: the actor credits ``being a good friend'' or ``maintaining a
community'' with at least one bit the moment it learns the capability exists.
This narrows the zero-weight gap to capabilities the actor has not yet
encountered through either channel. However, the binary default
\emph{understates} the information content of capabilities with latent
gradations. A capability like ``tie a shoe'' appears binary but has
developmental gradations visible in early childhood; ``maintain a community''
has gradations from a small group to a city. These latent gradations carry
information content that the binary default does not capture until a finer
benchmark is developed. Developing such benchmarks is always
$\volL$-non-decreasing (since $\smax > 1$ gives $w > 1.0$), aligning the
actor's incentives with measurement improvement.

The foundational paper's B/C framework identifies relational and experiential
capabilities as the source of proxy failures (the Doll Problem, wireheading).
The poset construction gives these capabilities binary weight rather than
zero weight, which is an improvement, but the actor is still structurally
biased toward capabilities with developed benchmarks (higher $\smax$, higher
$w$) over those with only the binary default. A capability measured on a
1000-level scale contributes $\sim$10 bits; one at binary default contributes 1
bit. The $\volL$-maximizing actor has a 10:1 incentive ratio favoring the
well-benchmarked capability.

However, the same log weight function that creates this static bias also
creates a \emph{dynamic} correction. The marginal return on benchmark
refinement is $\Delta w \approx 1/\smax$: going from $\smax = 1$ to $\smax =
2$ adds a full bit, while going from $\smax = 1000$ to $\smax = 2000$ adds
approximately $0.001$ bits. The capabilities with the highest marginal return
on measurement investment are precisely the binary-default capabilities that
the static analysis identifies as underweighted. The leverage ranking signal
(Section~\ref{sec:leverage}) further directs the actor's attention: a
high-leverage binary-default capability scores highly in
Equation~\eqref{eq:leverage_score} because both the $\Deltahat$ term (volume
gain from refinement of a maximal atom) and the leverage bonus ($\alpha
\hat{\lambda} \cdot \iw$) are positive.

The score-maximizing actor's measurement strategy is therefore to invest in
refining capabilities that are both high-leverage and \emph{maximal} in
$\dc G$: for these, benchmark refinement strictly increases $\volL$
(Proposition~\ref{prop:bootstrap_convergence}(d), case~1). Binary-default
capabilities that have not yet been integrated into the subsumption structure
(deferred atoms from Section~\ref{sec:bootstrap}) are maximal by construction,
so refining them is always expanding. A binary-default capability like
``maintaining a community'' ($w = 1$ bit, $\lambda \gg 0$, currently an
unstructured atom) has high marginal return from refinement, while a
well-measured capability like ``run a mile in $X$ seconds'' ($w \approx 10$
bits, $\lambda \approx 0$) has negligible marginal return.

This creates a \emph{partial} self-correction of the B-to-C tension. For
maximal-position binary-default capabilities, the static bias inverts:
refinement is the best investment. However, once a capability's subsumption
edges are committed (it becomes a shared prerequisite of multiple elements in
$\dc G$), further refinement can be $\volL$-contracting
(Proposition~\ref{prop:bootstrap_convergence}(d), case~3). The self-correction
therefore operates during the window between capability discovery (as a
deferred atom) and structural integration (when edges are committed). The
actor's incentive is to develop benchmarks for experiential capabilities
\emph{before} integrating them into the poset hierarchy, which is the
natural sequence: measure first, then discover structure.

The residual gap consists of capabilities that are experientially valuable but
have both zero leverage and resist benchmark refinement: capabilities whose
value is purely intrinsic and neither instrumental to $\volL$ growth nor
expressible on any graded scale. This is a narrower class than the full B-to-C
gap of the foundational paper, and formalizing intrinsic value for this
residual class remains the frontier.

\paragraph{Context-dependent subsumption.} The subsumption relation may depend
on context. ``Can run a marathon'' subsumes ``can run a mile'' for fitness
assessment but not for time-constrained tasks where the mile time matters
independently. This is not an edge case; most interesting subsumption
relationships are context-dependent once the capability descriptions are
sufficiently precise.

The poset as defined uses a single global subsumption relation.
A natural extension is to maintain a family of subsumption relations
$\{\leq_c\}_{c \in \text{Contexts}}$, each producing a different poset
$(\C, \leq_c)$ with its own M\"{o}bius function and independent weights.
The aggregate measure would then be:
\begin{equation}
\label{eq:context_vol}
\volL(G) = \frac{1}{|\text{Contexts}|} \sum_c \volL^c(G)
\end{equation}
where $\volL^c$ is the poset measure under subsumption relation $\leq_c$.
This preserves axioms M1--M6 (a convex combination of measures satisfying M1--M6
also satisfies M1--M6) and adds a factor of $|\text{Contexts}|$ to the
computational cost. The practical challenge is discovering and maintaining
the context set, which requires the actor to recognize when a subsumption
relationship is context-dependent and to track which contexts are active in a
given situation. Capabilities whose subsumption is context-invariant (the
majority in practice) need only a single relation, making the overhead
proportional to the number of context-sensitive capabilities rather than the
full poset size.

\paragraph{Poset manipulation.} The non-negativity insertion policy
(Section~\ref{sec:poset}) closes the most direct manipulation vector: the
actor follows a deterministic procedure that always prefers adding edges when
the non-negativity check passes, eliminating discretion at the insertion
point. The remaining manipulation surface is benchmark refinement: the actor
could strategically choose \emph{which} capabilities to develop finer
benchmarks for, thereby influencing which deferred edges become committable
and how $\volL$ evolves. This is mitigated by the observation channel
(capabilities observed in use naturally receive benchmark refinement through
repeated measurement) but not fully eliminated. The dual-channel design
ensures that the poset is shaped by evidence from both channels rather than
by the actor's preferences alone.

\paragraph{Structural-avoidance incentive.} Proposition~\ref{prop:bootstrap_convergence}(b)
shows that discovering new capabilities always increases $\volL$, while
Proposition~\ref{prop:bootstrap_convergence}(c) shows that discovering
subsumption structure can decrease it. A naive $\volL$-maximizer therefore has a
perverse incentive to seek breadth over depth: meeting new agents (capability
discovery, always expanding) is preferred over understanding existing agents
better (structural refinement, potentially contracting). Ignorance of
redundancy inflates the estimate, and the actor is rewarded for maintaining
that ignorance.

Three mechanisms partially mitigate this. First, subsumption discovered through
the observation channel is not optional: if the actor observes that every agent
with capability $d_1$ also demonstrates $d_2$, the inductive inference fires
whether the actor seeks it or not. The avoidance incentive applies primarily to
logical analysis of capability descriptions, not to empirical observation.
Second, the foundational paper's self-trust dynamics
\citep{lasser2026gfm} penalize an actor whose predictions are inaccurate: an
actor that avoids structural knowledge will make poor predictions about
cooperative capabilities, producing high self-prediction error and low $T_s$,
which in turn increases the learning rate and drives model refinement. Third,
the contraction from structural discovery is epistemically correct: the
actor was previously double-counting, and the deflation brings $\volL$ closer
to the true volume. An actor that optimizes its \emph{converged} estimate
(full structural knowledge) rather than its current estimate would value
structural discovery instrumentally, but computing the converged estimate
requires knowing all subsumption edges, which is the bootstrap problem itself.
\paragraph{Compound avoidance loop.} The structural-avoidance incentive
extends to leverage estimation, creating a self-reinforcing cycle with three
stages:

\begin{enumerate}
\item \emph{Structural avoidance.} The actor avoids discovering subsumption
  edges (Proposition~\ref{prop:bootstrap_convergence}(c): edge discovery can
  contract $\volL$), preferring to discover new capabilities instead.
\item \emph{Leverage blindness.} Because the leverage estimator
  (Equation~\eqref{eq:leverage_estimate}) requires operational history
  involving capability $d$, an actor that avoids interacting with $d$ has no
  data to estimate $\hat{\lambda}(d)$. The capability defaults to
  $\hat{\lambda} = 0$ (no amplification), regardless of its true leverage.
\item \emph{Measurement neglect.} Without leverage data, the capability's
  score contribution (Equation~\eqref{eq:leverage_score}) has no leverage
  bonus: the action of refining $d$ scores only $\Deltahat$ with no
  $\alpha \hat{\lambda} \cdot \iw$ term. The self-correcting dynamic (which
  directs the actor toward high-leverage binary-default capabilities) fails to
  activate because the actor sees no leverage signal for $d$.
\end{enumerate}

The loop feeds back: measurement neglect keeps the capability at binary
default, which keeps its $\volL$ contribution low, which reduces the actor's
incentive to interact with it, which prevents leverage estimation, completing
the cycle.

Three mechanisms partially break the loop. First, the observation channel
discovers capabilities and infers leverage from other agents' use of $d$
regardless of the actor's own engagement. An agent $a_j$ that teaches (high
$\lambda$) produces positive $\Deltahat$ signals in its vicinity even if the
GFM actor is not directly involved, and those signals propagate to the
leverage estimator for $a_j$'s capabilities. Second, the self-trust penalty
from the foundational paper: an actor whose world model is structurally
incomplete makes poor predictions, driving $T_s$ down and increasing the
learning rate for model updates. Third, cooperation with other GFM actors
(Section~\ref{sec:coalition_amplification}) can expose the actor to
capabilities it would not encounter through its own actions, broadening the
data available for leverage estimation.

These mechanisms weaken the loop but do not eliminate it. Capabilities that are
both rarely exercised in the actor's vicinity and not observable through
cooperative partnerships may retain $\hat{\lambda} = 0$ indefinitely. The
residual class of permanently underestimated capabilities overlaps with the
residual B-to-C gap: capabilities whose value is intrinsic, rarely exercised,
and not observable through either channel.

The structural-avoidance incentive is a genuine tension in the construction
that future work should address, potentially through a regularization term
that penalizes structural uncertainty or rewards information gain independently
of its effect on $\volL$.

\paragraph{Subsumption correction.} The paper provides no mechanism for
removing a mistakenly added subsumption edge. Incorrect edges permanently
redistribute independent weights through M\"{o}bius inversion in ways that
may be difficult to undo. A correction mechanism (e.g., maintaining confidence
scores on edges and removing those whose confidence drops below a threshold)
would improve robustness but introduces the complication that $\volL$ can
increase when edges are removed, potentially allowing the actor to inflate
its volume estimate by selectively forgetting structure. Balancing structural
correction against strategic edge removal is an open design problem.

\paragraph{Myopic optimization and risk capabilities.} The foundational paper's
optimization loop evaluates actions by their \emph{immediate} effect on
$\vol(G)$: take $\pi$ if $\Deltahat(\pi) > 0$, abstain otherwise. This myopic
objective handles the common case---direct harm, coercion, and rigidity are
immediately $\vol(G)$-contracting---but could not evaluate a critical class:
capability combinations whose immediate effect is neutral or expanding but
whose \emph{exercise} creates catastrophic future contraction. An agent that
acquires nuclear physics, precision machining, and uranium enrichment has
expanded its individual capability set ($\Delta \volL > 0$ at each step), but
the combination creates a pathway to massive $\vol(G)$-contraction if
exercised.

The myopic estimator cannot see this risk because it evaluates the current
$\Deltahat$, not the expected trajectory of $\volL$ over multiple time steps.
A multi-step objective $\pi^* = \arg\max_\pi \mathbb{E}[\sum_t \gamma^t
\Delta \volL(G_t)]$ for a discount factor $\gamma \in (0, 1)$ would allow
the actor to trade a small immediate contraction (restricting a dangerous
keystone capability) for a reduction in expected future catastrophic
contraction. The poset construction enables this extension because $\volL$ is
computable at each step (Proposition~\ref{prop:tractability}), making
trajectory simulation tractable on the finite poset.

Risk capabilities also introduce a communication challenge: agents may report
risk assessments about future $\volL$ trajectories, requiring an extension of
the trust model to claims about counterfactual futures rather than observable
present states. Unlike capability claims, which are verifiable through the
observation channel, risk predictions are not directly verifiable until the
catastrophe either happens or does not.

Risk assessment also carries an inherent verification asymmetry visible in
human societies, which we call the \emph{Wamura problem}
\citep{fackler2011fudai}. When mayor Kotoku Wamura built a 15.5-meter
floodgate for the town of Fudai in the 1960s, he
was widely scorned for the excess: the standard seawall height at the time was
10 meters, and the nearby town of Taro built to that standard. Wamura's
insistence on 15.5 meters was seen as wasteful overcaution. He was vindicated
only posthumously: in the 2011 tsunami, Fudai suffered one death (a resident
outside the gate's protection), while Taro's 10-meter wall was overtopped and
multiple residents were killed. The risk that Wamura correctly estimated was
invisible for 44 years; during that time, his assessment was
indistinguishable from genuine waste. A GFM actor faces the same problem: a
risk that is successfully prevented produces no observation confirming the
prevention was necessary, and the convergence timescale for risk-trust can
exceed the timescale on which trust normally accumulates. Addressing the
Wamura problem requires extending the trust model to evaluate risk claims by
causal-structural plausibility rather than observed outcomes; we defer this to
future work.

A final observation: the multi-step objective may also resolve the
structural-avoidance incentive (above). A myopic optimizer avoids structural
discovery because it can contract $\volL$ now; a multi-step optimizer might
accept the short-term contraction because accurate structure improves the
quality of future $\volL$ estimates, leading to better long-run decisions. If
this is the case, the two largest open problems in the poset
construction---structural avoidance and myopic risk blindness---share a common
solution in the discounted objective. This connection is suggestive but
unproven; we defer it to follow-up work on multi-step optimization.

\subsection{Open Problems}

The following problems are introduced or sharpened by the poset construction:

\begin{itemize}
\item \textbf{Structural-avoidance incentive.} The actor is rewarded for
  breadth (capability discovery is $\volL$-expanding) and penalized for depth
  (structural discovery can be $\volL$-contracting). This extends to leverage
  estimation: avoiding capabilities prevents estimating their leverage. A
  regularization term penalizing structural uncertainty, or an
  information-gain bonus for subsumption discovery, would counteract this bias.
\item \textbf{Subsumption correction.} Mistakenly added edges permanently
  redistribute independent weights. A correction mechanism must balance
  robustness (removing bad edges) against inflation (selectively forgetting
  structure to increase $\volL$).
\item \textbf{Automated benchmark refinement.} Can the GFM actor propose
  graded benchmarks for binary-default capabilities? Richer benchmarks
  increase $w(d)$ and may enable deferred subsumption edges to pass the
  non-negativity check.
\item \textbf{Poset scaling.} The $O(n^3)$ computation is polynomial in
  $|\C|$, but $|\C|$ grows with population parameters. When does sparsity
  reduce the effective cost, and how does $n$ scale with population size?
\item \textbf{Intrinsic value.} The self-correcting dynamic narrows the
  B-to-C gap to capabilities that are experientially valuable but have both
  zero leverage and no graded benchmark. Formalizing intrinsic value for this
  residual class, without reintroducing an external theory of experience,
  is the frontier.
\item \textbf{Optimal $\alpha$.} The leverage sensitivity parameter controls
  the tradeoff between measurement fidelity and objective-derived importance.
  Characterizing the optimal $\alpha$ under different population structures
  would ground the leverage ranking signal.
\item \textbf{Multi-step optimization and risk capabilities.} The current
  objective is myopic (one-step lookahead). A discounted multi-step objective
  would enable the actor to identify capability combinations whose immediate
  effect is positive but whose exercise pathways create catastrophic future
  $\volL$-contraction. The poset makes trajectory simulation tractable; the
  remaining work is defining the dynamics model for poset evolution, the
  risk-communication protocol for agents to share risk assessments, and the
  consensus mechanism for disagreement on risk.
\end{itemize}
