\section{Introduction}
\label{sec:introduction}

Goal-frontier maximization (GFM) proposes that an aligned agent should maximize
$\vol(G)$, the volume of the jointly achievable capability space across all
agents in a population \citep{lasser2026gfm}. The foundational paper proves
that this objective has a self-balancing property: destruction, coercion, and
self-imposed rigidity are all anti-maximizing under a single
measure-monotonicity argument, and cooperation is structurally favored. These
results are conditional on a measure satisfying axioms M1--M6, but the paper
provides only one concrete measure, the population empowerment measure
\citep{klyubin2005empowerment, salge2014empowerment}, which is itself
intractable.

This paper provides a tractable construction and discovers an unexpected
structural consequence. We build the capability space $\G$ as a finite weighted
poset ordered by subsumption, discovered through agent communication and
behavioral observation. The poset measure $\volL$, defined via M\"{o}bius inversion, satisfies
axioms M1--M6 and is computable in polynomial time. $\volL$ can be read as a
concrete answer to the problem of ranking opportunity sets by the freedom they
offer \citep{pattanaik1990ranking, sen1999development}: rank them by
M\"{o}bius-inverted weight on a capability poset. This resolves the
foundational paper's central open problem: the structure-of-$\G$ question.

On this poset, a second result emerges. We define \emph{leverage} as the
marginal contribution of a capability to $\volL$ growth rate and show that
cooperation between high-leverage agents is \emph{disproportionately} rewarded
by the objective (Proposition~\ref{prop:leverage_cooperation}). The
foundational paper showed cooperation is favored; the leverage analysis shows
that a specific kind of cooperation (investment in capability-multiplying
infrastructure) is amplified by a factor proportional to the partners'
leverage. Because $\volL$ counts the full population's benefit (not the
coalition's), the cooperative output is outward-facing: it builds
infrastructure that benefits everyone, not just the cooperating parties. We argue that this amplification constitutes a gradient toward
civilization-scale institutions: education, research, communication, and
governance are all instances of outward-facing capability-multiplying
infrastructure, and the formal result says the objective disproportionately
rewards investment in exactly this class of cooperative output. The formal
theorem is local (amplified scoring for high-leverage cooperation); the
civilization-building interpretation is an inference from it.

The poset construction also creates a self-correcting dynamic for the B-to-C
proxy gap identified in the foundational paper. The log weight function
produces diminishing returns on measurement precision ($\Delta w \approx
1/\smax$), making binary-default capabilities---disproportionately the
relational and experiential ones which humanity struggles to
define---high-value targets for the agent's own benchmark development. The
actor is incentivized by its own objective to close the gap between capability
volume and experiential optionality.

\paragraph{Contributions.}
\begin{enumerate}
\item A concrete construction of $\G$ as a finite weighted poset with
  subsumption ordering, discovered through dual-channel sensing
  (Section~\ref{sec:poset}).
\item Proof that the poset measure $\volL$ satisfies axioms M1--M6
  (Section~\ref{sec:axioms}) and is computable in polynomial time
  (Section~\ref{sec:tractability}).
\item An upgraded estimator with real-valued signals and magnitude bounds
  (Section~\ref{sec:estimator}).
\item Leverage as an endogenous importance signal, with proof that
  cooperation between high-leverage agents is amplified, and an argument
  that this amplification constitutes a gradient toward
  civilization-building (Section~\ref{sec:leverage}).
\item A bootstrap procedure with convergence bounds and adversarial
  analysis (Section~\ref{sec:bootstrap}).
\item A worked example with explicit computations demonstrating
  self-balancing and leverage amplification on a concrete poset
  (Section~\ref{sec:example}).
\end{enumerate}
