% =============================================================================
% CONCLUSION (0.5-1 page)
%
% Spec requirements:
%   - Lead with result: self-balancing from measure monotonicity
%   - Local estimator makes GFM tractable and sign-correct
%   - What makes GFM different: one objective for control + scorpion problems
%   - B/C framework predicts its own proxy failures
%   - Forward-looking: aligned to ongoing expansion, not fixed specification
% =============================================================================

\section{Conclusion}
\label{sec:conclusion}

The self-balancing property of goal-frontier maximization follows from a single geometric fact: removing elements from a measurable set cannot increase its measure. Proposition~\ref{prop:self_balancing} shows that this measure monotonicity is sufficient to penalize destruction (Lemma~\ref{lem:destruction}), coercion (Lemma~\ref{lem:coercion}), and self-imposed rigidity (Lemma~\ref{lem:rigidity}) from the same objective. No separate safety mechanism, deontological constraint, or utility-balancing term is required. The three failure modes that divide the existing alignment literature---Goodhart divergence in fixed-utility approaches, brittleness in rule-based approaches, and individual sacrifice in aggregate-welfare approaches---are all instances of $\vol(G)$-contraction, and a GFM actor that faithfully maximizes $\vol(G)$ resists all three for the same reason.

Exact $\vol(G)$ computation is intractable, but the local finite-difference estimator (Definition~\ref{def:local_volume_estimator}) reduces the requirement from volume computation to sign estimation. Proposition~\ref{prop:sign_correctness} establishes that the estimator preserves sign-correctness under stated assumptions, and the assumptions hold most comfortably for the action class that alignment research cares about most: direct harm, coercion, resource destruction, and capability expansion. The estimator is weakest for pure coalition-level changes (Remark~\ref{rem:failure_modes}, case~4), and this boundary is stated rather than hidden.

What distinguishes GFM from competing alignment proposals is not a novel safety constraint but the unification of two problems usually treated separately. The control problem---preventing a powerful agent from pursuing destructive objectives---and the scorpion problem---detecting and responding to agents that persistently contract the joint capability space---are both addressed by the same $\vol(G)$ objective. Tradeoffs between permissiveness and restriction are made explicit through geometric accounting on $G$, not resolved by fiat or hidden behind categorical rules.

The B/C framework adds a second layer of analytical value. By distinguishing the formal proxy (capabilities) from its experiential justification, GFM predicts its own failure modes: the Doll Problem and self-wireheading are cases where the capability proxy diverges from experiential optionality, and the exercised-fraction heuristic (Section~\ref{sec:proxy_failure_detector}) sketches a detection criterion. A framework that can name the conditions under which its own objective function fails is more trustworthy than one that cannot.

The open problems are substantial. The structure of $\G$, the measurement problem for real-world capability signals, the convergence properties of the local estimator, and the formalization of B-to-C divergence all remain unresolved. But the foundation is a single objective aligned not to a fixed human specification---which would be subject to the same Goodhart dynamics the framework diagnoses---but to the ongoing expansion of what all agents can achieve. If the alignment problem is fundamentally a problem of objective selection, then the objective should be one that grows with the population it serves. GFM is a candidate for that objective.
