% =============================================================================
% CONNECTIONS TO EXISTING FRAMEWORKS (1-2 pages)
%
% Spec requirements:
%   - Empowerment maximization (formal bridge)
%   - Sen's capability approach (philosophical bridge)
%   - Free energy principle (structural bridge)
%   - Constitutional AI / RLHF (practical bridge)
%
% Reference material:
%   - Klyubin et al. (2005), Sen (1999), Friston (2013)
%   - Christiano et al. (2017), Bai et al. (2022)
% =============================================================================

\section{Connections to Existing Frameworks}
\label{sec:connections}

Four existing frameworks independently converge on related problems and
collectively supply the tools, intuitions, and empirical track record that GFM
inherits and extends.

\subsection{Empowerment Maximization}
\label{sec:connections_empowerment}

Empowerment, introduced by \citet{klyubin2005empowerment} and surveyed in
\citet{salge2014empowerment}, formalizes the intuition that a capable agent is
one with many distinguishable futures. Formally, an agent's empowerment is the
channel capacity $I(A; S')$ between its available action sequences $A$ and the
resulting states $S'$: the maximum mutual information achievable over all
input distributions on actions. An agent that can reach many distinguishable
states from its current position has high empowerment; an agent trapped in a
narrow corner of state space has low empowerment. The measure is intrinsic (it
depends only on the agent's causal influence over future states, not on any
external reward signal) and information-theoretically tractable: empowerment
can be estimated via variational bounds on mutual information, enabling
practical computation.

GFM is, structurally, \emph{social empowerment}: maximizing $\vol(G)$ is
maximizing the empowerment of the population, measured over the joint
capability space rather than over any individual agent's action-state channel.
Equation~\eqref{eq:joint_goal_space} makes the relationship explicit: $G =
\bigcup_k \Gind{k} \cup \Gcoop$, where $\Gcoop$ captures the cooperative
capabilities that no individual empowerment calculation sees. An agent
reasoning over $I(A_k; S')$ misses the capabilities that exist only in
$\Gcoop$; GFM treats these as first-class objects.

The key difference is scope. Empowerment is agent-centric: it measures what a
single agent can do, and maximizing empowerment for one agent can contract the
capability space of others (seizing territory expands individual empowerment
while destroying cooperative potential). GFM is population-centric: it
penalizes unilateral expansion that shrinks $\vol(G)$ even if individual
empowerment increases. A related approach is attainable utility preservation
\citep{turner2020conservative}, which penalizes actions that reduce the agent's
ability to optimize auxiliary reward functions---a single-agent analogue of
GFM's multi-agent optionality preservation. A natural question is why GFM
introduces the set-theoretic construction $G = \bigcup_k \Gind{k} \cup \Gcoop$
rather than defining social empowerment directly as $\sup_{p(A)} I(A; S')$ over
the joint action space. The answer is structural: the decomposition of $G$ into
individual and cooperative components is what enables
Lemmas~\ref{lem:destruction}--\ref{lem:rigidity}. Social empowerment as a
scalar channel capacity does not decompose the same way; it cannot distinguish
whether a volume change came from destroying an agent's individual
capabilities, restricting cooperation, or imposing rigid rules on the actor.
The set-theoretic construction buys the self-balancing property at the cost of
requiring structure on $\G$ that a channel-capacity formulation avoids. The
practical inheritance from empowerment is tractability: the variational and
sampling techniques developed for single-agent empowerment estimation carry
over to the local finite-difference estimator $\hat{V}$ of
Proposition~\ref{prop:sign_correctness}, with the population-level
generalization being the primary added complexity.

\subsection{Sen's Capability Approach}
\label{sec:connections_sen}

Sen's capability approach \citep{sen1999development} reorients welfare
economics around a deceptively simple claim: what matters for human development
is not income or subjective happiness but the set of \emph{functionings} a
person has genuine freedom to exercise. A person who could be well-nourished
but chooses to fast for religious reasons is differently situated from one who
is undernourished because they lack food. Both have the same nutritional
outcome; only one has the capability. Sen's framework measures the space of
achievable functionings, what people \emph{can} do and be, rather than what
they happen to do or how much they report enjoying it.

GFM adopts the same structural move for alignment: what matters is not the AI's
utility score, not human-reported satisfaction, but the set of capabilities all
agents can actually exercise. Definition~\ref{def:goal} --- goals as
capabilities, $\vol(G)$ as the canonical measure --- is a direct formalization
of Senian well-being applied at population scale and extended to include
non-human agents. The alignment question becomes: does the AI's operation
expand or contract the functioning space of the population it affects?

The key difference is orientation. Sen's capability approach is
\emph{descriptive and evaluative}: it provides a framework for measuring
whether development has occurred and for comparing alternative states of
affairs. It does not prescribe an optimization target or specify how an agent
should act to increase capabilities. GFM is \emph{prescriptive and optimizing}:
it transforms the evaluation criterion into an objective function that an agent
can pursue, with formal self-balancing guarantees
(Proposition~\ref{prop:self_balancing}) and tractable estimation
(Proposition~\ref{prop:sign_correctness}). GFM also inherits a structural advantage from the capability framing: by
measuring the volume of a capability set rather than aggregating individual
preference orderings, it sidesteps the impossibility results that constrain
social choice theory \citep{arrow1951social}. The problem of ranking
opportunity sets by the freedom they offer has been studied formally by
\citet{pattanaik1990ranking}, and $\vol(G)$ can be read as a specific answer to
their question: rank opportunity sets by measure. The decades of formal
development in the capability approach, including Nussbaum's specification of
central human capabilities \citep{nussbaum2000women} and Robeyns's theoretical
survey of capability metrics \citep{robeyns2005capability}, provide a reservoir
of worked examples and partial formalizations that GFM can draw on as the
capability space $\G$ receives more structure in future work.

\subsection{Free Energy Principle}
\label{sec:connections_fep}

The Free Energy Principle (FEP) \citep{friston2013life} proposes that
self-organizing systems minimize \emph{variational free energy} $\mathcal{F}$,
a quantity that upper-bounds the surprise of sensory observations given an
agent's generative model. Under FEP, an agent acts to drive observations toward
model predictions and updates its model when predictions persistently fail.

GFM's trust maintenance dynamics (Appendix~\ref{app:trust_model}, Section~C.3)
exhibit a structural parallel at the world-model level:

\begin{remark}[GFM--FEP Structural Correspondence]
\label{rem:fep_correspondence}
Define the following correspondence between the two frameworks:
\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{FEP} & \textbf{GFM} \\
\midrule
Variational free energy $\mathcal{F}_t$ & Self-prediction error $\sigma_s^2(t)$ \\
Generative model $p(s, o)$ & World model $M_k(G)$ \\
Posterior belief $q(s \mid o)$ & Trust-weighted estimate $\hat{V}$ \\
Belief update (perception) & Model update (Equation~\eqref{eq:model_update}) \\
Active inference (action) & GFM optimization loop (Section~\ref{subsec:optimization_loop}) \\
\bottomrule
\end{tabular}
\end{center}
Under stationarity of the environment, $\sigma_s^2(t) \to \mathbb{E}[r_s^2]$
    (Equation~\eqref{eq:convergence}), so the GFM actor's world-model dynamics
    minimize mean squared prediction error. This is \emph{structurally
    analogous} to minimizing $\mathcal{F}$ under a Gaussian generative model:
    both reduce to minimizing expected squared surprise. The analogy is
    suggestive---both frameworks produce agents whose coherent world models
    emerge from prediction-error minimization---but it falls short of formal
    equivalence: the GFM trust model specifies no variational objective, no KL
    term, no posterior family, and no explicit generative density. Establishing
    a rigorous reduction would require embedding the EWMA tracker of
    Appendix~\ref{app:trust_model} within a variational inference framework,
    which we do not attempt here.
\end{remark}

The differences are in scope and direction. FEP is descriptive. It
characterizes how biological systems behave; it does not prescribe a target or
prove properties of agents that optimize it. GFM is normative. It specifies
$\vol(G)$ and derives the self-balancing property. FEP is individual-centered;
GFM is population-centered. Remark~\ref{rem:fep_correspondence} identifies a
structural parallel between GFM's model-maintenance dynamics and the
prediction-error minimization that FEP describes, but the $\vol(G)$ objective
and its self-balancing consequences are not derived from any biological
imperative; they follow from measure monotonicity applied to a joint
capability space.

\subsection{Constitutional AI and RLHF}
\label{sec:connections_rlhf}

The dominant alignment approach in deployed systems uses human feedback to
shape AI behavior: reinforcement learning from human preferences (RLHF)
\citep{christiano2017deep} collects pairwise human judgments to learn a reward
model, and Constitutional AI \citep{bai2022constitutional} augments this with a
set of principles that guide both the AI's self-critique and the feedback
collection process. Both approaches treat human evaluation as the ground truth
for alignment and use it to train AI systems that humans find helpful and
harmless. A complementary formalization, cooperative inverse reinforcement
learning \citep{hadfield2017cooperative}, models alignment as a cooperative
game in which the AI and the human jointly optimize the human's (initially
unknown) reward function.

GFM provides a theoretical foundation for why this works and a predictive
account of where it will fail. Under GFM, human feedback is a \emph{noisy
channel for communicating the goal frontier}: when a human evaluator prefers
response A over response B, they are, in GFM terms, reporting that A is more
$\vol(G)$-preserving than B from their vantage point. The RLHF reward model is
an estimator of the goal frontier averaged over the evaluator population.
Constitutional principles are domain constraints that encode the evaluators'
understanding of which regions of capability space are inviolable.

GFM makes two predictions about where this estimator will fail. First,
\emph{narrow evaluator pool}: if the evaluators do not represent the full
population of agents whose capabilities are affected, the learned reward model
will systematically underweight capability contractions that affect
non-evaluators (future generations, non-human agents, marginalized
communities). The estimator is biased toward the experiences and capabilities
of whoever is in the feedback pool. Second, \emph{reward hacking}: the trained
AI will find actions that score high on the reward model without actually
expanding $\vol(G)$, responses that evaluators find persuasive or satisfying
but that contract the goal frontier in ways evaluators cannot detect. Both
failure modes are predictions of the GFM framework they follow from treating
RLHF as a noisy estimator of a population-level quantity rather than as a
reliable proxy for alignment itself.
