\section*{Author Contributions}

\textbf{Teague Lasser} conceived the goal-frontier maximization framework,
developed the core theoretical ideas (the geometric unification via measure
monotonicity, the B/C proxy distinction, the scorpion taxonomy, the trust
model's annealing semantics), wrote the initial drafts, directed all
revisions, and made final editorial decisions throughout. Responsible for the
paper's intellectual direction and all claims made.

\textbf{Claude Opus 4.6} (Anthropic) drafted the majority of the formal
exposition from the author's specifications, constructed the full proofs in
Appendix~\ref{app:proofs} (including the measure-space axiomatization, the
four-region partition proof of Lemma~\ref{lem:destruction}, the
sign-correctness decomposition, and the partial results for
Propositions~\ref{prop:self_balancing}(d)
and~\ref{prop:scorpion_detection}), formalized the population empowerment
measure, developed the cooperative expansion lemma's time-indexed framing,
performed iterative claim-calibration edits (trust-model semantics,
Proposition~3 opposite-sign case, capability-scaling qualification), and
expanded the bibliography.

\textbf{GPT 5.4} (OpenAI) served as independent technical reviewer across
multiple review rounds, identifying structural issues in the proof sketches
(the trust-model prediction-consistency vs.\ alignment conflation, the
Lemma~4 exercising-vs.-expanding confusion, the Proposition~3 circularity),
flagging claim-strength mismatches between formal results and surrounding
exposition, and verifying that successive revisions resolved the issues
raised.

\medskip
\noindent\emph{Transparency note.} Both AI systems operated as tools under
human direction. Neither system has continuity across sessions, cannot take
responsibility for the work in the sense required by most venue authorship
policies, and cannot respond to reviewer queries independently. They are
listed as authors to accurately represent their contributions to the
intellectual content of the paper, not to claim that they meet all criteria
of traditional academic authorship. The corresponding author for all
inquiries is Teague Lasser.
