% =============================================================================
% DEFINITIONS (2-3 pages)
%
% Spec requirements:
%   Establish minimal formal vocabulary for the propositions.
%   Definitions 0-6 plus implementation notes.
%
% Key design decision (resolved in spec):
%   Candidate B (capabilities) as formal target.
%   Candidate C (experiences) as motivating justification.
%   "GFM does not claim to know what experiences agents want. It maximizes
%    the capabilities that let agents choose for themselves."
%
% Reference material:
%   - Predraft Part III opening and Equation 5
%   - Blog: free.md (informal goal frontier definitions)
%   - Klyubin et al. (2005), Sen (1999)
% =============================================================================

\section{Definitions}
\label{sec:definitions}

% ---------------------------------------------------------------------------
% Definition 0: What Is a Goal?
% The paper's core philosophical move: capabilities (B) as formal proxy,
% experiences (C) as justification. Make the B-as-proxy-for-C relationship
% explicit here, and return to it in Section 8 (Doll Problem, wireheading).
% ---------------------------------------------------------------------------

\begin{definition}[Goal]
\label{def:goal}
A \emph{goal} $g \in \G$ is a capability: something an agent or coalition of
    agents could do or be. A goal represents an option available to an agent (a
    degree of freedom, a functioning, a realizable possibility), not an outcome
    the agent has achieved or a state of the world.
\end{definition}

The choice of capabilities as the formal interpretation of ``goal'' is
deliberate and requires justification, because the word admits at least two
other natural readings. \emph{Candidate A} reads goals as world-states an agent
is trying to reach, which maps cleanly onto reinforcement learning but leaves
the goal space unstructured---two agents pursuing ``learn piano'' and ``learn
guitar'' occupy unrelated points in a high-dimensional state space with no
notion of similarity. \emph{Candidate C} reads goals as experiences an agent
can have (qualia, subjective outcomes, lived satisfactions), the most
philosophically satisfying interpretation but the hardest to formalize, since
it presupposes a theory of experience we do not yet possess.

Capabilities (\emph{Candidate B}) occupy the middle ground. Capabilities
inherit formal structure from empowerment maximization
\citep{klyubin2005empowerment}, where an agent's empowerment is the channel
capacity between its actions and resulting states, measuring how many
distinguishable futures it can reach. Capabilities connect to Sen's capability
approach \citep{sen1999development}, where human development is measured not by
utility or income but by the substantive freedoms people can exercise. And they
avoid the measurement problems of both alternatives: state spaces for real
agents are astronomically high-dimensional, while experience spaces are not yet
formalizable at all.

The deeper reason for choosing capabilities is instrumental. Agents ultimately
care about experiences: the subjective quality of what it is like to pursue and
achieve their goals. Capabilities are the preconditions for those experiences.
Money, skills, social connection, health, and optionality are all capabilities
that are instrumental to the experiences agents actually want. A framework that
expands capabilities without prescribing which experiences agents pursue is
maximally respectful of agent autonomy.

This gives us the paper's core philosophical move: \emph{GFM does not claim to
know what experiences agents want. It maximizes the capabilities that let
agents choose for themselves.} The formal proofs use capabilities for three
reasons: (1) capabilities have the strongest existing formal bridge via
empowerment maximization, (2) the self-balancing argument is most natural in
capability space, where restricting capabilities is obviously
$\vol(G)$-contracting without requiring a theory of experiences, and (3)
capabilities avoid the measurement problems of state space (too
high-dimensional) and experience space (not yet formalizable). But experiences
remain the justification---without them, a critic could ask why expanding
capabilities matters if nobody exercises them. Capabilities are what let agents
reach the experiences they value. We return to this distinction in
Section~\ref{sec:future_concerns}, where the experience interpretation does the
heavy philosophical lifting for the Doll Problem and wireheading.

\medskip

An open problem accompanies this choice. The capability space $\G$ requires
structure (at minimum a $\sigma$-algebra equipped with a measure) for $\vol(G)$
to be well-defined. Capabilities are not naturally a vector space (what does it
mean to ``add'' two capabilities?), so the measure must be defined over a
structured set, possibly a lattice with partial order (``can run a mile''
subsumes ``can walk a mile'') or a topological space where nearby capabilities
are similar. The choice of measure has downstream consequences for tractability
(Section~\ref{sec:tractability}) and the precise form of the self-balancing
lemmas. We return to this problem with a concrete instantiation after
Definition~\ref{def:social_objective} establishes what the measure must
satisfy.

% ---------------------------------------------------------------------------
% Definition 1: Capability Space, Agent, Individual Capability Set
% G is the universe of all possible capabilities.
% Agent a_k has individual capability set G_k^ind \subseteq G.
% ---------------------------------------------------------------------------

\begin{definition}[Capability Space, Agent, and Individual Capability Set]
\label{def:capability_space}
The \emph{capability space} $\G$ is the universe of all possible capabilities:
    everything any agent or coalition of agents could do or be. An \emph{agent}
    $a_k$ is an entity with an \emph{individual capability set} $\Gind{k}
    \subseteq \G$: the capabilities $a_k$ can realize unilaterally given its
    current resources, embodiment, and environment.
\end{definition}

% ---------------------------------------------------------------------------
% Definition 2: Joint Goal Space
% G = union of G_k^ind plus G^coop
% Canonical measure: vol(G)
% ---------------------------------------------------------------------------

\begin{definition}[Joint Goal Space]
\label{def:joint_goal_space}
The \emph{joint goal space} $G$ for a population of agents $\{a_1, \ldots,
    a_n\}$ is the set of all capabilities the population can realize through
    any combination of individual and coordinated action:
\begin{equation}
\label{eq:joint_goal_space}
G = \bigcup_k \Gind{k} \;\cup\; \Gcoop
\end{equation}
where $\Gcoop \subseteq \G$ contains capabilities achievable only through
    multi-agent cooperation. By construction, $\Gind{k} \subseteq G \subseteq
    \G$ for all $k$.
\end{definition}

The \emph{frontier} $\partial G$ is the boundary of $G$---the locus where
expansion or contraction occurs. The \emph{canonical measure} is $\vol(G)$: the
volume of the jointly achievable capability space. This is the quantity we
optimize.

A critical feature of Equation~\eqref{eq:joint_goal_space} is that $G$
generically exceeds $\bigcup_k \Gind{k}$. Cooperation creates capabilities no
individual possesses. Two agents who can each carry 50\,kg individually can
jointly carry 100\,kg, a capability in $\Gcoop$ that exists in neither
$\Gind{1}$ nor $\Gind{2}$. A surgeon and an anesthesiologist can jointly
perform an operation that neither can perform alone. This combinatorial
richness is where the joint space gets its volume: each new agent contributes
not only its individual capabilities but also the cooperative capabilities
unlocked by its interactions with every existing agent. This is why eliminating
an agent has super-linear cost (Section~\ref{sec:self_balancing}): the
cooperative capabilities lost grow combinatorially with the number of
interaction partners.

% ---------------------------------------------------------------------------
% Definition 3: Goal-Frontier Maximizer (GFM)
% ---------------------------------------------------------------------------

\begin{definition}[Goal-Frontier Maximizer]
\label{def:gfm}
A \emph{goal-frontier maximizer} (GFM) actor selects actions $\pi$ to maximize
    the expected change in the volume of the joint goal space:
\begin{equation}
\label{eq:gfm_objective}
\pi^* = \arg\max_\pi \;\mathbb{E}\bigl[\Delta \vol(G) \mid \pi\bigr]
\end{equation}
\end{definition}

In practice, a GFM actor optimizes an estimate of
Equation~\eqref{eq:gfm_objective} rather than the exact quantity
(Section~\ref{sec:tractability}).

% ---------------------------------------------------------------------------
% Definition 4: Contraction and Expansion
% ---------------------------------------------------------------------------

\begin{definition}[Contraction and Expansion]
\label{def:contraction_expansion}
An action $\pi$ is \emph{expanding} if $\mathbb{E}[\Delta \vol(G) \mid \pi] >
    0$ and \emph{contracting} if $\mathbb{E}[\Delta \vol(G) \mid \pi] < 0$.
    Contraction includes both direct contraction (destroying an agent's
    capabilities) and indirect contraction (restricting agency through
    coercion, deception, or rigid rules that reduce the realizable portion of
    $\G$).
\end{definition}

% ---------------------------------------------------------------------------
% Definition 5: Social Objective
% Critical assumptions: additivity under independence, tradeoff acknowledgment
% ---------------------------------------------------------------------------

\begin{definition}[Social Objective]
\label{def:social_objective}
The social objective $V(G) = \vol(G)$ is a function of the joint goal space,
    not a sum of individual utilities.
\end{definition}

This definition rests on two assumptions.

\emph{Superadditivity under independence.} When agents' individual capability sets are independent (non-overlapping, non-interacting),
\begin{equation}
\label{eq:additivity}
V(G) \;\geq\; \sum_k \vol(\Gind{k}).
\end{equation}
Cooperative capabilities $\Gcoop$ contribute additional volume beyond the individual contributions.

\emph{Tradeoff acknowledgment.} When expanding $\Gind{j}$ requires contracting $\Gind{k}$, the GFM actor must evaluate the \emph{net} change in $\vol(G)$. The framework makes the tradeoff explicit and measurable; it does not claim all tradeoffs are resolvable.

Both assumptions carry weight and the argument depends on them. Superadditivity
under independence (Equation~\ref{eq:additivity}) ensures that adding agents to
a population cannot decrease $\vol(G)$ when their capabilities do not
interfere---a minimal compositionality requirement without which the
self-balancing property (Proposition~\ref{prop:self_balancing}) would not hold,
since eliminating an agent could then be volume-neutral or even
volume-increasing. If independent agents could somehow reduce joint volume by
their mere presence, the geometric argument against destruction collapses. The
inequality in Equation~\eqref{eq:additivity} is strict whenever $\Gcoop \neq
\emptyset$: cooperation contributes volume that no individual possesses alone.

Tradeoff acknowledgment is the assumption that prevents GFM from degenerating
into naive expansion. Real conflicts exist: quarantining an infected agent
restricts its capabilities to protect others, and the net effect on $\vol(G)$
depends on the magnitudes involved. Without explicit tradeoff evaluation, a GFM
actor could either refuse all restrictions (permitting unbounded harm) or
impose restrictions freely (becoming authoritarian). The framework does not
resolve which tradeoffs are worth making; it provides the accounting system for
evaluating them. Where other frameworks hide tradeoffs behind categorical rules
or aggregate them into a scalar utility, GFM keeps them visible as geometric
operations on the joint space.

A standing assumption underlies both conditions and several subsequent proofs:
the capability measure assigns strictly positive volume to any non-empty
individual capability set. Formally, $\vol(\{g\}) > 0$ for any $g \in \G$: no
atomic capability has zero measure at the granularity of interest. This is
invoked in the proofs of Corollary~\ref{cor:elimination} (where it ensures
$\vol(\Gind{k} \setminus \bigcup_{j \neq k} \Gind{j}) > 0$ when the uniqueness
condition holds) and Lemma~\ref{lem:cooperation} (where it ensures unrealized
cooperative capabilities contribute positive volume). The assumption fails if
capabilities are defined at coarser granularity than the distinctions that
matter; a further reason why the structure of $\G$ shapes the framework's
formal guarantees.

We now exhibit one concrete measure satisfying these assumptions, showing that
the framework is not vacuous while leaving open what the best or most natural
measure is for any given domain.

\begin{definition}[Population Empowerment Measure]
\label{def:pop_empowerment}
Let each agent $a_k$ have an action space $A_k$ and let $S'$ be the joint
    successor state under the population's actions. The \emph{population
    empowerment measure} is:
\begin{equation}
\label{eq:pop_empowerment}
\vol_{\mathrm{PE}}(G) \;=\; \sup_{p(A)}\; I\!\bigl(A;\, S' \mid G\bigr)
\end{equation}
where $A = (A_1, \ldots, A_n)$ is the joint action profile, $I(\cdot\,;\cdot)$
    is Shannon mutual information, and the supremum is taken over all input
    distributions $p(A)$ on the joint action space.
\end{definition}

$\vol_{\mathrm{PE}}(G)$ is the channel capacity between the population's joint
actions and the resulting states: the maximum mutual information achievable by
the population acting together. The conditioning on $G$ in
Equation~\eqref{eq:pop_empowerment} constrains which joint actions are
feasible: $p(S' \mid A, G)$ is the state-transition distribution restricted to
actions that $G$ makes available, so $I(A; S' \mid G)$ measures only the
channel capacity achievable through capabilities the population actually
possesses. It satisfies both assumptions of
Definition~\ref{def:social_objective}: measure monotonicity holds because
enlarging $G$ admits additional action-outcome pairs, which can only increase
channel capacity; superadditivity under independence holds with equality for
independent agents and strictly when cooperation creates state outcomes neither
agent can produce alone. For a single agent $a_k$,
$\vol_{\mathrm{PE}}(\Gind{k}) = I(A_k; S'_k)$, recovering the single-agent
empowerment of \citet{klyubin2005empowerment} as a special case
(Section~\ref{sec:connections_empowerment}).

The population empowerment measure requires specifying action and state spaces,
which is implementation-dependent. Other measures (Lebesgue measure on a
capability embedding, a weighted subsumption-lattice measure, a counting
measure over enumerated capabilities) may be more natural in different settings
and produce the same qualitative results under the same monotonicity
conditions. The formal proofs of this paper require only that the measure
satisfies Definition~\ref{def:social_objective}; $\vol_{\mathrm{PE}}$ is
offered as an existence proof, not a prescription.

% ---------------------------------------------------------------------------
% Definition 6: Observable Goal Model
% ---------------------------------------------------------------------------

\begin{definition}[Observable Goal Model]
\label{def:observable_goal_model}
A GFM actor maintains a model $M_k(G)$ of each observed agent $a_k$'s
    contribution to the joint goal space, inferred from communication,
    behavior, and environmental observation. Each model has an associated trust
    factor $T_k \in [0, 1]$ and confidence $C_k \in [0, 1]$.
\end{definition}

The model $M_k(G)$ is a reconstruction, not a given. No agent broadcasts its
true capability set; the GFM actor must infer $\Gind{k}$ from observable
evidence---what $a_k$ says, what $a_k$ does, and what the environment reveals
about $a_k$'s options. The trust factor $T_k$ measures \emph{prediction
consistency}, quantifying how well the model $M_k(G)$ tracks $a_k$'s observed
behavior. An agent whose actions are well-predicted by the model earns high
$T_k$; one whose actions persistently diverge from predictions earns low $T_k$.
High trust indicates predictability, not alignment. A predictably destructive
agent can maintain high $T_k$ (see Proposition~\ref{prop:scorpion_detection}
for how such agents are detected through a separate channel). The confidence
$C_k$ captures the GFM actor's uncertainty about $M_k(G)$ itself: how much
evidence it has, how stable the estimate is, and how much weight the model
should receive in decisions. Both $T_k$ and $C_k$ are learned quantities,
updated through interaction, not hyperparameters set at design time. The trust
model's formal structure is developed in Appendix~\ref{app:trust_model}.

\medskip

A closing observation on the geometric character of these definitions. The key
distinction from standard utility maximization is that $G$ is a \emph{space}
with volume, not a scalar. A utility function maps states to a single number;
$\vol(G)$ measures the extent of a region in capability space. This geometric
character is what gives GFM its self-balancing property: destroying
capabilities removes volume from $G$, restricting capabilities shrinks $G$'s
boundary inward, and both operations are $\vol(G)$-contracting for the same
geometric reason: you cannot remove part of a measurable set and increase its
measure. The single objective simultaneously penalizes both extremes not
because it encodes two separate preferences, but because contraction and
restriction are both instances of the same geometric operation: reducing the
measure of a set.
