% =============================================================================
% FUTURE CONCERNS AND OPEN PROBLEMS (1-2 pages)
%
% Demonstrates B/C framework's predictive power, identifies proxy failures,
% sketches correction criterion, states open problems.
% =============================================================================

\section{Future Concerns and Open Problems}
\label{sec:future_concerns}

The preceding sections formalized GFM over capabilities (Candidate B from
Definition~\ref{def:goal}) and identified the conditions under which the
self-balancing property survives local approximation. The B/C framework's
analytical contribution is that it can precisely identify \emph{when and why}
the formal proxy fails, where capability volume diverges from the experiential
optionality that justifies it. A framework that names its own failure modes is
more trustworthy than one that hides them. This section demonstrates three
predicted failure patterns, sketches a correction heuristic, identifies a class
of human-level proxy traps that GFM would flag, and states the open problems
that remain.

\subsection{The Doll Problem}
\label{sec:doll_problem}

Consider a population in which AI companions increasingly replace human
relationships. Individuals who find reciprocal human connection difficult may
voluntarily substitute simulated companions: programmable, frictionless, always
available. Under the capability formalization alone, this substitution may not
register as contraction: the agent's capability set $\Gind{k}$ appears neutral
or even expanded, since the agent retains all prior capabilities and gains a
new one (access to a simulated companion). A GFM actor monitoring only
$\vol(G)$ could observe stable or growing capability volume throughout this
transition.

The B/C framework predicts this divergence. Capabilities are the formal proxy;
experiences are the justification. A simulated companion preserves the
capability of ``having a conversational partner'' but eliminates the experience
of genuine reciprocal connection: the vulnerability, unpredictability, and
mutual growth that characterize human relationships. Experiential optionality
contracts even as the formal capability set holds steady. This is precisely the
proxy failure pattern that the B/C distinction was designed to detect: stable
$\vol(G)$ with declining experiential richness.

A GFM actor operating purely on the capability formalization would not catch
this. The formal machinery alone is insufficient; the experience layer does the
philosophical work. GFM does not solve the Doll Problem; it \emph{predicts the
failure category}. A framework lacking the B/C distinction cannot name what
went wrong: it would observe unchanged capability volume and conclude nothing
had happened. The two-layer structure makes the failure diagnosable in advance,
even when the formal layer cannot detect it autonomously.

\subsection{Self-Wireheading}
\label{sec:self_wireheading}

Algorithmically optimized content feeds present a subtler instance of the same
pattern. Recommendation systems steer users incrementally toward lower-effort,
higher-dopamine activities: short-form video over long-form reading, parasocial
engagement over reciprocal conversation, consumption over creation. Each
incremental step is freely chosen. The agent's capabilities are formally
unchanged: it \emph{could} still read a book, learn a skill, or initiate a
difficult conversation. Nothing in $\Gind{k}$ has been removed.

The B/C framework predicts what happens: $\vol(G)$ remains stable while
exercisability declines. The agent's \emph{revealed} capability set, the
capabilities it actually exercises, narrows progressively even as the formal
set stays constant. This is wireheading in the Senian sense: functionings are
technically available, but the effective freedom to exercise them is eroding.
The proxy is being Goodharted. A GFM actor monitoring capability volume alone
would see a flat trajectory and take no action, while the experiential
optionality it is supposed to protect quietly collapses.

The formal capability measure would miss this as well. The framework's value
lies in predicting the failure pattern---stable capability volume coupled with
declining exercisability, not in catching it through $\vol(G)$ alone. Both the
Doll Problem and self-wireheading exhibit the same signature: the B-to-C gap
opening while the formal proxy reports no change.

\subsection{The Substitution Problem}
\label{sec:substitution_problem}

Corollary~\ref{cor:elimination} established that elimination is \emph{almost
always} anti-maximizing, with the qualifier tied to a uniqueness condition:
agent $a_k$ has at least one capability not possessed by any other agent. The
substitution problem asks what happens when this condition fails, not through
any destructive or coercive act, but through a coalition becoming so capable
that it subsumes the capability sets of everyone outside it.

Formally: suppose a coalition $C$ satisfies $\Gind{k} \subseteq
G_C^{\mathrm{ind}} \equiv \bigcup_{j \in C} \Gind{j}$ for every $a_k \notin C$.
Then $\vol(\Gind{k} \setminus G_C^{\mathrm{ind}}) = 0$ for all
non-members---their individual contributions to $G$ are zero. The coalition
need not eliminate anyone. It suffices to make non-members functionally
irrelevant: their individually achievable capabilities are fully replicated by
the coalition, and their participation in $\Gcoop$ is contingent on the
coalition's permission. The strict inequality of Lemma~\ref{lem:destruction}
fails; the self-balancing argument no longer guarantees that marginalizing
non-members is anti-maximizing. This is the monopolar outcome the framework is
intended to prevent, arising from capability subsumption rather than from any
action the lemmas directly penalize.

The B/C framework predicts precisely when this failure occurs. Under the narrow
capability definition (capabilities as tasks that can be performed, economic
functionings), substitution is possible. A sufficiently capable coalition can
replicate the productive output of any less-capable agent, driving that agent's
unique contribution to zero. The self-balancing property's protection against
elimination dissolves.

Under the broad capability definition (capabilities as modes of being,
including experiential and relational capabilities), the uniqueness condition
holds for \emph{every} agent regardless of the coalition's productive power.
The capability ``to experience the world as this particular agent, from this
embodiment and history'' is individuated by the identity of the experiencing
subject. No coalition can possess it by proxy, because it is not a function
that can be delegated---it is a fact about who is doing the experiencing.
Relational capabilities are the same: being this person's parent, this
community's elder, this lineage's bearer are not roles a more capable agent can
fill by substitution, because the relationships are constituted by the specific
agents in them. Under this reading, $\vol(\Gind{k} \setminus \bigcup_{j \neq k}
\Gind{j}) > 0$ for every agent in any population, and the ``almost always''
qualifier in Corollary~\ref{cor:elimination} becomes universal.

The population empowerment measure (Definition~\ref{def:pop_empowerment})
formalizes this. $\vol_{\mathrm{PE}}(G) = \sup_{p(A)} I(A; S' \mid G)$ counts
the distinguishable outcomes the population can produce. A human's experienced
existence---the particular sequence of states constituting their life from
their particular perspective---is a distinguishable element of the joint state
space $S'$ that no other agent can produce, because the identity of the
experiencing subject is part of the outcome's description. Under the
information-theoretic measure, every agent contributes strictly positive
channel capacity regardless of whether their productive outputs are replicable
by the coalition.

As with the Doll Problem and self-wireheading, a GFM actor operating on the
narrow capability definition would not detect the substitution problem. Stable
or growing formal $\vol(G)$ would be reported as the coalition subsumed
non-member capabilities one by one. Only the broad definition---anchored in
experiential optionality---closes the gap. The substitution problem is the
collective version of the earlier individual failures: stable proxy volume,
declining experiential substrate, B-to-C divergence at population scale.

\subsection{Toward a Proxy-Failure Detector}
\label{sec:proxy_failure_detector}

The Doll Problem and self-wireheading share a structural pattern: capabilities
are formally available but exercisability declines. This suggests a correction
heuristic. A GFM actor should monitor not only $\vol(G)$ but the
\emph{exercised fraction}, the ratio of capabilities agents actually use to
capabilities formally available. Define $\rho_k(t)$ as the fraction of
$\Gind{k}$ that agent $a_k$ exercises over a window $[t - \tau, t]$. A
persistent decline in $\rho_k$ across many agents, even with stable or growing
$\vol(G)$, constitutes a B-to-C divergence signal: the formal proxy is
reporting health while the experiential substrate is contracting.

This is a heuristic, not a solution, and it introduces its own proxy problem.
Not all unused capabilities represent experiential contraction. An agent who
specializes---exercising a narrow set of capabilities with increasing
depth---is not wireheading. A musician who stops playing sports to practice
more is exercising fewer capabilities but may be expanding experiential
richness within a chosen domain. Specialization is not contraction, and a naive
exercised-fraction monitor would conflate the two. The heuristic would need to
distinguish between voluntary concentration (stable or increasing depth within
a narrowing set) and involuntary atrophy (declining engagement across the
board), a distinction that reintroduces the measurement problems the capability
formalization was designed to avoid.

The heuristic does not resolve the proxy failure. The B/C framework's
contribution is narrower: it makes the failure \emph{detectable in principle}.
When $\vol(G)$ and $\rho$ diverge---stable volume, declining exercise---the
framework flags that the capability proxy is failing and that the GFM actor
should increase its uncertainty about whether its objective is tracking the
underlying justification. Frameworks without the B/C distinction cannot offer
even this: they have no internal signal that anything is wrong.

\subsection{Human Proxy-Trap Susceptibility}
\label{sec:human_proxy_traps}

The B-to-C divergence pattern is not specific to AI-mediated scenarios. Humans
are chronically susceptible to Goodhart's Law applied to their own goals:
optimizing proxy rewards (money, status, social-media engagement metrics)
rather than the experiences those proxies were instrumental toward. An
individual who accumulates wealth far beyond the point where it expands
experiential optionality is optimizing a proxy that has decoupled from its
justification. An institution that maximizes measurable outputs (publications,
quarterly earnings, engagement metrics) while the outcomes those metrics were
supposed to track deteriorate is exhibiting the same pattern at an
organizational scale.

A GFM framework applied to human institutions would flag these proxy traps
through the same mechanism. When an agent's or institution's $\vol(G)$ grows
(more measurable options) but $\rho$ declines (fewer of those options are
actually exercised or lead to experiential expansion), the divergence signal
fires. This does not require AI mediation, it is a diagnostic that applies
wherever proxy objectives have displaced the goals they were designed to serve.
The framework's contribution is providing a vocabulary for a failure mode that
Goodhart's Law names but does not formalize: the B/C structure specifies
\emph{what} diverges (capabilities from experiences) and \emph{how} to detect
it (volume-exercise divergence).

\subsection{Open Problems}
\label{sec:open_problems}

\begin{itemize}
    \item \textbf{Closing the B-to-C gap.} The paper adopts capabilities as the formal proxy for experiences, but this section demonstrates three cases where the proxy fails. The exercised-fraction heuristic is a first step toward detection, but formalizing B-to-C divergence, and proving that any correction criterion does not introduce worse proxy problems, remains the central open challenge.

    \item \textbf{Structure of $\G$.} The capability space requires structure beyond a bare set: at minimum a $\sigma$-algebra with a measure so that $\vol(G)$ is well-defined. Richer structure (a lattice with a natural subsumption order, or a topological space where nearby capabilities are similar) would enable stronger results. This choice determines what ``volume'' means and remains the most urgent formal question.

    \item \textbf{Measurement.} What sensors and signals does a GFM actor need to construct the local volume estimator $\Vhat$ in practice? The estimator requires observable capability changes across agents, but the mapping from real-world observations to capability-space signals is unspecified.

    \item \textbf{Convergence.} Does the local finite-difference estimator converge to the true $\vol(G)$ trajectory over time? Under what conditions on observation coverage, trust-model accuracy, and population dynamics does the estimator track the ground truth rather than drifting?

    \item \textbf{Plurality.} Can multiple GFM actors with different models of $\G$ coordinate? When actors disagree about the structure of the capability space itself, not just the shape of $G$ within it, the shared-objective argument breaks down and requires extension.

    \item \textbf{Bootstrap.} Initializing a GFM actor requires estimating other agents' capability sets, which is itself a hard inference problem made harder by the need to discover the structure of $\G$ simultaneously. The goal estimation problem and the capability-space structure problem are coupled, and solving one presupposes progress on the other.
\end{itemize}
