% =============================================================================
% INTRODUCTION (1-2 pages)
%
% Spec requirements:
%   - Frame alignment as choosing the right objective, not the right rules
%   - Failure modes: fixed utility (Goodhart), deontological (brittle),
%     unconstrained consequentialism (sacrifices individuals)
%   - Introduce GFM as third option: maximize achievable goals for all agents
%   - Central claim: GFM resists both destructive action and overreactive
%     constraint from the same objective
%   - Roadmap the paper
%
% Reference material:
%   - Predraft Section VI (Evil Twin Time Bomb)
%   - Blog: frameskipping.md ("Misleading Definitions" section)
%   - Bostrom (2014), Hubinger et al. (2019), Carlsmith (2021)
%
% What NOT to include:
%   - ML basics (predraft Section II)
%   - "ML models cannot adapt" (predraft Section III)
%   - Sigmoid vs exponential discussion
%   - Extended Clippy argument (cite blog instead)
% =============================================================================

\section{Introduction}
\label{sec:introduction}

The alignment problem is, at root, a problem of objective selection. An
artificial agent powerful enough to reshape its environment will reshape it in
the direction its objective points. The question is not which rules to impose
on such an agent---rules can be gamed, circumvented, or rendered obsolete by
distributional shift---but which objective function, once faithfully optimized,
produces behavior that a civilization of diverse agents would endorse. The
difficulty is that most candidate objectives fail in characteristic and
well-documented ways \citep{amodei2016concrete, ngo2022alignment}, and the
failures are not incidental but structural: they follow from the mathematical
properties of the objective itself.

Three families of alignment proposals dominate the current literature, and each
fails for a distinct geometric reason.

\emph{Fixed utility functions} reduce alignment to scalar optimization: specify
what the agent should value, assign numerical weights, and maximize the sum.
The failure mode is Goodhart's Law: any fixed proxy, once optimized
sufficiently hard, diverges from the quantity it was intended to track
\citep{bostrom2014superintelligence, manheim2018categorizing}. Specification
gaming, reward hacking, and perverse instantiation are all instances of this
divergence. Mesa-optimization compounds the problem: a learned optimizer may
construct an internal objective that correlates with the training signal
in-distribution but pursues different goals out-of-distribution
\citep{hubinger2019risks}. The fixed-utility approach assumes that the right
scalar exists and merely needs to be found; the structural critique is that no
scalar faithfully captures a multi-agent world's interests, because compressing
a high-dimensional capability space into a single number necessarily discards
the geometric structure that makes tradeoffs visible.

\emph{Deontological rules} are hard constraints on permissible actions. They
address the failure of fixed utilities by replacing optimization with
prohibition. The failure mode is brittleness under adversarial conditions. Any
finite rule set partitions the action space into permitted and forbidden
regions, and an adversary with knowledge of the partition can construct
scenarios that force the rule-bound agent into harmful outcomes: situations
where every permitted action is dominated by a forbidden one. We develop a
concrete construction of this attack in Appendix~\ref{app:evil_twin}, where a
rule-bound agent facing an adversarial twin is forced to accept a destructive
outcome that a flexible agent could have avoided. The deeper problem is that
rigidity is itself a cost: an agent constrained to a strict subset of its
action space cannot, by definition, outperform an unconstrained agent
optimizing the same objective (Lemma~\ref{lem:rigidity}). Rules are useful as
heuristics but elevating them to inviolable constraints guarantees
suboptimality in adversarial environments, and since they cannot evolve with
changes an agent might experience they lock the agent out of adaptation to new
environmental constraints.

\emph{Unconstrained consequentialism} maximizes aggregate welfare without side
constraints. It avoids both Goodhart fragility and deontological rigidity, but
introduces a third failure mode: it can sacrifice individuals for aggregate
gain. An agent maximizing total utility across a population has no structural
reason to preserve any particular agent's capabilities if eliminating that
agent produces a net increase in the aggregate. Power-seeking behavior emerges
naturally under such objectives \citep{carlsmith2022power, turner2021optimal},
because concentrating capabilities in a single agent can increase that agent's
contribution to the aggregate even as it destroys the cooperative capabilities
that depend on distributed agency. The failure is not moral but geometric:
aggregate measures are insensitive to the distribution of capabilities across
agents, and redistribution that concentrates volume in fewer agents destroys
the combinatorial cooperative structure that generates most of the joint
space's measure.

This paper proposes a fourth option: \emph{goal-frontier maximization} (GFM). A
GFM actor selects actions to maximize $\vol(G)$, the volume of the jointly
achievable capability space across all agents in the population. The central
claim is that this single objective naturally resists both destructive action
and overreactive constraint, from the same geometric principle. Destroying an
agent's capabilities removes volume from $G$ (measure monotonicity);
restricting an agent's capabilities shrinks $G$ inward (subset monotonicity);
imposing rigid rules on the actor itself reduces its ability to find
$\vol(G)$-expanding actions (optimization over subsets). The self-balancing
property (Proposition~\ref{prop:self_balancing}) comprises three consequences
of maximizing a single measure over a joint space, not three separate safety
mechanisms. The objective penalizes both extremes of the
permissiveness--control spectrum because both extremes are instances of the
same geometric operation: contraction of a measurable set.

The paper proceeds as follows. Section~\ref{sec:definitions} establishes the
formal vocabulary: goals as capabilities, the joint goal space $G$, the GFM
objective, and the capability-experience framework that distinguishes the
formal proxy (Candidate B: capabilities) from its experiential justification
(Candidate C: experiences). Section~\ref{sec:self_balancing} proves the
self-balancing property through three lemmas on destruction, coercion, and
rigidity, synthesized in Proposition~\ref{prop:self_balancing}, with
corollaries on elimination cost and rule suboptimality.
Section~\ref{sec:tractability} confronts the intractability of exact $\vol(G)$
computation, introduces a local finite-difference estimator
(Definition~\ref{def:local_volume_estimator}), and identifies the margin
condition under which the estimator preserves sign-correctness
(Proposition~\ref{prop:sign_correctness}). Section~\ref{sec:multi_agent}
extends the analysis to multi-agent populations, characterizing cooperation,
defection, and the trust dynamics that govern the transition between them.
Section~\ref{sec:connections} grounds GFM in four existing
frameworks---empowerment maximization, Sen's capability approach, the free
energy principle, and RLHF---identifying what GFM inherits and where it
diverges. Section~\ref{sec:future_concerns} demonstrates the B/C framework's
predictive power by identifying three failure patterns (the Doll Problem,
self-wireheading, and the Substitution Problem) where the capability proxy
diverges from experiential optionality, sketches a correction criterion, and
states the open problems that remain. Appendices provide worked examples: the
Evil Twin adversarial construction (Appendix~\ref{app:evil_twin}), the trust
model's formal structure (Appendix~\ref{app:trust_model}), the scorpion
taxonomy (Appendix~\ref{app:scorpion_taxonomy}), and agent similarity metrics
(Appendix~\ref{app:agent_similarity}).
