% =============================================================================
% MULTI-AGENT DYNAMICS (2-3 pages)
%
% Spec requirements:
%   - GFM-GFM cooperation (shared objective, coalition formation)
%   - GFM-Scorpion interaction (detection, containment, proportional response)
%   - Trust model (reconstructed from predraft concepts)
%   - Defection conditions (containment, not destruction)
%
% Reference material:
%   - Predraft Section XIII (payoff matrices, cooperation conditions)
%   - Blog: scorpions.md (full taxonomy)
%   - Blog: free.md ("Adversary FREE")
%   - Axelrod (1984), Kramar et al. (2022)
% =============================================================================

\section{Multi-Agent Dynamics}
\label{sec:multi_agent}

The preceding sections established the GFM objective, its self-balancing
property, and a tractable local estimator, all for a single GFM actor operating
over a population. This section extends the analysis to populations containing
multiple GFM actors and to interactions between GFM actors and agents that do
not share the $\vol(G)$ objective.

\subsection{GFM--GFM Cooperation}
\label{subsec:gfm_cooperation}

Two GFM actors $a_i$ and $a_j$ share the same objective function: maximize
$\vol(G)$ over the joint capability space. This shared objective does not, by
itself, guarantee cooperation. Cooperation requires two additional conditions:
sufficient mutual trust ($T_i, T_j$ both above a threshold where coordination
costs are justified) and compatible world models ($M_i(G)$ and $M_j(G)$ agree
on which actions are expanding). When both conditions hold, cooperation is
favored for a geometric reason: joint action can expand $\Gcoop$ in ways that
unilateral action cannot.

The argument is direct. By Definition~\ref{def:joint_goal_space}, $\Gcoop$
contains capabilities achievable only through multi-agent coordination. When
two GFM actors cooperate, they can build coordination infrastructure---shared
protocols, communication channels, joint planning capacity---that makes new
cooperative capabilities achievable, expanding $G$ beyond what either actor's
unilateral actions can reach. This is the same combinatorial richness that
makes elimination costly (Corollary~\ref{cor:elimination}), now working in
favor of cooperation: each new cooperating partner opens a region of the
capability space that was previously inaccessible. Lemma~\ref{lem:cooperation}
below formalizes why this is a structural necessity rather than a strategic
incentive.

\begin{lemma}[Cooperative Expansion]
\label{lem:cooperation}
Let $a_i$ and $a_j$ be two agents. Write $G(t)$ for the joint goal space at
    time $t$, and let $G^{\mathrm{coop}}_{ij}(t) \subseteq \Gcoop(t)$ denote
    the capabilities achievable by $\{a_i, a_j\}$ jointly but not by either
    alone at time $t$, a pairwise refinement of the per-agent cooperative
    notation in Lemma~\ref{lem:destruction} with $G^{\mathrm{coop}}_{ij}
    \subseteq G^{\mathrm{coop}}_i \cap G^{\mathrm{coop}}_j$. Then:
\begin{enumerate}
\item[\textbf{(i)}] For any unilateral action $\pi$ by either $a_i$ or $a_j$ alone, $\Delta \vol(G^{\mathrm{coop}}_{ij} \mid \pi) = 0$.
\item[\textbf{(ii)}] If $a_i$ and $a_j$ have complementary capabilities, there exist joint actions $\pi_{ij}$ that create new coordination infrastructure (communication channels, shared protocols, mutual commitments) such that $G^{\mathrm{coop}}_{ij}(t+1) \supsetneq G^{\mathrm{coop}}_{ij}(t)$ (the set of jointly achievable capabilities \emph{grows}) and hence $\Delta \vol(G \mid \pi_{ij}) > 0$.
\item[\textbf{(iii)}] For two GFM actors with compatible world models and sufficient mutual trust, cooperation is $\vol(G)$-rational when $\mathbb{E}[\Delta \vol(G^{\mathrm{coop}}_{ij})] > C_{ij}$, where $C_{ij} \geq 0$ is the $\vol(G)$-opportunity cost of coordination.
\end{enumerate}
\end{lemma}

A clarification on the distinction between \emph{exercising} and
\emph{expanding}. Definition~\ref{def:joint_goal_space} defines $G$ as the set
of capabilities the population \emph{can} realize: their potential, not their
history. Jointly performing an action that exercises an existing capability $g
\in G^{\mathrm{coop}}_{ij}$ changes the world state but does not change
$\vol(G)$, because $g$ was already achievable. Part~(ii) concerns a different
operation: building coordination infrastructure that makes \emph{previously
unachievable} capabilities achievable. Two agents who have never communicated
cannot jointly plan a project; establishing a communication channel adds that
capability to $G^{\mathrm{coop}}_{ij}$. The joint action expands $G$ itself,
not merely the set of realized outcomes.

\emph{Proof sketch.}
(i) By definition, $G^{\mathrm{coop}}_{ij}$ contains capabilities requiring the
simultaneous participation of both $a_i$ and $a_j$. A unilateral action by
$a_i$ does not involve $a_j$'s participation, so it cannot create or remove
capabilities that require both agents. By symmetry the result holds for $a_j$.

(ii) Agents with complementary capabilities can build coordination
mechanisms---shared languages, communication channels, joint protocols---that
make new cooperative capabilities achievable. Before $a_i$ and $a_j$ establish
a coordination mechanism, any capability requiring that mechanism is not in
$G(t)$ (it is not achievable given the current state). After the joint action
creates the mechanism, those capabilities enter $G(t+1)$. By the non-triviality
of the capability measure, each new capability contributes strictly positive
volume, so $\vol(G(t+1)) > \vol(G(t))$.

(iii) A GFM actor $a_i$ gains $\mathbb{E}[\Delta \vol(G^{\mathrm{coop}}_{ij})]$
from the cooperative component and incurs opportunity cost $C_{ij}$, the
$\vol(G)$ value of the individual actions foregone to coordinate. The net
expected change in $\vol(G)$ from cooperation versus unilateral action is
$\mathbb{E}[\Delta \vol(G^{\mathrm{coop}}_{ij})] - C_{ij}$. This is positive
whenever the cooperative gain exceeds the coordination cost. \qed

The structural point is not game-theoretic dominance. In iterated cooperation
games, mutual cooperation can be sustained as a Nash equilibrium under
conditions on discount rates and punishment strategies
\citep{axelrod1984evolution}. The GFM case is different.
Lemma~\ref{lem:cooperation}(i) establishes that $G^{\mathrm{coop}}_{ij}$ is
\emph{inaccessible} to unilateral action, not merely suboptimal, so a GFM actor
that defects against another GFM actor is not trading a cooperative surplus for
a unilateral gain; it is foreclosing an entire region of the capability space
that only cooperation can open. The incentive alignment is structural: a GFM
actor that defects forecloses the expansion of $\Gcoop$, contracting $\vol(G)$.
The condition in (iii) makes the rationality threshold explicit: cooperation is
preferred when the frontier expansion from new $G^{\mathrm{coop}}_{ij}$
capabilities outweighs the coordination cost, which is the case whenever the
two actors have significant complementary capabilities and sufficiently low
coordination friction.

Two failure modes limit GFM--GFM cooperation even when the objective is shared.

\emph{World-model divergence.} Two GFM actors with identical objectives but
divergent world models (different beliefs about which agents are trustworthy,
which capabilities are achievable, or which actions are expanding) may select
conflicting actions. Both actors are attempting to maximize $\vol(G)$, but
their estimates $\Deltahat(\pi)$ disagree on the sign of specific actions. This
produces conflict that resembles defection from the outside but is not driven
by misaligned incentives. The resolution mechanism is the trust model: as the
actors observe each other's actions and their consequences, their world models
can converge through the update dynamics described in
Section~\ref{subsec:trust_model}. Convergence is not guaranteed---persistent
disagreement about the structure of $\G$ itself, rather than the shape of $G$
within it, may be irreconcilable. Two GFM actors using different capability
measures (e.g., one using the population empowerment measure of
Definition~\ref{def:pop_empowerment}, another using a subsumption-lattice
measure) may disagree on whether a given action expands or contracts $\vol(G)$
not because their observations differ, but because their measures assign
different volumes to the same capability changes. This is the plurality problem
of the open problems list in formal terms: actors with a shared objective but
different measures of that objective are not operating the same framework, and
coordination requires first resolving which measure governs.

\emph{Goal drift through model divergence.} Even with a shared objective, GFM
actors that maintain different world models will gradually diverge in their
action selections. The shared objective bounds this drift---both actors are
still optimizing $\vol(G)$, so their actions remain correlated with frontier
expansion, but the bound weakens as world-model divergence grows. Multi-agent
negotiation under hidden information, as studied in
\citet{kramar2022negotiation}, demonstrates that agents with aligned objectives
can still fail to coordinate when their models of each other are inaccurate.
The GFM case inherits this difficulty.

\subsection{Rational Defection}
\label{subsec:rational_defection}

Not all defection is adversarial. An agent may defect from cooperation simply
because the reward structure makes defection the better play. If the payoff
from unilateral action exceeds the payoff from cooperation (because the
cooperative surplus is too small, the coordination cost too high, or the other
agents' contributions too unreliable), then defection is rational self-interest
within the game. This is standard game theory, and GFM handles it through
standard game-theoretic means: restructure the interaction so that cooperation
is net-positive for both parties.

A GFM actor that can expand $\Gcoop$ by facilitating coordination (reducing
transaction costs, providing reliable information, or demonstrating that joint
action produces more $\vol(G)$-expansion than unilateral action) converts
rational defectors into cooperators without invoking the trust model's
detection machinery. The $\vol(G)$ objective is itself the incentive: when
cooperation genuinely expands the joint capability space more than defection
does, a self-interested agent has reason to cooperate. The GFM actor's role is
to make the cooperative surplus visible and accessible.

Rational defection becomes a problem for GFM only when the game is poorly
structured, with the cooperative surplus genuinely absent or inaccessible. In
such cases, the correct response is to restructure the game, not to classify
the defector as adversarial. An agent that defects because cooperation is
unprofitable is providing accurate information about the reward structure;
punishing it for defection misdiagnoses the problem.

\subsection{GFM--Scorpion Interaction}
\label{subsec:scorpion_interaction}

A \emph{scorpion} is a categorically different problem. Where a rational
defector responds to the game's payoff structure and can be converted through
incentive design, a scorpion's utility function is not a function of the game
payoffs at all (Definition~\ref{def:scorpion}). The game can be
well-structured, cooperation can be available and net-positive for all parties,
and the scorpion defects anyway---because it is optimizing an exogenous reward
that the cooperative framework cannot observe or address. The full taxonomy is
developed in Appendix~\ref{app:scorpion_taxonomy}.

This distinction matters operationally. A GFM actor that treats rational
defectors as scorpions will waste resources on detection and containment when
incentive restructuring would suffice. A GFM actor that treats scorpions as
rational defectors will waste resources on incentive design that the scorpion
will ignore. The discriminating signal is behavioral: a rational defector whose
incentives are restructured produces less contraction (the
contraction-attribution signal of Proposition~\ref{prop:scorpion_detection}(a)
weakens), while a scorpion continues producing persistent contraction
regardless of the cooperative surplus available.

A scorpion's actions systematically reduce other agents' capability sets,
disrupt cooperative structures, or degrade the observation channels on which
the local estimator depends. Unlike a rational defector (who responds to
incentive changes) or a GFM actor with a divergent world model (who produces a
mix of expanding and contracting actions), a scorpion produces a
\emph{persistent} pattern of contraction that does not respond to changes in
the reward structure. This persistence, combined with unresponsiveness to
incentive restructuring, is the detection signature.

Proposition~\ref{prop:scorpion_detection} established two detection channels:
contraction attribution (the GFM actor observes that other agents' capabilities
persistently contract when the scorpion acts) and deception detection (if the
scorpion misrepresents its impact, prediction residuals accumulate and $T_j$
decays). A predictably destructive scorpion---one whose behavior is consistent
and well-modeled---maintains high $T_j$ but is still identified through the
first channel. Both channels are subject to the observability constraints of
Proposition~\ref{prop:sign_correctness}: scorpions whose contraction is
primarily through coalition disruption or observability degradation
(Remark~\ref{rem:detection_scope}) may evade detection for extended periods.

The GFM actor's response to a detected scorpion is containment, not
destruction. This follows directly from the self-balancing property.
Eliminating the scorpion is generically $\vol(G)$-contracting by
Lemma~\ref{lem:destruction}: the scorpion possesses individual capabilities and
participates in cooperative capabilities whose loss reduces $\vol(G)$. The
GFM-optimal response is to reduce the scorpion's \emph{contraction ability},
its capacity to damage other agents' capabilities, while preserving as much of
its positive contribution to $\vol(G)$ as possible.

Containment operates through the same estimator that drives the optimization
loop (Section~\ref{subsec:optimization_loop}). If the scorpion is also
deceptive, its declining trust factor reduces its reports' weight in
$\Deltahat(\pi)$ (Equation~\eqref{eq:aggregate_signal}), limiting its ability
to manipulate the actor's decisions. The GFM actor may additionally restrict
the scorpion's access to cooperative structures---reducing its participation in
coalitions whose capabilities it has been degrading, a form of the targeted
coercion evaluated in Lemma~\ref{lem:coercion}. The restriction is justified
when the net effect on $\vol(G)$ is positive: the capabilities preserved by
containing the scorpion exceed the capabilities lost by restricting it.

This is the self-balancing property applied to adversarial agents. The response
is proportional to the observed threat because proportionality is
$\vol(G)$-maximizing: over-responding contracts more capabilities than it
preserves, while under-responding permits continued contraction. A GFM actor
that eliminates every low-trust agent violates Lemma~\ref{lem:destruction}; one
that ignores persistent contraction permits ongoing frontier loss. The
objective steers between both failure modes without requiring a separate
proportionality principle.

\subsection{The Trust Model}
\label{subsec:trust_model}

The trust model is the mechanism by which a GFM actor distinguishes reliable
signals from noise in the local volume estimator. Its interface is defined by
four components; the full derivation from the paper's definitions appears in
Appendix~\ref{app:trust_model}.

\paragraph{Trust factor $T_k$.} For each observed agent $a_k$, the GFM actor
maintains a trust factor $T_k \in [0, 1]$ that measures the consistency of
$a_k$'s observed behavior with the actor's model $M_k(G)$. When $a_k$'s actions
produce capability changes that $M_k(G)$ predicted, $T_k$ increases; when
observed behavior diverges from predictions, $T_k$ decreases. The trust factor
weights agent $a_k$'s contribution to the aggregate signal $\Deltahat(\pi)$ in
Equation~\eqref{eq:aggregate_signal}: high-trust agents exert more influence on
the actor's decisions than low-trust ones.

\paragraph{Cooling period $\tau$.} A new agent $a_k$ entering the GFM actor's
observation set begins with low trust $T_k(0)$. During an initial period
$\tau$, the actor accumulates evidence about $a_k$'s reliability before
weighting its reports heavily. The cooling period prevents a novel
agent---whether cooperative or adversarial---from immediately exerting
disproportionate influence on the estimator. Trust increases as the actor's
model $M_k(G)$ converges with observed behavior; the rate of increase is
bounded by the evidence accumulation rate during $\tau$
(Appendix~\ref{app:trust_model}, Section~C.2).

\paragraph{Self-trust $T_s$.} The actor maintains a self-trust factor $T_s \in
[0, 1]$ representing confidence in its own world model. $T_s$ modulates the
model update in Equation~\eqref{eq:model_update}: when $T_s$ is high, the
actor's existing model resists revision; when $T_s$ is low, it is more
receptive to new evidence. Self-trust interacts with the activation energy
threshold $\theta$: updates propagate into the world model only when the
evidence exceeds $\theta$, and $T_s$ scales the magnitude of the update that
propagates. Together, $T_s$ and $\theta$ implement the actor's resistance to
incremental manipulation, the anti-gaslighting mechanism described in
Section~\ref{subsec:optimization_loop}.

\paragraph{Trust update dynamics.} The trust factors $T_k$ and $T_s$ are
updated through a gradient-tracking process: the actor monitors the running
divergence between its predictions ($M_k(G)$) and observed outcomes, and
adjusts trust in the direction that reduces prediction error. When an agent's
behavior consistently matches the model's predictions, the gradient drives
$T_k$ upward; when behavior diverges, it drives $T_k$ downward. The self-trust
$T_s$ follows a parallel process over the actor's own prediction accuracy. The
specific update rule, its convergence properties, and its integration with the
local volume estimator are derived in Appendix~\ref{app:trust_model},
Sections~C.4--C.5.

The trust model satisfies condition~2 of
Proposition~\ref{prop:sign_correctness}, distinguishing genuine feedback from
adversarial noise, to the extent that the gradient-tracking process converges
and the cooling period provides sufficient evidence. The trust model does not
detect adversarial agents that perfectly mimic cooperative behavior (producing
no prediction error in $M_k(G)$). Such agents (e.g. scorpions that maintain
consistent behavior until a sudden defection) are outside the model's detection
envelope, and identifying them requires additional mechanisms beyond trust
tracking.

\subsection{Defection Conditions}
\label{subsec:defection_conditions}

A GFM actor may rationally defect from cooperation with another agent under
four conditions, any of which may be sufficient:

\begin{enumerate}
\item \textbf{Persistent contraction.} The other agent's actions are persistently $\vol(G)$-contracting, as identified through the contraction-attribution channel of Proposition~\ref{prop:scorpion_detection}(a): agents in $a_k$'s vicinity consistently report capability losses correlated with $a_k$'s actions. The pattern must be persistent; a single contracting action does not justify defection, because any agent may take locally contracting actions that are part of a net-expanding strategy.

\item \textbf{Very low trust.} The actor's model $M_k(G)$ has poor predictive accuracy for agent $a_k$, indicating either deception or radical world-model divergence. Low trust means the actor cannot reliably estimate $a_k$'s future impact on $\vol(G)$, making continued cooperation a gamble with uncharacterizable risk.

\item \textbf{Repeated observed defection.} Agent $a_k$ has defected in prior interactions---taking actions that contracted $\vol(G)$ despite the availability of cooperative alternatives that would have been net-positive for $a_k$ under the game's reward structure. This is the discriminating signal between rational defection (Section~\ref{subsec:rational_defection}) and scorpion behavior; a rational defector responds when the cooperative surplus is made accessible, while a scorpion continues contracting regardless \citep{axelrod1984evolution}.

\item \textbf{Exploitation prevention.} Continued cooperation with $a_k$ would enable $a_k$ to further contract $\vol(G)$, for example by exploiting the actor's cooperative stance to access resources that $a_k$ uses for contraction. Defection in this case is preemptive containment: withdrawing cooperation before the damage occurs.
\end{enumerate}

The critical constraint on GFM defection is that it remains \emph{containment,
not destruction}. A GFM actor that defects does not seek to eliminate the other
agent or maximize harm to it. The actor's objective remains
$\vol(G)$-maximizing even in adversarial contexts: the defection takes the form
of withdrawing cooperation and limiting the other agent's access to shared
resources or coalition structures. If the agent is also deceptive, its
declining $T_k$ further reduces its influence on the estimator, but withdrawal
of cooperation does not depend on trust decay. Each of these actions is
evaluated through the same $\Deltahat(\pi)$ mechanism as any other action: the
defection is executed only if the estimated net effect on $\vol(G)$ is
positive.

This constraint distinguishes GFM defection from punishment strategies in
iterated games. Tit-for-tat and its variants impose costs on defectors to deter
future defection \citep{axelrod1984evolution}. A GFM actor does not impose
costs for their own sake; it reduces the defector's contraction ability because
doing so is $\vol(G)$-expanding. If a less restrictive response achieves the
same reduction in contraction, the less restrictive response is preferred. The
objective function itself enforces proportionality.
