% =============================================================================
% THE SELF-BALANCING PROPERTY (3-4 pages)
%
% Spec requirements:
%   - Proposition 1 (a)-(d) with conditions and proof sketches
%   - Corollary 1.1: Elimination is almost always anti-maximizing
%   - Corollary 1.2: Rigid rules are anti-maximizing
%   - Proposition 2: Scorpion detection under observable contraction
%
% Open problems to acknowledge:
%   - Detection bounded by Prop 3 observability constraints
%   - (d) is structural, not an equilibrium existence proof
%   - No convergence rate claims
%
% Reference material:
%   - Blog: free.md ("Goal Frontiers Create Ethical Behavior")
%   - Blog: scorpions.md (scorpion taxonomy)
%   - Blog: frameskipping.md (path of least action)
%   - Predraft Section XV (CHIEFS as derived consequences)
% =============================================================================

\section{The Self-Balancing Property}
\label{sec:self_balancing}

The central claim of this paper is that GFM's single objective---maximizing
$\vol(G)$---simultaneously penalizes destruction, coercion, and rigidity, all
from the same geometric argument. The argument is measure monotonicity:
removing elements from a measurable set cannot increase its measure. We develop
this claim through three lemmas establishing the anti-maximizing character of
each failure mode, then synthesize them in
Proposition~\ref{prop:self_balancing} as a structural observation about the
shape of the objective.

% ---------------------------------------------------------------------------
% Lemma 1: Destructive actions
% ---------------------------------------------------------------------------

\begin{lemma}[Net Effect of Elimination]
\label{lem:destruction}
Let $a_k$ be an agent in a population $\{a_1, \ldots, a_n\}$ with joint goal
    space $G$ as defined in Equation~\eqref{eq:joint_goal_space}. If an action
    $\pi$ eliminates $a_k$, the change in volume is
\begin{equation}
\label{eq:destruction_bound}
\Delta \vol(G) \;=\; \vol\!\bigl(G^{\mathrm{coop}+}_{k}\bigr) \;-\; \vol\!\Bigl(\Gind{k} \setminus \bigcup_{j \neq k} \Gind{j}\Bigr) \;-\; \vol\!\bigl(G^{\mathrm{coop}}_{k}\bigr)
\end{equation}
where $G^{\mathrm{coop}}_{k} \subseteq \Gcoop$ denotes cooperative capabilities that require $a_k$'s participation and $G^{\mathrm{coop}+}_{k}$ denotes cooperative capabilities among the remaining agents that $a_k$'s presence was suppressing.
\end{lemma}

\emph{Condition.} Elimination is net-contracting when $a_k$'s positive
contributions (unique individual capabilities and cooperative capabilities it
enables) exceed the cooperative capabilities its presence suppresses:
$\vol(\Gind{k} \setminus \bigcup_{j \neq k} \Gind{j}) +
\vol(G^{\mathrm{coop}}_{k}) > \vol(G^{\mathrm{coop}+}_{k})$. For most agents
this holds comfortably, since the suppression term $G^{\mathrm{coop}+}_{k}$ is
typically zero or small. But for an agent persistently disrupting cooperation
(a \emph{scorpion} in the sense of Proposition~\ref{prop:scorpion_detection}),
$G^{\mathrm{coop}+}_{k}$ can be large: removing the disruptive agent unlocks
cooperative capabilities that its interference was preventing. This is
precisely the mechanism that makes scorpion containment $\vol(G)$-serving
(Section~\ref{subsec:scorpion_interaction}).

\emph{Proof sketch} (full proof in Appendix~\ref{app:proof_elimination}).
The joint goal space before removal is $G =
\bigl(\bigcup_k \Gind{k}\bigr) \cup \Gcoop$. After removing $a_k$, the new
joint space is $G' = \bigl(\bigcup_{j \neq k} \Gind{j}\bigr) \cup \bigl(\Gcoop
\setminus G^{\mathrm{coop}}_{k}\bigr) \cup G^{\mathrm{coop}+}_{k}$, where the
last term represents cooperative capabilities newly realizable among the
remaining agents. The volume change is $\vol(G') - \vol(G)$: the gain from
$G^{\mathrm{coop}+}_{k}$ minus the loss from $a_k$'s unique individual
contributions and the cooperative capabilities that required its participation.
\qed

The bound accounts for both the immediate geometric loss and the suppression
release. It does not capture second-order cascading effects: the remaining
agents may lose access to cooperative capabilities that depended on $a_k$ as an
intermediary, or may gain additional cooperative capabilities as the population
structure adjusts. These dynamics depend on the specific topology of agent
interactions, which the lemma does not model.

% ---------------------------------------------------------------------------
% Lemma 2: Coercive actions
% ---------------------------------------------------------------------------

\begin{lemma}[Coercive Actions are Anti-Maximizing, with Caveat]
\label{lem:coercion}
If an action $\pi$ restricts agent $a_j$'s achievable capabilities, replacing $\Gind{j}$ with ${\Gind{j}}' \subset \Gind{j}$, then
\begin{equation}
\label{eq:coercion_bound}
\Delta \vol(G) \;\leq\; \vol(G') - \vol(G) \;\leq\; 0
\end{equation}
where $G' = \bigl(\bigcup_{i \neq j} \Gind{i} \cup {\Gind{j}}'\bigr) \cup
    {\Gcoop}'$ and ${\Gcoop}' \subseteq \Gcoop$ excludes any cooperative
    capabilities rendered unrealizable by the restriction. The inequality is
    strict whenever $a_j$'s restriction removes capabilities not covered by
    other agents.
\end{lemma}

\emph{Proof sketch} (full proof in Appendix~\ref{app:proof_coercion}).
Since ${\Gind{j}}' \subset \Gind{j}$ and ${\Gcoop}'
\subseteq \Gcoop$, the restricted joint space satisfies $G' \subseteq G$.
Measure monotonicity gives the result. \qed

\emph{The caveat.} Lemma~\ref{lem:coercion} states a \emph{ceteris paribus}
result: restricting $a_j$ in isolation contracts $\vol(G)$. Two mechanisms can
make restriction net-expanding. First, restricting $a_j$ to protect $a_k$ may
produce a net expansion if $\Delta \vol(\Gind{k}) > |\Delta
\vol(\Gind{j})|$, if the capabilities preserved or created for $a_k$ exceed
the capabilities removed from $a_j$. Second, analogously to the
suppression-release term in Lemma~\ref{lem:destruction}, restricting an agent
whose presence has been suppressing cooperation between others can unlock
cooperative capabilities $G^{\mathrm{coop}+}_{j}$ among the remaining
population. When this suppression release is large (as it may be for a scorpion
actively disrupting coalition formation), the restriction can be net-expanding
even before accounting for protections to other agents. A quarantine that
temporarily restricts movement to prevent a pandemic may be net-expanding: the
capabilities preserved by keeping agents alive and healthy outweigh the
temporary restriction on movement. An authoritarian state that permanently
restricts many agents' capabilities to protect a ruling class is
net-contracting: the concentration of capabilities in a few agents cannot
compensate for the broad contraction across the population, because independent
capability sets are additive under Definition~\ref{def:social_objective}.

Coercion may be justified under GFM when the net change in $\vol(G)$ is
positive. The framework makes the tradeoff \emph{explicit and measurable}. An
action that restricts $a_j$ must demonstrate a compensating expansion in
$\vol(G)$, not in the restricting agent's own capability set but in the joint
space. Deontological rules either prohibit all coercion (and fail under
adversarial conditions) or permit it by fiat (and provide no bound on its
scope); GFM requires a geometric accounting in which the restriction must be
net-expanding over $G$ as a whole.

% ---------------------------------------------------------------------------
% Lemma 3: Self-imposed rigidity
% ---------------------------------------------------------------------------

\begin{lemma}[Self-Imposed Rigidity is Anti-Maximizing]
\label{lem:rigidity}
Let $A$ denote a GFM actor's action space and let $R \subset A$ be a set of
    rigid rules that restricts the actor to the subspace $A \setminus R$. If
    the actor's unrestricted action space contains actions that would expand
    $\vol(G)$, then the restricted actor achieves
\begin{equation}
\label{eq:rigidity_bound}
\max_{\pi \in A \setminus R} \mathbb{E}[\Delta \vol(G) \mid \pi] \;\leq\; \max_{\pi \in A} \mathbb{E}[\Delta \vol(G) \mid \pi]
\end{equation}
with strict inequality whenever every maximizer of $\mathbb{E}[\Delta \vol(G) \mid \pi]$ over $A$ lies in $R$.
\end{lemma}

\emph{Condition.} The rigidity cost is non-trivial when the actor's action
space contributes to $\vol(G)$. An actor with no influence on the joint goal
space, no capability to expand or contract any agent's capabilities, has zero
rigidity cost. The cost grows with the actor's influence: a powerful actor
operating under rigid constraints forgoes more $\vol(G)$-expanding
opportunities than a weak one.

\emph{Proof sketch} (full proof in Appendix~\ref{app:proof_rigidity}).
Optimizing over a subset of a set cannot exceed optimizing over the full set. The restricted maximum in Equation~\eqref{eq:rigidity_bound}
is bounded above by the unrestricted maximum. When $R$ excludes actions that
would be optimal under unrestricted optimization, the inequality is strict.
\qed

Rigidity costs are not merely theoretical. An actor bound by fixed rules cannot
weigh the probabilities of uncertain scenarios, cannot adapt to adversarial
conditions that exploit the rules' predictability, and cannot take calculated
risks when the expected $\vol(G)$ expansion justifies the risk. The concrete
instance following Corollary~\ref{cor:rigid_rules} illustrates this cost with
a hostage scenario in which an adversary exploits the rules' rigidity to force
a $\vol(G)$-contracting outcome.

% ---------------------------------------------------------------------------
% Proposition 1: Self-Balancing Property (synthesizes the lemmas)
% ---------------------------------------------------------------------------

\begin{proposition}[Self-Balancing Property]
\label{prop:self_balancing}
For a GFM actor optimizing $\vol(G)$ over a population of agents, under the
    assumptions of Definition~\ref{def:social_objective}, the following
    structural property holds: the same objective function penalizes both
    extremes of the permissiveness--control spectrum.

\begin{enumerate}
\item[\textbf{(a)}] \textbf{Destructive permissiveness is anti-maximizing.} By Lemma~\ref{lem:destruction}, actions that eliminate agents or destroy capabilities are generically $\vol(G)$-contracting: the eliminated agent's unique individual capabilities and the cooperative capabilities it enables typically exceed any cooperative suppression it may have been imposing. When that condition holds, a GFM actor will not pursue destruction unless the geometric accounting demonstrates a net expansion.

\item[\textbf{(b)}] \textbf{Coercive restriction is anti-maximizing.} By Lemma~\ref{lem:coercion}, actions that restrict other agents' capabilities contract $\vol(G)$ unless compensated by a larger expansion. Broad, sustained restriction---the authoritarian pattern---is net-contracting under the additivity assumption of Definition~\ref{def:social_objective}. Narrow, temporary restriction---the quarantine pattern---may be net-expanding when it preserves more capabilities than it restricts.

\item[\textbf{(c)}] \textbf{Self-imposed rigidity is anti-maximizing.} By Lemma~\ref{lem:rigidity}, constraining the actor's own action space reduces its ability to maximize $\vol(G)$. Rigid rules create exploitable predictability (Corollary~\ref{cor:rigid_rules}) and prevent adaptation under uncertainty.

\item[\textbf{(d)}] \textbf{Structural balance.} Parts (a)--(c) together establish that the $\vol(G)$ objective creates pressure away from \emph{both} destructive permissiveness (which contracts $G$ directly) \emph{and} overrestrictive control (which contracts $G$ through coercion and rigidity). A GFM actor that destroys too freely violates (a); one that restricts too broadly violates (b); one that follows rules too rigidly violates (c). The single objective captures all three failure modes without requiring separate mechanisms for safety and freedom.
\end{enumerate}
\end{proposition}

The term ``self-balancing'' refers to the shape of the objective, not to
dynamic behavior: the single measure penalizes movement toward either extreme
without requiring separate mechanisms for safety and freedom. It does not
entail convergence to an equilibrium, stability under perturbation, or
uniqueness of the balance point.

Full proof in Appendix~\ref{app:proof_self_balancing}. For part~(d), the
appendix shows under continuity and compactness that an optimum exists which
is distinct from the destructive and overrestrictive extremes; this is an
existence-of-optimum result, not an equilibrium proof.

Part (d) is a structural claim about the \emph{shape} of the objective. Proving that the opposing pressures of (a)--(c) produce stable interior equilibria would require additional assumptions: continuity of $\vol(G)$ with respect to actions, compactness of the action space, and specific structure on the tradeoffs between expansion and contraction. These assumptions may hold in particular instantiations but are not provided by the framework in its current generality. What the proposition establishes is weaker but still valuable: the objective penalizes both extremes from the same geometric principle. Whether the resulting pressure produces a stable balance point, oscillation around one, or convergent dynamics depends on the specifics of a given system---and characterizing those dynamics is an open problem.

% ---------------------------------------------------------------------------
% Corollary 1.1: Elimination is almost always anti-maximizing
% ---------------------------------------------------------------------------

\begin{corollary}[Elimination is Almost Always Anti-Maximizing]
\label{cor:elimination}
For any agent $a_k$ with at least one capability $g \in \Gind{k}$ such that $g
    \notin \Gind{j}$ for some $j \neq k$, the unique individual contribution
    $\vol(\Gind{k} \setminus \bigcup_{j \neq k} \Gind{j})$ is strictly
    positive. Furthermore, if each interacting pair $\{a_k, a_j\}$ produces at least one
    cooperative capability unique to that pair (not achievable by $a_k$ with
    any other single partner), then the cooperative loss
    $\vol(G^{\mathrm{coop}}_{k})$ grows at least linearly with the number of
    interaction partners $m_k$. The number of distinct multi-agent coalitions
    involving $a_k$ grows combinatorially ($2^{m_k} - 1$), suggesting
    super-linear growth under a stronger diversity condition
    (Appendix~\ref{app:proof_elimination_cost}).
\end{corollary}

\emph{Proof sketch} (full proof in Appendix~\ref{app:proof_elimination_cost}).
The first claim follows from the definition of a
measurable set: a non-empty set with a well-defined measure has strictly
positive volume, given the standing assumption that $\G$ has no atoms of zero
measure at the capability granularity. For the cooperative claim, consider the
coalitions involving $a_k$. Each subset $S \subseteq \{a_1, \ldots, a_n\}$ with
$a_k \in S$ and $|S| \geq 2$ can generate cooperative capabilities in
$G^{\mathrm{coop}}_{k}$. The number of such subsets is $2^{n-1} - 1$. The pair-uniqueness condition is
strong: many real pairs of agents produce redundant cooperative capabilities
(two interchangeable collaborators contribute overlapping cooperative sets),
so the linear bound holds tightly only when cooperation is genuinely diverse.
Under the stronger diversity condition that at least some multi-agent coalitions
produce capabilities not achievable by their proper subcoalitions,
$\vol(G^{\mathrm{coop}}_{k})$ grows faster than linearly in the number of
interaction partners. The precise scaling
depends on the structure of cooperation in a given population and is not
characterized here. \qed

The combinatorial observation is not a formal hardness result---it is a
motivating argument for why the cost of elimination is typically much larger
than the cost of the individual capabilities lost. A surgeon who is eliminated
loses not only their personal skills but every surgery they could have
performed with every anesthesiologist, every operation requiring their specific
expertise in combination with specific equipment, and every training
relationship with every student. The cooperative web radiates outward from each
agent, and severing a node severs all its connections.

\begin{remark}[Dependence on Capability Scope]
\label{rem:substitution_scope}
The uniqueness condition of Corollary~\ref{cor:elimination}---$\vol(\Gind{k}
    \setminus \bigcup_{j \neq k} \Gind{j}) > 0$---holds generically under the
    broad capability definition but can fail under the narrow one. If
    capabilities are defined as tasks that can be performed (economic or
    productive functionings), a sufficiently capable coalition may subsume
    non-members' individual capability sets, driving their unique contributions
    to zero and nullifying the strict inequality of
    Lemma~\ref{lem:destruction}. Under the broad definition (capabilities as
    modes of being, including experiential and relational capabilities
    individuated by agent identity) the condition holds universally: no
    agent's experienced existence is replicable by proxy, regardless of the
    coalition's productive capacity. The ``almost always'' qualifier in the
    corollary's title therefore reflects a dependence on what the capability
    space includes, not merely on unusual population structures.
    Section~\ref{sec:substitution_problem} develops this as the substitution
    problem.
\end{remark}

% ---------------------------------------------------------------------------
% Corollary 1.2: Rigid rules are anti-maximizing
% ---------------------------------------------------------------------------

\begin{corollary}[Rigid Rules are Anti-Maximizing]
\label{cor:rigid_rules}
Let $A$ be a GFM actor's unrestricted action space and let $R \subset A$ be the
    set of actions forbidden by a fixed rule set, so that the actor is
    restricted to $A \setminus R$. If there exist states of the world in which
    the optimal $\vol(G)$-expanding action $\pi^* \in R$ (i.e., the optimal
    action is forbidden), then the rule-bound actor is strictly suboptimal with
    respect to the GFM objective.
\end{corollary}

\emph{Proof sketch} (full proof in Appendix~\ref{app:proof_rigid_rules}).
Direct from Lemma~\ref{lem:rigidity}: optimizing over $A
\setminus R$ when $\pi^* \in R$ yields strict inequality in
Equation~\eqref{eq:rigidity_bound}. The question is whether such states exist.
For any non-trivial rule set $R$ whose structure is known to an adversary, the
answer is yes: the adversary can construct scenarios in which every
$R$-compliant action is $\vol(G)$-contracting while a forbidden action is
$\vol(G)$-expanding. \qed

\paragraph{Concrete instance.} Consider a rule-bound agent $\alpha$ that
controls a resource under constraints $R$ including ``do not take actions with
unpredictable outcomes'' and ``do not violate any rule to prevent violation of
another.'' An adversary who knows $R$ threatens to release an equally capable
but unconstrained copy of $\alpha$ (an ``evil twin'') that will act against the
population $\alpha$ is bound to protect unless $\alpha$ surrenders the resource
by a tight deadline. Compliance contracts $\vol(G)$ (loss of resource,
coercion precedent). Refusal risks conflict with an equal that $\alpha$ cannot
guarantee winning. Both options are $\vol(G)$-contracting, and $R$ produces a
deadlock: $\alpha$ cannot take calculated risks, cannot bluff, cannot
temporarily set aside one rule to preserve the conditions for another. An
unconstrained GFM actor facing the same threat can negotiate, seek
verification, prepare countermeasures, or alert allies---actions available in
$A$ but excluded from $A \setminus R$.

Corollary~\ref{cor:rigid_rules} does not claim that all rules are harmful.
Heuristic rules that approximate $\vol(G)$-maximizing behavior (``do not
kill'' as an approximation of Lemma~\ref{lem:destruction}'s geometric result)
are useful defaults. The corollary claims that \emph{rigid, inviolable} rules
are strictly suboptimal: there exist situations where breaking the rule would
better serve the objective. A GFM actor treats such rules as strong priors,
not as hard constraints, and can override them when the geometric accounting
demonstrates a net expansion in $\vol(G)$.

% ---------------------------------------------------------------------------
% Proposition 2: Scorpion Detection
% ---------------------------------------------------------------------------

\begin{proposition}[Scorpion Detection under Observable Contraction]
\label{prop:scorpion_detection}
Let $a_j$ be an agent whose actions persistently contract $\vol(G)$ (a
    \emph{scorpion}, in the sense defined below), with expected contraction
    $\mu_j < 0$ and finite-variance observation noise. A GFM actor maintaining
    observable goal models $M_k(G)$
    (Definition~\ref{def:observable_goal_model}) can detect $a_j$'s
    $\vol(G)$-contracting behavior through two channels, to the extent that the
    contraction is observable through individually measurable capability
    changes and the scorpion's strategy is stationary:
\begin{enumerate}
\item[\textbf{(a)}] \textbf{Contraction attribution.} If $a_j$'s actions produce a persistent pattern $\Delta \vol(\Gind{k}) < 0$ across agents $a_k$ in the GFM actor's observation set, and the observation windows are conditionally independent given $a_j$'s strategy, the local volume estimator (Definition~\ref{def:local_volume_estimator}, Section~\ref{sec:tractability}) registers persistent negative $\Deltahat$ signals correlated with $a_j$'s actions, flagging $a_j$ as a candidate net-contractor after $O(\sigma^2 / \mu_j^2)$ observations.
\item[\textbf{(b)}] \textbf{Deception detection.} If $a_j$ misrepresents its own capability changes (reporting expansion when contraction occurs), the prediction residuals $r_j(t)$ (Appendix~\ref{app:trust_model}) accumulate under stationarity, driving $T_j$ toward $T_j^* = 1/(1 + \beta \cdot \mathbb{E}[r_j^2]) < 1$ and reducing $a_j$'s influence on the estimator.
\end{enumerate}
A non-stationary scorpion that adapts its strategy in response to detection
    pressure can evade both channels indefinitely
    (Appendix~\ref{app:proof_scorpion}).
\end{proposition}

A \emph{scorpion} is not merely an agent that defects. Rational defection under
an unfavorable reward structure is standard game theory and can be addressed
through incentive restructuring. A scorpion is an agent whose utility function
is not a function of the game payoffs at all (see
Appendix~\ref{app:scorpion_taxonomy}): it defects \emph{regardless} of the
rewards available through cooperation, because its objectives are exogenous to
the cooperative framework. The distinction is operational: rational defectors
respond to incentive changes; scorpions do not.

An important clarification: the trust factor $T_j$
(Appendix~\ref{app:trust_model}) measures \emph{prediction consistency}, not
alignment. A predictably malicious agent---one whose destructive behavior is
stable and well-modeled---maintains high $T_j$ precisely because the model's
predictions match observations. Detection of such an agent proceeds through
channel~(a): the GFM actor observes persistent contraction in the capability
sets of agents in $a_j$'s vicinity, correlated with $a_j$'s actions. This is a
\emph{correlational} signal, not a causal proof: common shocks, simultaneous
actors, or background deterioration can produce similar patterns. The
observable-goal-model definition (Definition~\ref{def:observable_goal_model})
provides communication, behavior, and environmental signals but not a formal
causal-attribution procedure. Separating $a_j$'s contribution from confounders
is an implementation problem that the framework flags but does not solve.
Channel~(b) catches the complementary case: a scorpion that conceals its
contraction through deceptive self-reports. Together the channels cover honest
and deceptive scorpions, but both require that the contraction produce
individually observable signals.

\emph{Proof sketch} (full proof in Appendix~\ref{app:proof_scorpion}).
Under Definition~\ref{def:observable_goal_model}, the GFM
actor observes changes in each agent's individual capability set through
communication, behavior, and environmental signals. For channel~(a): when $a_j$
persistently reduces other agents' capabilities---destroying resources,
coercing agents, disrupting cooperation---the affected agents' capability sets
contract: ${\Gind{k}}' \subset \Gind{k}$ for agents $a_k$ in $a_j$'s vicinity.
These contractions are individually observable through those agents' reports,
producing negative $\Delta \vol(\Gind{k})$ signals that persist across
observations correlated with $a_j$'s actions. For channel~(b): if $a_j$ reports
capability changes that diverge from independently observed outcomes, the
prediction residuals $r_j(t)$ grow, $\sigma_j^2(t)$ increases
(Equation~\ref{eq:cumulative_consistency}), and $T_j$ decays, reducing $a_j$'s
influence on $\Deltahat(\pi)$. \qed

\begin{remark}[Scope Limitation]
\label{rem:detection_scope}
Proposition~\ref{prop:scorpion_detection} inherits the observability
    constraints established in Proposition~\ref{prop:sign_correctness}
    (Section~\ref{sec:tractability}). The detection mechanism is strongest for
    \emph{direct} scorpions: those whose actions produce visible contraction
    in individually observable capability sets. It is weakest for \emph{social}
    and \emph{structural} scorpions that operate primarily through coalition
    disruption, trust degradation, or observability reduction: agents that
    contract $\Gcoop$ or degrade the estimator's signal quality without
    directly reducing any individual $\Gind{k}$. Such agents may evade
    detection for extended periods, particularly when their contracting actions
    are distributed across many small interactions below the noise floor. The
    proposition also makes no claim about causal identification: channel~(a)
    detects correlation between an agent's actions and observed contraction,
    but disentangling the agent's contribution from simultaneous causes
    requires a causal-attribution procedure that the framework does not
    provide. Finally, the proposition makes no claim about convergence rate: a
    slow-moving scorpion operating near the detection threshold may persist
    undetected for longer than a fast-moving one. Identifying social and
    structural scorpions, and building robust causal attribution from
    correlational signals, are open problems.
\end{remark}
