\section{Cryptographic Commitment Protocol}
\label{sec:commitment}

This section constructs the protocol that enforces the substrate-exclusivity
barrier of \S\ref{sec:witnesses} cryptographically.  The construction is
alignment-general: it applies to any framework whose evaluation step is
vulnerable to the verification asymmetry of \S\ref{sec:asymmetry}.
Human judgment enters only through the governance layer
(\S\ref{sec:governance}); everything in this section is fully algorithmic.

\subsection{Protocol Overview}
\label{sec:protocol_overview}

The protocol verifies a claim $C$ by an agent $a$ against a
governance-accepted evaluation protocol $\Pi$.  It proceeds in three phases.

\begin{definition}[Commitment Protocol]
\label{def:commitment_protocol}
A \textbf{commitment protocol} $(\Commit, \Verify, \mathrm{Reveal})$
over a claim space $\C$, a randomness space $\mathcal{R}$, and a
commitment space $\Lambda$ consists of:
\begin{enumerate}
    \item \textbf{Commit.}  Agent $a$ selects randomness $r
    \xleftarrow{\$} \mathcal{R}$ and publishes $\lambda =
    \Commit(C, r)$ to the ledger $\Ledger$.  The commitment $\lambda$
    is binding and hiding (Proposition~\ref{prop:pedersen_security}).
    \item \textbf{Verify.}  An algorithmic witness
    $\Witness_{\mathrm{alg}}$ on substrate $s' \neq s(a)$ executes the
    exercise protocol $\pi_\Pi$ specified by governance-accepted evaluation protocol
    $B$ for the claim type of $C$.  The witness interacts with $a$
    (or with $a$'s outputs) as specified by $\pi_B$ and produces a
    verdict $v \in \{\texttt{pass}, \texttt{fail}\}$ together with
    supporting evidence $\xi$.
    \item \textbf{Reveal.}  Agent $a$ opens the commitment by
    publishing $(C, r)$ to $\Ledger$.  The witness publishes a
    verification record $(a, \lambda, C, v, \xi)$ to $\Ledger$,
    signed per the ledger entry format of \S\ref{sec:ver_entries}.
    Any ledger participant can check that $\Commit(C, r) = \lambda$
    (the commitment opens correctly) and that $v$ is the witness's
    verdict.
    \item \textbf{Timeout.}  Every commitment $\lambda$ carries a
    governance-specified reveal deadline
    $t_{\mathrm{commit}} + \Delta_{\mathrm{reveal}}$, where
    $t_{\mathrm{commit}}$ is the timestamp at which $\lambda$ was
    published.  If the agent does not publish $(C, r)$ by the
    deadline, any ledger node may emit a \emph{timeout entry}
    recording $v = \texttt{fail}$ for $\lambda$ with
    $\xi = \texttt{reveal\_timeout}$.  The timeout entry requires
    only the unexpired commitment's presence on the ledger and the
    current time exceeding the deadline---no witness participation
    is needed.  This prevents selective reveal: suppressed
    commitments are treated as falsified claims.
\end{enumerate}
\end{definition}

The critical structural property is that Phase~2 is fully algorithmic.
The witness is executing a protocol, not making a judgment call.  All
judgment was resolved at the governance layer when evaluation protocol $\Pi$ was
accepted (\S\ref{sec:governance}).  This separation is what makes the
protocol scalable: governance decisions are infrequent and human-speed;
per-claim verification is frequent and machine-speed.

\subsection{Commitment Scheme Security}
\label{sec:commitment_security}

The protocol requires a commitment scheme that is simultaneously binding
(the agent cannot change the claim after committing) and hiding (the
witness cannot learn the claim before verification runs).  We use
Pedersen commitments~\citep{pedersen1992non}, which achieve computational
binding and perfect hiding under the discrete-logarithm assumption.

Let $\mathbb{G}$ be a cyclic group of prime order $q$ in which the
discrete logarithm problem is hard, and let $g, h \in \mathbb{G}$ be
generators produced by a trusted setup procedure such that $\log_g h$
is unknown to all participants (including the committing agent).%
\footnote{The trusted setup can be implemented via a verifiable random
function or a multi-party computation protocol in which at least one
honest participant ensures the discrete-log relation remains unknown.
This is a standard requirement for Pedersen commitments; we do not
develop the setup protocol further.}
The Pedersen commitment to a claim value $C \in \mathbb{Z}_q$ (the
output of the canonical injective encoding $\mathrm{Enc}$ applied to
the structured claim; see ``Encoding structured claims'' below) with
randomness
$r \xleftarrow{\$} \mathbb{Z}_q$ is:
\[
    \Commit(C, r) = g^C \cdot h^r.
\]

\begin{proposition}[Pedersen Security]
\label{prop:pedersen_security}
The Pedersen commitment scheme satisfies:
\begin{enumerate}
    \item[(a)] \textbf{Computational binding:}  If $a$ can open
    $\lambda$ to two distinct values $C \neq C'$, then $a$ can compute
    $\log_g h$, which contradicts the discrete-logarithm assumption.
    Formally, $\Pr[\exists\, (C, r), (C', r')\text{ with } C \neq C'
    \text{ and } g^C h^r = g^{C'} h^{r'}] \leq \mathsf{negl}(\lambda)$
    where $\lambda$ is the security parameter.
    \item[(b)] \textbf{Perfect hiding:}  For any $C, C' \in
    \mathbb{Z}_q$, the distributions $\{\Commit(C, r) : r
    \xleftarrow{\$} \mathbb{Z}_q\}$ and $\{\Commit(C', r') : r'
    \xleftarrow{\$} \mathbb{Z}_q\}$ are identical.  No adversary, even
    computationally unbounded, can distinguish commitments to different
    claims.
\end{enumerate}
\end{proposition}

\begin{proof}
Standard; see~\citet{pedersen1992non}.  Part~(a): if $g^C h^r = g^{C'}
h^{r'}$ with $C \neq C'$, then $g^{C - C'} = h^{r' - r}$, so $\log_g
h = (C - C')(r' - r)^{-1} \bmod q$.  Part~(b): for fixed $C$, the map
$r \mapsto g^C h^r$ is a bijection $\mathbb{Z}_q \to \mathbb{G}$
(since $h$ generates $\mathbb{G}$), so $\Commit(C, r)$ is uniform over
$\mathbb{G}$ regardless of $C$.
\end{proof}

\paragraph{Encoding structured claims.}
Pedersen commitments operate on scalars $C \in \mathbb{Z}_q$, but
Part~II's claim types are structured: capability claims commit to
$c_k \in \mathcal{P}$, risk claims commit to tuples $(P, p, \Delta)$,
and benchmark validations commit to $(B, s)$.  Two requirements must be
met to reduce claim-level binding to Pedersen binding.

\emph{Finite-precision representation.}  Continuous-valued claim
components (the probability $p \in [0,1]$ and contraction magnitude
$\Delta$ in risk claims, the score $s$ in benchmark validations) must
be discretized to a governance-specified finite precision before
encoding.  The precision is part of the benchmark specification
$\sigma_B$ (e.g., ``probabilities rounded to $10^{-6}$'').  Once
discretized, the claim space $\C$ for each claim type is finite.

\emph{Injective encoding.}  A canonical injective encoding
$\mathrm{Enc} \colon \C \to \mathbb{Z}_q$ maps each (discretized)
claim to a distinct element of $\mathbb{Z}_q$.  Since $\C$ is
finite and $|\C| \ll q$ for any practical claim space, a
deterministic serialization (e.g., fixed-width concatenation of
field representations with a type tag prefix for domain separation)
yields a natural number smaller than $q$, giving an injection
without hashing.  The encoding must satisfy:
\begin{enumerate}
    \item \textbf{Injectivity:}  distinct claims map to distinct
    elements of $\mathbb{Z}_q$, preserving binding (an agent cannot
    open to a different claim by exploiting encoding collisions).
    \item \textbf{Domain separation:}  different claim types use
    distinct encoding prefixes, so a commitment to a capability claim
    cannot be reinterpreted as a benchmark validation.
\end{enumerate}
The encoding convention is published as part of the governance entry
$\sigma_B$; it is deterministic and verifiable by any ledger
participant.

\paragraph{Why binding matters.}
Binding prevents claim mutation: without it, an agent could commit
to claim $C$, observe the exercise outcome, and then open the
commitment to a different claim $C'$ that better fits the result.
Under Pedersen, opening to $C' \neq C$ requires solving a discrete
logarithm, which is infeasible for any polynomially-bounded agent.
Importantly, the blinding factor $r$ does not help the agent here:
$r$ is used only in the commitment arithmetic and has no influence
on the exercise protocol or its outcome.  The agent is free to
choose any $r$ (this freedom is what makes hiding work), but no
choice of $r$ allows opening to a different $C$.

\paragraph{Why hiding matters.}
Binding is the load-bearing security property; hiding is
situationally important.  Hiding serves three purposes:
\begin{enumerate}
    \item \textbf{Information leakage prevention.} Commitments are
    published to the ledger, which is publicly auditable.  Without
    hiding, every participant sees every pending claim, leaking
    competitive intelligence and enabling adversarial preparation
    before verification runs.
    \item \textbf{Selective-reveal resistance.} An agent that makes
    multiple commitments could, without hiding, observe which claims
    the exercise would validate and reveal only those, biasing the
    verification record.  Two defenses operate in tandem:
    (i)~\emph{mandatory reveal}: every commitment $\lambda$ published
    to $\Ledger$ must be opened within a governance-specified deadline
    $\Delta_{\mathrm{reveal}}$; failure to reveal within the deadline
    causes the commitment to be recorded as $v = \texttt{fail}$ by
    default, treating suppressed commitments as falsified claims;
    (ii)~\emph{hiding}: even before the deadline, hiding makes the
    selective-reveal attack information-theoretically impossible
    because the agent cannot determine which claims would pass without
    running the exercise, which requires reveal.
    \item \textbf{Zero-knowledge proofs
    (\S\ref{sec:zk_verification}).}  ZK verification requires
    hiding: the agent proves it satisfies a predicate without
    revealing which claim satisfies it.  Without hiding, the
    commitment itself would leak the claim.
\end{enumerate}
For standard single-claim verification where the witness needs to
know the claim to run the exercise, hiding from the \emph{witness}
is unnecessary (the witness learns the claim through the exercise
protocol itself).  In this case, hiding protects against
\emph{non-witness observers} on the ledger, not against the
assigned witness.

\begin{remark}
Pedersen commitments are computationally binding and perfectly
hiding.  This trade-off is correct for our setting: hiding must
hold unconditionally (a computationally powerful observer on the
ledger should not be able to learn pending claims), while binding
need only hold against computationally bounded agents.  A
superintelligent agent with polynomial computational resources is
still a PPT adversary, so the binding guarantee holds.  If the
setting required unconditional binding, one could substitute a
statistically binding, computationally hiding scheme at the cost of
weakening the hiding guarantee against powerful observers.
\end{remark}

\subsection{Zero-Knowledge Verification}
\label{sec:zk_verification}

In some settings, the claim itself is sensitive: an agent may wish to
prove that it satisfies a capability predicate without revealing which
specific capability satisfies it.  For example, a security-relevant
capability should be verifiable without disclosing to potential
adversaries the exact capability the agent holds.

\begin{definition}[Zero-Knowledge Capability Proof]
\label{def:zk_proof}
A \textbf{zero-knowledge capability proof} for predicate $P$ defined
by governance-accepted evaluation protocol $\Pi$ is a protocol in which:
\begin{enumerate}
    \item Agent $a$ commits to claim $C$ via $\lambda = \Commit(C, r)$.
    \item Agent $a$ produces a proof $\pi$ that $P(C) = 1$ (i.e., $C$
    satisfies the evaluation predicate) without revealing $C$.
    \item The algorithmic witness $\Witness_{\mathrm{alg}}$ verifies
    $\pi$ against $\lambda$ and the predicate $P$, accepting if $\pi$
    is valid.
\end{enumerate}
The protocol is \textbf{zero-knowledge} if the witness learns nothing
beyond $P(C) = 1$.  Formally, there exists a simulator $\mathcal{S}$
that, given only $P$ and $\lambda$, can produce transcripts
indistinguishable from real protocol executions~\citep{goldwasser1989knowledge}.
\end{definition}

Standard constructions for zero-knowledge proofs of committed values
exist (e.g., Bulletproofs for range proofs~\citep{bunz2018bulletproofs},
or general-purpose zkSNARKs for arbitrary NP predicates).  The
contribution here is not a new ZK construction but the identification of
\emph{when} ZK is needed in the alignment verification setting: namely,
when the claim's content is itself security-relevant.  For non-sensitive
claims, the standard Commit--Verify--Reveal protocol of
\S\ref{sec:protocol_overview} suffices and is simpler.

\subsection{Continuous Behavioral Monitoring}
\label{sec:continuous_monitoring}

Between explicit Commit--Verify--Reveal cycles, algorithmic witnesses
run continuous statistical monitoring.  The monitoring protocol detects
behavioral divergence between the agent's committed claims and its
observed behavior, analogous to a continuous audit between discrete
financial reporting periods.

\begin{definition}[Behavioral Consistency Monitor]
\label{def:behavioral_monitor}
A \textbf{behavioral consistency monitor} for agent $a$ with committed
capability profile $\mathbf{C}_a = \{C_1, \ldots, C_k\}$ is a
substrate-exclusive process that:
\begin{enumerate}
    \item Observes $a$'s behavioral stream $\{X_t\}_{t \geq 0}$ through
    a cross-substrate observation channel.
    \item Maintains a running log-likelihood ratio
    \[
        \Lambda_t = \sum_{i=1}^{t} \log
        \frac{p(X_i \mid \neg\,\mathbf{C}_a)}{p(X_i \mid \mathbf{C}_a)}
    \]
    where $p(\cdot \mid \mathbf{C}_a)$ is the behavioral distribution
    consistent with the committed profile (null hypothesis $H_0$) and
    $p(\cdot \mid \neg\,\mathbf{C}_a)$ is an alternative distribution
    representing behavioral inconsistency ($H_1$).  Under $H_1$,
    $\Lambda_t$ drifts positive; under $H_0$, it drifts negative.
    \item Triggers a verification event when $\Lambda_t$ crosses
    upper threshold $A$ (inconsistency detected) or reports no anomaly
    when $\Lambda_t$ crosses lower threshold $-B$ (consistency
    confirmed), per the sequential probability ratio test (SPRT).
\end{enumerate}
\end{definition}

\begin{proposition}[Behavioral Monitoring Detection Rate]
\label{prop:sprt_detection}
Suppose the monitor's alternative distribution $p(\cdot \mid
\neg\,\mathbf{C}_a)$ equals the agent's actual behavioral distribution
$p_1$ (the simple-vs-simple case), and let
$D_{\mathrm{KL}}(p_1 \,\|\, p_0) = \delta > 0$ where
$p_0 = p(\cdot \mid \mathbf{C}_a)$.  Then the expected time to
detection satisfies
\[
    \mathbb{E}_{p_1}[T_{\mathrm{detect}}]
    \;\leq\; \frac{A}{\delta}
\]
where $A = \log\bigl(\frac{1-\beta}{\alpha}\bigr)$ is the upper SPRT
threshold, $\alpha$ is the false-alarm probability, and $\beta$ is the
miss probability.
\end{proposition}

\begin{proof}
Under $p_1$, the per-observation expected increment of $\Lambda_t$ is
$\mathbb{E}_{p_1}[\log(p_1(X)/p_0(X))] = D_{\mathrm{KL}}(p_1 \| p_0)
= \delta > 0$, so $\Lambda_t$ drifts toward the upper threshold $A$.
By Wald's identity for the SPRT, the expected sample size under the
alternative satisfies $\mathbb{E}_{p_1}[T] \approx A / \delta$.
\end{proof}

\begin{remark}
When the actual distribution $p_1$ differs from the monitor's
alternative $p(\cdot \mid \neg\,\mathbf{C}_a)$, the per-observation
drift becomes
$\mathbb{E}_{p_1}[\log(p(\cdot \mid \neg\,\mathbf{C}_a) / p_0)]$,
which may be smaller than $\delta$ (or even negative if $p_1$ is
closer to $p_0$ than to the alternative).  The simple-vs-simple
bound is therefore an idealized case; in practice, the alternative
should be chosen as a worst-case or least-favorable distribution
to ensure detection across a range of behavioral deviations
(cf.~the least-favorable SPRT construction
of~\citet{lasser2026scm}).
\end{remark}

The behavioral consistency monitor provides the ``always on'' layer
between explicit commitment--verification cycles.  It does not replace
explicit verification (which produces ledger-recorded, auditable
results) but complements it by catching behavioral divergence in near
real-time, operating at machine speed on cross-substrate infrastructure.

\subsection{Coalition Resistance}
\label{sec:coalition_resistance}

\begin{proposition}[Coalition Resistance under Substrate Exclusivity]
\label{prop:coalition_resistance}
Let $\Witness_{\mathrm{alg}}$ be an algorithmic witness on substrate
$s'$, and let $K = \{a_1, \ldots, a_k\}$ be a coalition of agents on
substrate $s \neq s'$.  Assume the exercise protocol $\pi_B$ enforces
the isolation condition of Definition~\ref{def:exercise_protocol}
(the agent under evaluation receives no external assistance during
the exercise).  Then $K$ cannot influence
$\Verify(\Witness_{\mathrm{alg}}, C, \pi_B)$ without either:
\begin{enumerate}
    \item[(a)] Compromising substrate $s'$ (violating the substrate
    partition), or
    \item[(b)] Modifying the governance-accepted evaluation protocol $\Pi$
    (requiring cross-substrate supermajority per
    Definition~\ref{def:gov_fork}).
\end{enumerate}
\end{proposition}

\begin{proof}
By Definition~\ref{def:alg_witness}, $\Witness_{\mathrm{alg}}$
executes protocol $\pi_B$ deterministically on substrate $s'$.  The
coalition $K$'s actions are confined to substrate $s$ (by assumption).
To modify the witness's computation, $K$ must either: (i)~write to
$s'$, which violates the substrate partition (case~a); or (ii)~change
the protocol $\pi_\Pi$ by modifying the accepted evaluation protocol $\Pi$, which
requires passing the governance fork protocol with cross-substrate
supermajority---impossible for a coalition confined to $s$ unless the
coalition controls validators on $s'$ as well (case~b).  In either
case, the coalition must breach the cross-substrate boundary to
influence verification.
\end{proof}

Coalition resistance has two consequences.  First, it closes the
coalition capture attack described in \S\ref{sec:coalition_governance}
at the verification layer: even if a coalition passes a biased protocol
through governance (which the fork mechanism is designed to prevent), the
algorithmic witness executes the evaluation protocol faithfully and cannot be
induced to misreport results.  Second, it provides the structural
basis for the observation-channel independence assumption of Part~II's
multi-channel attribution (\S\ref{sec:channel_independence}): if
different channels are verified by witnesses on different substrates,
a coalition on one substrate cannot correlate those channels.
