\section{Conclusion}
\label{sec:conclusion}

Alignment frameworks address specification, generalization, and
verification.  This paper has argued that verification is the
prerequisite layer: when the evaluator is endogenous---reachable from
the agent's action space---solutions to the other two layers do not
bind, because the channel through which their success would be measured
is the channel that has failed.

The verification asymmetry has a precise form
(Definition~\ref{def:ver_asymmetry}): the ratio of evaluator inputs
reachable by the agent to the total evaluator input space.  This ratio
grows monotonically with capability
(Proposition~\ref{prop:monotonicity}), making the verification problem
strictly harder for more capable agents.  The only structural barrier
invariant under capability scaling is substrate exclusivity: the agent
cannot write to observation channels rooted in substrates it does not
occupy.

We constructed a cryptographic commitment protocol
(Definition~\ref{def:commitment_protocol}) that enforces the
substrate-exclusivity barrier without requiring trust: binding
commitments prevent retroactive claim modification, hiding prevents
premature claim disclosure (with an optional zero-knowledge variant
for security-sensitive claims), and algorithmic witnesses
(Definition~\ref{def:alg_witness}) execute fully specified verification
protocols on separate substrates.  Human judgment enters at a single
point---benchmark admissibility through a governance fork protocol
(Definition~\ref{def:gov_fork})---resolving the bandwidth problem by
separating the infrequent, human-speed governance decision from the
frequent, machine-speed per-claim verification.

Within the GFM framework, the construction closes three open issues.
First, it provides the observation-channel integrity that
multi-channel attribution~\citep{lasser2026scm} assumes: channel
isolation (Proposition~\ref{prop:channel_isolation}) holds
adversarially under substrate exclusivity alone, and full conditional
independence (Proposition~\ref{prop:channel_independence}) holds under
additional sufficiency assumptions.  Second, it upgrades
SPRT-based scorpion detection from actor-estimated baselines to committed
baselines that, when the committed profile is at least as discriminating
as the actor's estimate (KL-dominance), provide at least equal detection
power (Proposition~\ref{prop:committed_detection}).  Third, the verification
impossibility theorem (Theorem~\ref{thm:anti_monopolar}) establishes
that a monopolar population loses not only the $\volL$ safety
guarantee~\citep{lasser2026horizon} but the ability to verify whether
that guarantee holds, a compound failure that is both dangerous and
silent.

The construction is not GFM-specific.  Any alignment framework whose
evaluation step is vulnerable to the write-access asymmetry---which, as
\S\ref{sec:existing_frameworks} argues, includes all major current
frameworks---can use substrate-exclusive witnesses and cryptographic
commitments to exogenize its verification layer.  The framework-specific
work is in defining the claim types and exercise protocols; the
structural argument for why exogenous verification is needed, and the
protocol that provides it, are universal.

Verification does not solve alignment.  It is the infrastructure
that makes solutions to specification and generalization
\emph{checkable}.  A safety guarantee that cannot be verified is not a
guarantee.
