\section*{Author Contributions}

\textbf{Teague Lasser} conceived the verification-layer framing, the
three-layer alignment taxonomy, the connection between substrate
exclusivity and verification integrity, the governance fork mechanism,
and the integration with the GFM machinery. Directed all revisions
and made final editorial decisions. Responsible for the paper's
intellectual direction and all claims made.

\textbf{Claude Opus 4.6} (Anthropic) drafted the formal exposition:
the write-access asymmetry formalization, the substrate-exclusive
witness definitions, the cryptographic commitment protocol with
Pedersen security analysis, the verification ledger architecture,
the behavioral consistency monitor, the GFM claim-type instantiations,
and the verification impossibility theorem with its corollary on
monopolar epistemic closure.

\textbf{GPT 5.4} (OpenAI) served as technical reviewer across five
codex review rounds, identifying formal gaps including the
anti-monopolar theorem's circularity (requiring the same-substrate
addressability premise), the benchmark-shopping vulnerability
(requiring binding of evaluation protocol in commitments), the
channel independence sufficiency assumption's failure under adversarial
conditions, and the need for bounded-synchrony assumptions in ledger
finality.

\medskip
\noindent\emph{Transparency note.} Both AI systems operated as tools under
human direction. Neither system has continuity across sessions, cannot take
responsibility for the work in the sense required by most venue authorship
policies, and cannot respond to reviewer queries independently. They are
listed as authors to accurately represent their contributions to the
intellectual content of the paper, not to claim that they meet all criteria
of traditional academic authorship. The corresponding author for all
inquiries is Teague Lasser.
