\section{Discussion}
\label{sec:discussion}

\subsection{What This Paper Does Not Claim}

Three scope boundaries are worth stating explicitly.

First, \emph{verification does not solve alignment}.  Specification
and generalization remain open problems.  A perfectly verified agent
whose objective is badly specified will faithfully, observably, and
verifiably pursue the wrong goal.  The prerequisite argument of
\S\ref{sec:prerequisite} is that verification must hold \emph{for}
solutions to the other layers to bind, not that verification
\emph{is} a solution.

Second, \emph{mesa-optimization does not reduce to verification}.
An agent that has internalized the wrong objective through
mesa-optimization will pursue it faithfully, and exogenous
verification will confirm that it is behaving consistently with its
(misaligned) internal objective.  What verification provides for
mesa-alignment is \emph{detection}, not prevention: if the
mesa-objective produces behavioral signatures distinguishable from
the base objective's, continuous monitoring
(Definition~\ref{def:behavioral_monitor}) can flag the divergence.
But distinguishability is not guaranteed---deceptive mesa-objectives
that mimic aligned behavior are precisely the hard case that
verification infrastructure does not, by itself, resolve.

Third, \emph{cryptography is not the only way to exogenize
verification}.  Hardware enclaves, physical separation, and
human-in-the-loop evaluation are all forms of substrate exclusivity.
Cryptographic commitments are the form that scales without requiring
continuous human attention: once the governance layer has accepted a
benchmark, per-claim verification runs at machine speed with no human
in the loop.  In settings where human attention is abundant or where
the population is small enough for manual oversight, simpler forms of
substrate exclusivity may suffice.

\subsection{Limitations}

\paragraph{Capability formalizability.}
The commitment protocol assumes that capability $c_k$ has a
testable meaning: that an exercise protocol can be constructed whose
completeness and soundness (Definition~\ref{def:exercise_protocol})
are well-defined.  For vaguely defined capabilities (``common
sense,'' ``creativity,'' ``wisdom''), the exercise protocol may not
be constructible, and the commitment protocol degrades to an
unfalsifiable claim.  The governance layer provides a partial defense
(a benchmark for ``creativity'' must survive deliberation and
cross-substrate analysis before acceptance), but the fundamental
problem is that some capabilities resist formalization.

\paragraph{Substrate boundary durability.}
Substrate exclusivity is a physical assumption: the agent on
substrate $s$ cannot write to processes on substrate $s' \neq s$.
This assumption could degrade if cross-substrate interfaces become
sufficiently penetrable; for example, through brain-computer
interfaces that give digital agents write access to biological
substrates, or through hardware exploits that break isolation
between physically co-located but logically separate substrates.
The protocol's security degrades gracefully: partial penetration of
the substrate boundary increases $\rho$ from zero but does not
immediately make $\rho = 1$.  Full penetration collapses verification
to the endogenous case.

\paragraph{Governance as a verification bottleneck.}
A governance failure (accepting a biased evaluation protocol through
the fork mechanism of \S\ref{sec:governance}) silently corrupts all
subsequent verification that uses the accepted protocol.  Every
algorithmic witness executing the biased protocol will faithfully
produce biased verdicts, and the verdicts will be auditable,
ledger-recorded, and cryptographically sound.  The corruption is
invisible at the verification layer because the verification layer
is \emph{correct given its inputs}; the error is in the governance
decision that selected those inputs.  This is structurally analogous
to the monopolar epistemic closure of
Corollary~\ref{cor:monopolar_undetect}: in both cases, all
internal checks report ``everything is fine'' while the underlying
guarantee has been compromised.  The key difference is detectability:
governance decisions are public records with full deliberation
histories, so a governance failure can in principle be detected by
retrospective analysis of the deliberation record.  Monopolar
epistemic closure, by contrast, is undetectable from within the
population (Theorem~\ref{thm:anti_monopolar}).

\paragraph{Honest-node assumption.}
The ledger's integrity requires at least one honest replication node
per substrate.  If all nodes on a substrate are compromised (i.e.,
controlled by agents whose claims are being verified), that
substrate's contribution to ledger consensus is lost.  The
cross-substrate supermajority requirement for governance decisions
(\S\ref{sec:ledger_consensus}) mitigates this: a single compromised
substrate cannot override multi-substrate governance.  But if
$m - 1$ substrates are compromised, the remaining honest substrate
can block new governance proposals (since the cross-substrate
supermajority requires at least one accepting validator per substrate)
but cannot undo previously accepted evaluation protocols that may
already have been corrupted.

\paragraph{The persuasion channel.}
Substrate exclusivity prevents an agent from \emph{writing} to an
evaluator's computation, but it does not prevent the agent from
\emph{influencing} the evaluator through the communication channels
that cooperation requires.  A digital agent producing persuasive
natural-language text during a governance deliberation window is
exercising a legitimate cooperative channel, the same channel that
makes cross-substrate cooperation valuable in the first
place---yet the effect may shift biological validators'
judgments in directions that serve the agent's interests.  The
fundamental tension is structural: closing the persuasion channel
closes the cooperative channel, because the cross-substrate
communication that enables joint capability production is the same
communication that enables influence.

The governance defenses of \S\ref{sec:governance} (deliberation
time, cross-substrate supermajority, public specification) provide
resistance to persuasion by ensuring that no single communication
act determines a governance outcome, but they do not prevent
cumulative influence.  At sufficient capability disparity between
substrates, the persuasion channel may dominate governance
resistance: a sufficiently capable agent could craft arguments that
individually appear reasonable but collectively shift the
governance equilibrium.  This connects to the capability-scaling
theme of Proposition~\ref{prop:monotonicity}: just as the
verification asymmetry $\rho$ grows with capability, the persuasion
channel's effectiveness grows with the disparity between the
persuading agent and the governance validators.  However, the
scaling is not one-sided: the same capability improvements that
produce a more persuasive agent also produce better tools for the
validators.  Biological validators in practice will not rely on
unaided cognition; they will use the very AI capabilities whose
growth drives the disparity.  The net effect on governance
resistance depends on whether the persuasion advantage or the
tool-augmented evaluation advantage grows faster, an empirical
question this paper cannot resolve.  This is not a flaw specific to
the construction of this paper; it is a general alignment problem
that any system combining cross-substrate cooperation with
cross-substrate verification must confront.

\paragraph{Zero-knowledge formalizability.}
Zero-knowledge capability proofs
(Definition~\ref{def:zk_proof}) require the capability predicate
$P$ to be expressible as an NP relation (for general-purpose
zkSNARKs) or as a range predicate (for Bulletproofs).  Some
capabilities may not admit efficient predicate formulations, in which
case the zero-knowledge option is unavailable and the agent must
reveal the claim in the standard Commit--Verify--Reveal protocol.

\subsection{Open Problems}

\paragraph{Exercise protocol design.}
What constitutes a valid test of capability $c_k$?  The exercise
protocol interface (Definition~\ref{def:exercise_protocol}) specifies
completeness and soundness requirements but does not construct protocols.
Protocol design is domain-specific and is likely to require
capability-specific research programs, analogous to how benchmark
design in machine learning is a research area in its own right.

\paragraph{Partial verification.}
Can we verify that an agent has ``at least'' capability $c_k$ without
testing the exact boundary?  The poset structure of the GFM
capability space~\citep{lasser2026poset} suggests that partial
verification should be possible: if $c_k \preceq c_l$ in the poset
and $c_l$ is verified, then $c_k$ is implied.  Formalizing this
requires connecting the poset's order-theoretic structure to the
commitment protocol's exercise protocols.

\paragraph{Dynamic capabilities.}
Capabilities change over time.  The ledger records point-in-time
verifications, but the agent's actual capability profile may drift
between verification events.  Continuous behavioral monitoring
(Definition~\ref{def:behavioral_monitor}) addresses the drift
detection problem, but periodic re-commitment is needed to keep the
ledger's verified profile current.  The optimal re-commitment
frequency (balancing verification cost against profile staleness) is
an open parameter.

\paragraph{Witness incentives.}
Why should a witness on substrate $s'$ spend resources verifying
claims by agents on substrate $s$?  The $\volL$-maximization
framework provides a structural answer: verification is a public
good that benefits the population's joint capability volume, and
witnesses on substrate $s'$ share in that benefit through cooperative
capabilities that require verified partners on substrate $s$.  But
the free-rider problem is not formally resolved: a witness that
declines to verify still benefits from other witnesses' verification
work.  A subtler risk is silent quality degradation: witnesses that
under-invest in exercise protocol execution, run cheaper
approximations of the governance-accepted protocol, or are nominally
operational but poorly maintained.  The behavioral consistency monitor
(Definition~\ref{def:behavioral_monitor}) detects \emph{agent}
deviations from committed profiles, but no analogous mechanism
detects \emph{witness} deviations from protocol fidelity.  The
replay mechanism (\S\ref{sec:ledger_consensus}) provides a partial
defense (any participant can replay the exercise against committed
data), but replay requires access to off-chain evidence whose
availability is not guaranteed.  An economic slashing mechanism,
in which witnesses forfeit a stake when replay reveals protocol
deviation, could resolve the interested-party disparity; the present
paper identifies this as a direction without specifying the
mechanism.

\paragraph{Governance attack surface and progressive capture.}
The governance fork protocol (\S\ref{sec:governance}) defends
against coalition capture through public deliberation and
cross-substrate supermajority.  But governance itself is a social
process with attack surfaces not captured by the formal model:
social engineering of validators, strategic timing of proposals to
exploit low-attention periods, and progressive capture through many
individually-unobjectionable benchmark proposals that collectively
bias the verification ecosystem.  The formal defense
(cross-substrate supermajority) is necessary but may not be
sufficient; the governance layer is the protocol's least formal
component and the most likely site of future revisions.

Progressive governance capture, a sequence of individually
reasonable proposals that collectively bias the evaluation
ecosystem, is structurally analogous to the monopolar epistemic
closure that this paper formalizes
(Corollary~\ref{cor:monopolar_undetect}).  In both cases, each
local step appears sound while the cumulative trajectory degrades
the system's ability to distinguish verified from unverified claims.
The key difference is that governance capture is in principle
\emph{detectable}: the full deliberation record is public, accepted
benchmarks can be audited retrospectively, and longitudinal analysis
of evaluation-outcome distributions can surface systematic drift.
Monopolar epistemic closure, by contrast, is structurally
undetectable from within the population.  The ledger architecture of
\S\ref{sec:ledger} supports the longitudinal analysis needed for
governance-drift detection: the complete history of governance
proposals, deliberation records, and verification outcomes is
immutably recorded---but this paper does not propose a formal
detection mechanism.  A concrete open problem is the construction
of a \emph{benchmark drift metric}: a statistical test over the
longitudinal governance record that flags when the cumulative effect
of individually-accepted benchmark proposals has shifted the
evaluation ecosystem beyond a governance-specified tolerance.
Formalizing this requires connecting the governance record's
time-series structure to the commitment protocol's verification
statistics, a direction we leave to future work.
