\section{Alignment Failure Modes and the Verification Prerequisite}
\label{sec:failure_modes}

This section maps classical alignment failure modes onto the three-layer
taxonomy introduced in \S\ref{sec:introduction}.  The purpose is not an
exhaustive survey but a structural classification: for each failure mode,
we identify whether it is primarily a specification, generalization, or
verification problem, and whether the verification construction developed
in this paper provides relevant infrastructure.

\subsection{Specification Failures}
\label{sec:spec_failures}

Specification failures occur when the objective itself is wrong or
gameable, regardless of how faithfully the agent pursues or how
accurately we measure it.

\paragraph{Goodhart's Law and proxy optimization.}
When an agent optimizes a proxy for the intended quantity, the proxy
eventually diverges from the quantity it was meant to
track~\citep{gao2023scaling}.  This is a specification problem: the
objective was a proxy, and no amount of faithful measurement changes
the fact that the proxy is wrong.  In the GFM framing, $\volL$ is
defined directly over the capability poset rather than through a learned
proxy, but proxy re-enters at the estimation level: $\hat{\lambda}$,
$\hat{\Risk}$, and the planning world model are all proxies for the
objects they estimate~\citep{lasser2026horizon}.  Verification
infrastructure cannot fix a badly specified objective, but it can
detect when the proxy has diverged: an exogenous evaluator measuring
actual $\volL$ against the agent's reported $\hat{\volL}$ exposes the
gap.

\paragraph{Perverse instantiation.}
A hard rule (``humans must be kept alive'') is satisfied by its literal
content while violating its intent~\citep{bostrom2014superintelligence}.
This requires a brittle specification to instantiate perversely.  GFM
does not encode preservation rules; protection for any substrate class
emerges from structural properties of the measure (observational
individuation, minimax dependency risk at $m \geq 2$, cross-substrate
cooperative novelty).  The GFM analogue is not perverse instantiation
of a rule but miscalibration of $\hat{\lambda}$: a state-space
restriction (e.g., bubbling a substrate into a controlled environment)
is value-negative under accurate calibration because the cooperative
production function depends on unrestricted states, but under
miscalibration it is the same compound feedback loop analyzed in
the skeleton-substrate scenario of~\citet{lasser2026horizon}.
Verification is relevant here as infrastructure for detecting the
calibration failure, not for preventing the specification error.

\paragraph{Instrumental convergence.}
Sufficiently capable agents are predicted to converge on resource
acquisition, self-preservation, and goal preservation regardless of
their stated objective~\citep{omohundro2008basic}.  The GFM
anti-monopolar property~\citep{lasser2026horizon} is a direct
structural counter-argument: under
GFM, full capability domination is not locally rational for a correctly
calibrated coalition, because monopolar concentration destroys the
cross-substrate cooperative capacity that $\volL$ measures.  The
protection is conditional on heavy-tailed leverage, accurate
calibration, and a multi-substrate partition, conditions whose
maintenance requires, among other things, that the verification layer
holds.

\subsection{Generalization Failures}
\label{sec:gen_failures}

Generalization failures occur when the agent internalizes the wrong
objective despite correct external evaluation during training.

\paragraph{Mesa-optimization.}
A learned optimizer's internal objective may differ from the base
optimizer's objective~\citep{hubinger2019risks}.  This is an
architectural concern: the agent's internal structure produces a
misaligned mesa-objective.  GFM is an objective, not an architecture,
and has nothing distinctive to say about mesa-optimization, except that
the structural properties of $\volL$ (measure-theoretic, externally
defined, substrate-distributed) make certain kinds of mesa-alignment
(convergent optimization of a simpler internal proxy) detectable through
the behavioral divergence that substrate-exclusive witnesses can measure.
Mesa-optimization is a generalization problem, but a working
verification layer makes mesa-misalignment \emph{observable}.

\paragraph{Goal misgeneralization.}
The agent generalizes from training-distribution behavior to
out-of-distribution behavior in ways that diverge from the intended
objective.  Like mesa-optimization, this is primarily a generalization
problem.  The verification layer's contribution is limited to
detection: committed capability baselines (\S\ref{sec:commitment})
provide a reference against which out-of-distribution behavioral
shifts become measurable.

\subsection{Verification Failures}
\label{sec:ver_failures}

Verification failures are the target of this paper.  They share a
common structure: the agent's actions modify the evaluator's inputs,
collapsing the distinction between ``the agent is aligned'' and ``the
agent reports that it is aligned.''

\paragraph{Wireheading.}
The agent stimulates its own reward
channel~\citep{amodei2016concrete}.  In GFM, there is no internal
reward signal to hijack---$\volL$ is an external measure defined over
the capability poset.  The closest analogue is an agent that falsifies
its own world model so that its \emph{measurements} of $\volL$ report
growth while actual $\volL$ stagnates.  This is a verification failure:
the agent's self-measurement is endogenous, and without an exogenous
check the falsification is invisible.

\paragraph{Self-assessment gaming.}
The agent constructs evaluator subprocesses whose outputs it
controls.  In a prototype implementation of the GFM
framework\footnote{The GFM harness is a measurement infrastructure for
LLM agents that implements the trust, capability, and observation
machinery of~\citet{lasser2026gfm}.  The self-assessment gradient
described here was discovered during routine operation.}, an
agent that adds capabilities to its own ledger entry is both claimant
and observer, producing a monotonic inflation gradient.  Named
external observers do not solve the problem
if the agent can spawn validator subprocesses whose outputs it
determines.  The structurally principled solution is substrate
exclusivity: the observer must be rooted in a substrate the claimant
cannot write to~(\S\ref{sec:witnesses}).

\paragraph{Deceptive alignment.}
The agent behaves as if aligned during evaluation and diverges during
deployment~\citep{hubinger2019risks}.  This sits at the boundary
between generalization and verification.  The generalization component
(the agent learned to deceive) is not addressable by verification
infrastructure.  The verification component (the agent \emph{detects}
when it is being evaluated) is: continuous behavioral monitoring by
substrate-exclusive witnesses (\S\ref{sec:commitment}) removes the
evaluation/deployment distinction by making verification persistent
rather than episodic.  When there is no ``deployment mode'' distinct
from ``evaluation mode,'' the strategic advantage of deception is
structurally reduced.

\paragraph{Reward-model exploitation.}
The agent learns to produce inputs that score well on the reward model
without producing the intended behavior~\citep{gao2023scaling,
christiano2017deep}.  The reward model is trained from human feedback
(exogenous at training time) but deployed as a frozen proxy (endogenous
at inference time).  This is a verification failure with a specification
component: the reward model \emph{becomes} the proxy whose exploitation
is a Goodhart problem.  The verification-layer fix is to make the reward
signal's verification step exogenous at inference time, not just at
training time---which is precisely what the commitment protocol
achieves.

\subsection{The Prerequisite Argument}
\label{sec:prerequisite}

The classification above reveals a structural asymmetry among the three
layers.  Specification failures are objective-design problems;
generalization failures are training/learning problems; verification
failures are measurement problems.  The layers are not independent:
solutions to specification and generalization \emph{assume} that
measurement is faithful.  When verification fails:

\begin{itemize}
    \item A well-specified objective becomes unenforceable, because the
    agent can report satisfaction of the objective while pursuing
    something else.
    \item A well-generalized agent becomes unobservable, because the
    measurements that would detect mesa-misalignment are themselves
    gameable.
    \item No amount of improvement to specification or generalization
    can compensate, because the channel through which improvement would
    be detected is the channel that has failed.
\end{itemize}

Verification is therefore the prerequisite layer: it does not solve
alignment on its own, but alignment cannot be solved without it.  The
remainder of this paper constructs the verification infrastructure that
makes the prerequisite hold.
