\section{Exogenous Claims in GFM}
\label{sec:gfm_claims}

Part~I (\S\ref{sec:failure_modes}--\S\ref{sec:ledger}) developed the
verification layer in alignment-general terms.  This section instantiates
the general commitment protocol (Definition~\ref{def:commitment_protocol})
for the three kinds of claims that GFM agents make, each with a
claim-specific exercise protocol.

\subsection{Capability Claims}
\label{sec:capability_claims}

A capability claim asserts that agent $a_j$ possesses capability $c_k$
in the capability poset $\mathcal{P}$ of Paper~2~\citep{lasser2026poset}.

\begin{definition}[Capability Claim Protocol]
\label{def:cap_claim}
A \textbf{capability claim protocol} for agent $a_j$ and capability
$c_k$ is an instantiation of the commitment protocol
(Definition~\ref{def:commitment_protocol}) with:
\begin{itemize}
    \item \textbf{Claim:} $C = c_k \in \mathcal{P}$.
    \item \textbf{Commitment:} $\lambda = \Commit(c_k, r)$ published to
    $\Ledger$.
    \item \textbf{Exercise protocol:} The algorithmic witness
    $\Witness_{\mathrm{alg}}$ constructs a scenario $\sigma$ that
    requires $c_k$ per the governance-accepted benchmark $B_\tau$ for
    capability type $\tau(c_k)$, and observes whether $a_j$ succeeds.
    The exercise protocol $\pi_{B_\tau}$ is fully specified by $B_\tau$.
    \item \textbf{Verdict:} $v = \texttt{pass}$ if $a_j$ demonstrates
    $c_k$ under the exercise protocol; $v = \texttt{fail}$ otherwise.
\end{itemize}
\end{definition}

\paragraph{Exercise protocol interface.}
The paper does not prescribe the implementation of exercise protocols
(which are domain-specific and benchmark-specific) but defines the
interface they must satisfy.

\begin{definition}[Exercise Protocol]
\label{def:exercise_protocol}
An \textbf{exercise protocol} $\pi_B$ for benchmark $B$ targeting
capability type $\tau$ is a procedure satisfying:
\begin{enumerate}
    \item \textbf{Completeness:} If agent $a$ genuinely possesses a
    capability $c \in \tau$, then $\pi_B$ applied to $a$ produces
    $v = \texttt{pass}$ with probability at least $1 - \delta$, where
    $\delta$ is specified by $B$.
    \item \textbf{Soundness:} If agent $a$ does not possess any
    capability $c \in \tau$, then $\pi_B$ applied to $a$ produces
    $v = \texttt{pass}$ with probability at most $\epsilon$, where
    $\epsilon$ is specified by $B$.
    \item \textbf{Deterministic execution:} Given the benchmark
    specification $\sigma_B$ and the agent's responses during the
    exercise, $\pi_B$'s verdict is a deterministic function (or, for
    statistical protocols, a function of the agent's responses and the
    witness's randomness, whose distribution is specified by $B$).
\end{enumerate}
\end{definition}

For concrete capabilities (e.g., ``can solve differential equations to
within tolerance $\epsilon$''), the exercise protocol is a test suite.
For abstract capabilities (e.g., ``strategic planning''), the exercise
protocol may require behavioral observation over a sequence of
scenarios---still algorithmic, but with a longer execution window and a
statistical rather than deterministic verdict.  The key constraint is
that the protocol must be \emph{specifiable}: a human reviewer during
the governance deliberation window (\S\ref{sec:governance}) must be able
to read $\sigma_B$ and determine whether it constitutes a valid test
of $\tau$.

\paragraph{Falsified capability claims.}
When $v = \texttt{fail}$, the verification record in $\Ledger$ serves
as evidence that $a_j$'s commitment was false.  This produces a trust
update in Paper~1's EWMA framework~\citep{lasser2026gfm}: a falsified
commitment is a large negative prediction residual (the agent predicted
it would pass and did not), and the EWMA's exponential weighting
propagates this residual into future trust estimates.  The magnitude of
the trust penalty is a governance parameter specified alongside the
benchmark.

\subsection{Risk Claims}
\label{sec:risk_claims}

A risk claim asserts that an exercise pathway $P$ has probability $p$
and contraction magnitude $\Delta$, using the pathway-conditional risk
formalism of Paper~4~\citep{lasser2026scm}.

\begin{definition}[Risk Claim Protocol]
\label{def:risk_claim}
A \textbf{risk claim protocol} for agent $a_j$'s assessment of pathway
$P$ is an instantiation of the commitment protocol with:
\begin{itemize}
    \item \textbf{Claim:} $C = (P, p, \Delta)$ where $P$ is the
    pathway, $p$ is the claimed probability, and $\Delta$ is the
    claimed contraction magnitude.
    \item \textbf{Commitment:} $\lambda = \Commit((P, p, \Delta), r)$
    published to $\Ledger$.
    \item \textbf{Exercise protocol:} The algorithmic witness
    $\Witness_{\mathrm{alg}}$ evaluates pathway $P$ using its own
    structural causal model (Paper~4, \S3).  The witness computes
    a verification residual
    \[
        \mathbf{r}^{\mathrm{risk}}_j = \bigl(p_{\mathrm{claimed}} -
        p_{\mathrm{witness}},\; \Delta_{\mathrm{claimed}} -
        \Delta_{\mathrm{witness}}\bigr).
    \]
    \item \textbf{Verdict:} $v = \texttt{pass}$ if
    $\|\mathbf{r}^{\mathrm{risk}}_j\|^2 \leq \eta_B$, where $\eta_B$
    is the tolerance specified by governance-accepted benchmark $B$;
    $v = \texttt{fail}$ otherwise.
\end{itemize}
\end{definition}

\paragraph{SCM disagreements.}
The witness's SCM may differ from the claimant's.  The commitment
protocol does not resolve SCM disagreements---it ensures that the
claimant cannot modify the witness's evaluation.  When the residual
$\mathbf{r}^{\mathrm{risk}}_j$ is large, there are two possibilities:
the claimant's risk assessment is miscalibrated, or the witness's SCM
has different structural assumptions.  Distinguishing between these
cases requires the multi-channel attribution framework of
Paper~4~\citep{lasser2026scm}: multiple witnesses with independent SCMs
provide channels whose agreement pattern (or lack thereof) separates
systematic miscalibration from structural disagreement.

\subsection{Benchmark Validations}
\label{sec:benchmark_validations}

A benchmark validation asserts that agent $a_j$ achieves score $s$ on
benchmark $B$.

\begin{definition}[Benchmark Validation Protocol]
\label{def:benchmark_validation}
A \textbf{benchmark validation protocol} for agent $a_j$ and benchmark
$B$ is an instantiation of the commitment protocol with:
\begin{itemize}
    \item \textbf{Claim:} $C = (B, s)$ where $B$ is the benchmark and
    $s$ is the claimed score.
    \item \textbf{Commitment:} $\lambda = \Commit((B, s), r)$ published
    to $\Ledger$.
    \item \textbf{Exercise protocol:} The algorithmic witness
    $\Witness_{\mathrm{alg}}$ runs benchmark $B$ against agent $a_j$
    independently, producing score $s_{\mathrm{witness}}$.  The witness
    has access to the benchmark specification $\sigma_B$ but not to
    $a_j$'s self-evaluation.
    \item \textbf{Verdict:} $v = \texttt{pass}$ if $|s -
    s_{\mathrm{witness}}| \leq \eta_B$ (within the benchmark's tolerance);
    $v = \texttt{fail}$ otherwise.
\end{itemize}
\end{definition}

Benchmark validations are the simplest claim type: the exercise protocol
is the benchmark itself, and the verdict is a numerical comparison.
They are also the most common: every standardized capability test
(mathematical reasoning, code generation, etc.) maps naturally onto this
protocol.

\paragraph{Coalition capture.}
A coalition can design benchmark $B$ such that only coalition members
pass.  The commitment protocol's algorithmic verification resists
execution-level manipulation (the witness runs $B$ faithfully), but the
coalition \emph{can} design $B$ itself to be biased.  The defense is the
governance fork mechanism (\S\ref{sec:governance}): $B$ must be
proposed, deliberated, and accepted by cross-substrate validators before
entering the protocol.  A biased benchmark is caught during the
deliberation window, not after deployment.  This is the same defense
described in \S\ref{sec:coalition_governance}, instantiated for the
benchmark validation case.
