\section{Integration with GFM Machinery}
\label{sec:integration}

This section connects the exogenous verification construction of
\S\ref{sec:gfm_claims} to the existing machinery of
Papers~1--4, showing how committed claims improve the fidelity of each
mechanism.

\subsection{Closing the Observation-Channel Integrity Assumption}
\label{sec:channel_independence}

Paper~4's multi-channel attribution (\S2.4 of~\citet{lasser2026scm})
aggregates evidence from multiple observation channels via log-odds
combination.  The aggregation's validity depends on the channels being
\emph{independently informative}: each channel's evidence is
conditionally independent of the others given the true state.  If an
agent can write to multiple channels simultaneously (because they all
run on the same substrate), the independence assumption fails, and the
log-odds aggregation can be arbitrarily manipulated.

The committed-claim construction closes this gap at two levels of
strength: a \emph{channel isolation} result that holds adversarially
(no sufficiency assumptions on the agent's internal state), and a
\emph{channel independence} result that provides full conditional
independence under additional assumptions.

\begin{proposition}[Channel Isolation under Substrate Exclusivity]
\label{prop:channel_isolation}
Let $\{E_1, \ldots, E_n\}$ be observation channels for agent $a$, and
let each $E_i$ be verified by an algorithmic witness
$\Witness^{(i)}_{\mathrm{alg}}$ on substrate $s_i$ using
governance-accepted benchmark $B_i$.  Assume the witnesses are on
distinct substrates ($s_i \neq s_j$ for $i \neq j$) and each
$s_i \neq s(a)$.
Then no coalition confined to a single substrate can simultaneously
write to the inputs of two or more witnesses.  Formally, for any
coalition $K$ on substrate $s$ and any $i \neq j$:
if $s \neq s_i$ and $s \neq s_j$, then $K$ has no write path to
either $\Witness^{(i)}_{\mathrm{alg}}$ or
$\Witness^{(j)}_{\mathrm{alg}}$; if $s = s_i$ for some $i$, then
$K$ has no write path to any $\Witness^{(j)}_{\mathrm{alg}}$ with
$j \neq i$.
\end{proposition}

\begin{proof}
By Definition~\ref{def:alg_witness}, each witness
$\Witness^{(i)}_{\mathrm{alg}}$ executes on substrate $s_i$.  By
assumption, the substrates are pairwise distinct and disjoint
from $s(a)$.  A coalition $K$ on substrate $s$ can write only to
processes on $s$ (substrate exclusivity).  If $s \neq s_i$, then $K$
cannot write to $\Witness^{(i)}_{\mathrm{alg}}$'s inputs.  Since
the $s_i$ are pairwise distinct, $K$'s substrate $s$ can equal at
most one $s_i$, so $K$ can write to at most one witness's inputs.
In particular, $K$ cannot simultaneously corrupt two or more
channels.
\end{proof}

Channel isolation is the adversarially robust result for
single-substrate coalitions: it holds regardless of the agent's
internal state, strategic behavior, or latent variables not captured
by the capability profile.  The absence of a write path means that
a coalition confined to one substrate cannot simultaneously corrupt
multiple channels, restoring \emph{channel integrity} for Paper~4's
multi-channel attribution~\citep{lasser2026scm}: no single-substrate
actor can inject coordinated false evidence across channels.
Channel integrity is weaker than the full conditional independence
that optimal log-odds aggregation requires; natural correlations
between channels (e.g., from shared agent state) may still exist
under isolation alone.  The stronger conditional-independence
guarantee, which licenses exact log-odds combination, requires the
additional assumptions of
Proposition~\ref{prop:channel_independence} below.

Under additional assumptions about the agent's internal state, the
isolation result strengthens to full conditional independence:

\begin{proposition}[Channel Independence under Committed Verification]
\label{prop:channel_independence}
Under the assumptions of
Proposition~\ref{prop:channel_isolation}, assume additionally:
\begin{enumerate}
    \item[(ii)] \textbf{Independent governance:} The benchmark
    parameters $\{B_1, \ldots, B_n\}$ are independently specified
    by governance.  (This assumption is absorbed by conditioning on
    $\{B_1, \ldots, B_n\}$ in the proposition's conclusion; its role
    is to ensure that the conditioning set is well-defined and that
    each benchmark was produced by an independent governance process
    rather than a correlated batch proposal.)
    \item[(iii)] \textbf{Joint sufficiency:} $\mathbf{C}_a$ is a
    jointly sufficient statistic for agent $a$'s exercise-level
    behavior across all benchmarks, i.e., conditional on
    $\mathbf{C}_a$, the agent's performance on protocol $\pi_{B_i}$
    is independent of its performance on $\pi_{B_j}$ for $i \neq j$.
    This excludes persistent cross-exercise state (e.g., the agent
    learning from one exercise and adapting its strategy for the next)
    and shared agent-side randomness not absorbed by $\mathbf{C}_a$.
    \item[(iv)] \textbf{Independent randomness:} For statistical
    exercise protocols, each witness $\Witness^{(i)}_{\mathrm{alg}}$
    draws its randomness independently of other witnesses' randomness.
\end{enumerate}
Then the verified outputs $\{v_1, \ldots, v_n\}$ are conditionally
independent given $\mathbf{C}_a$ and $\{B_1, \ldots, B_n\}$.
\end{proposition}

\begin{proof}
Each $v_i = \pi_{B_i}(a, \Witness^{(i)}_{\mathrm{alg}})$ is determined
by three inputs: (1)~$a$'s behavior under exercise protocol $\pi_{B_i}$,
(2)~the benchmark specification $B_i$, and (3)~the witness's
execution on substrate $s_i$ (deterministic or with independent
randomness per assumption~(iv);
Definition~\ref{def:alg_witness}).

Conditional on $(\mathbf{C}_a, B_1, \ldots, B_n)$, all shared causes
are blocked:
\begin{itemize}
    \item \textbf{No agent-mediated path:}  By assumption~(iii),
    $\mathbf{C}_a$ is jointly sufficient: conditional on
    $\mathbf{C}_a$, the agent's performance on $\pi_{B_i}$ is
    independent of its performance on $\pi_{B_j}$.  This blocks
    correlations from shared latent state (training history, strategic
    disposition, memory of prior exercises), persistent cross-exercise
    adaptation (the agent learning from one exercise and modifying its
    approach to the next), and shared task content across benchmarks
    (e.g., two benchmarks testing overlapping skills).  The agent's
    performance on each $\pi_{B_i}$ is then characterized by the
    completeness and soundness parameters $(\delta_i, \epsilon_i)$ of
    Definition~\ref{def:exercise_protocol}, which are set independently
    per benchmark by governance (assumption~(ii)).
    \item \textbf{No witness-mediated path:}  By
    Proposition~\ref{prop:channel_isolation}, no single-substrate
    coalition can write to two witnesses simultaneously.  Since each
    witness is on a distinct substrate ($s_i \neq s_j$; inherited from
    Proposition~\ref{prop:channel_isolation}),
    witness $\Witness^{(i)}_{\mathrm{alg}}$ cannot influence
    $\Witness^{(j)}_{\mathrm{alg}}$.
    \item \textbf{No shared-randomness path:}  Per assumption~(iv),
    witnesses draw independent randomness, so stochastic verdicts
    do not correlate through shared seeds.
    \item \textbf{No shared-environment path:}  Benchmark parameters
    $B_i$ are conditioned on, removing correlation through shared
    governance.  Any remaining common causes from the shared physical
    environment are absorbed by joint sufficiency (assumption~(iii)),
    which guarantees that conditioning on $\mathbf{C}_a$ blocks all
    cross-exercise dependencies.
\end{itemize}
Therefore, conditional on $(\mathbf{C}_a, B_1, \ldots, B_n)$, the
verdicts $v_i$ are conditionally statistically independent: each is
determined by the agent's $\mathbf{C}_a$-characterized behavior on its
benchmark, the witness's independent execution on its own substrate,
and independent randomness (assumption~(iv)).
\end{proof}

\begin{remark}
The relationship between the two propositions reflects a general
pattern in adversarial analysis: the weaker result
(Proposition~\ref{prop:channel_isolation}) holds under fewer
assumptions and is therefore robust to adversarial violation of
the stronger assumptions, while the stronger result
(Proposition~\ref{prop:channel_independence}) provides tighter
statistical guarantees when the additional assumptions hold.  For a
deceptive agent with latent strategic state not captured by
$\mathbf{C}_a$ (violating assumption~(iii)), channel isolation
still holds---the agent cannot simultaneously corrupt multiple
channels---even though conditional independence may not.  For a
non-deceptive agent whose capability profile is an accurate
summary, the full conditional-independence guarantee supports
optimal log-odds aggregation.
\end{remark}

\begin{remark}
Both propositions require $n$ distinct witness substrates for $n$
channels.  When the substrate partition has $m$ substrates total, at
most $m - 1$ independent channels can be constructed (one substrate is
occupied by $a$, and each remaining substrate hosts at most one
independent witness).  This bounds the effective multiplicity of
independent verification channels by $m - 1$, connecting verification
depth to substrate diversity.
\end{remark}

\subsection{Committed Baselines for SPRT Detection}
\label{sec:committed_baselines}

The SPRT-based scorpion detection of~\citet{lasser2026scm} compares an agent's behavioral
observations against a baseline.  In the current framework, the baseline
is the \emph{actor's own estimate} of the agent's expected behavior, an
endogenous quantity that is subject to the actor's model uncertainty and
potential manipulation.

Committed capability claims (\S\ref{sec:capability_claims}) provide an
exogenous alternative.  When agent $a_j$ has committed to capability
profile $\mathbf{C}_a$ and the commitment has been verified by a
substrate-exclusive witness, the behavioral distribution $p(\cdot \mid
\mathbf{C}_a)$ consistent with the committed profile becomes the
baseline.  The SPRT then tests whether $a_j$'s observed behavior is
consistent with its own committed claims, rather than with the actor's
model of what $a_j$ should be doing.

\begin{proposition}[Detection Improvement under Committed Baselines]
\label{prop:committed_detection}
Let $p_0$ be the behavioral distribution under the committed capability
profile $\mathbf{C}_a$, and let $\hat{p}_0$ be the actor's estimated
baseline.  If the agent's true distribution is $p_1 \neq p_0$, then
the SPRT detection rate satisfies:
\[
    \frac{1}{\mathbb{E}_{p_1}[T_{\mathrm{detect}}^{\mathrm{committed}}]}
    \;\geq\;
    \frac{1}{\mathbb{E}_{p_1}[T_{\mathrm{detect}}^{\mathrm{estimated}}]}
\]
whenever $D_{\mathrm{KL}}(p_1 \| p_0) \geq D_{\mathrm{KL}}(p_1 \|
\hat{p}_0)$, i.e., when the committed baseline is at least as
discriminating as the estimated baseline.  Equality holds if and only if
$\hat{p}_0 = p_0$.
\end{proposition}

\begin{proof}
By Proposition~\ref{prop:sprt_detection}, the expected detection time
is inversely proportional to the KL divergence between the true and
baseline distributions.  If $D_{\mathrm{KL}}(p_1 \| p_0) \geq
D_{\mathrm{KL}}(p_1 \| \hat{p}_0)$, then $\mathbb{E}[T^{\mathrm{committed}}]
\leq \mathbb{E}[T^{\mathrm{estimated}}]$, and detection is faster
(or equally fast) under the committed baseline.
\end{proof}

The committed baseline is at least as discriminating as the estimated
baseline whenever the agent's commitment is \emph{informative}: that
is, when the committed profile $\mathbf{C}_a$ constrains the expected
behavioral distribution more tightly than the actor's prior estimate
does.  Since agents commit to specific capability claims (not vague
ranges), committed baselines are generically more informative than
actor-estimated baselines, especially during the initial-period
noise that the trust EWMA~\citep{lasser2026gfm} is most vulnerable to.

\subsection{Trust Dynamics with Verified Claims}
\label{sec:trust_dynamics}

The trust EWMA $T_j$ of~\citet{lasser2026gfm} updates from prediction
residuals.  Committed and verified claims anchor
the prediction baseline and affect trust dynamics in three ways.

\paragraph{Reduced initial-period noise.}
A new agent entering a population with no trust history begins with
$T_j$ at a prior value, and the EWMA requires multiple observations to
converge.  Verified capability claims provide a shortcut: a commitment
that is verified by a substrate-exclusive witness establishes a baseline
that the actor's trust model can condition on, reducing the number of
observations needed for convergence.  The verified claim does not
replace the behavioral evidence; it provides a structural prior that
the EWMA refines.

\paragraph{Falsification as accelerated detection.}
When a commitment is falsified ($v = \texttt{fail}$), the prediction
residual is large by construction: the agent predicted it would
demonstrate capability $c_k$ and did not.  In the EWMA framework, this
produces a correspondingly large negative trust update.  The key
structural property is that falsification is \emph{binary} (pass/fail
against an algorithmic protocol), not graded (how consistent is
behavior?), so the signal is clean and unambiguous.  A single falsified
commitment can move $T_j$ more than many observations of mildly
inconsistent behavior.

\paragraph{Exogenous risk residuals.}
The risk trust $\Trisk_j$ of~\citet{lasser2026scm} is computed from the residual between
claimed and realized risk.  Currently, both the claim and the
realization are endogenous to the actor's model.  Committed risk claims
(\S\ref{sec:risk_claims}), verified by cross-substrate witnesses,
replace the endogenous claim with an exogenous one: the residual
$\mathbf{r}^{\mathrm{risk}}_j$ is now between what the agent committed
to and what the witness measured, not between what the actor estimated
and what the actor observed.  This sharpens $\Trisk_j$ by removing one
of the two endogenous quantities from the residual computation.

\subsection{Verification Impossibility under Monopolarity}
\label{sec:anti_monopolar}

Section~\ref{sec:witness_m2} observed that substrate-exclusive witnesses
require $m \geq 2$.  The following theorem establishes the stronger
claim: in a monopolar population, not only do substrate-exclusive
witnesses fail to exist, but \emph{no rearrangement of the population's
computational resources can produce exogenous verification}.  The
impossibility is grounded in a physical premise about same-substrate
causal access, not in our definitions.

\begin{definition}[Same-Substrate Addressability]
\label{def:same_substrate}
Two processes $p, q$ executing on the same substrate $s$ satisfy
\textbf{same-substrate addressability}: there exist actions available
to $p$ that modify at least one input to $q$, and vice versa.
Formally, writing $\mathrm{Addr}(p, q)$ for the set of $q$'s inputs
addressable by $p$ (analogous to $\mathcal{R}(a, E)$ from
Definition~\ref{def:ver_asymmetry} but applied to arbitrary process
pairs), we require $\mathrm{Addr}(p, q) \neq \emptyset$ and
$\mathrm{Addr}(q, p) \neq \emptyset$.
\end{definition}

\begin{remark}
Same-substrate addressability is a physical premise, not a logical
necessity.  It asserts that processes sharing a computational medium
can causally influence each other's inputs.  This holds for software
processes on shared hardware (they share memory, network, and OS
resources), for biological agents in shared environments (they share
sensory channels and physical space), and more generally for any
processes whose computation is mediated by the same physical
substrate.  It would fail if two processes occupied the same substrate
but were perfectly isolated (e.g., by an impenetrable hardware
partition).  The premise is that such perfect intra-substrate isolation
is not the default condition for co-substrate processes.
\end{remark}

\begin{theorem}[Verification Impossibility under Monopolarity]
\label{thm:anti_monopolar}
Let $\mathcal{A}$ be a population of agents with substrate partition
$\{S_1, \ldots, S_m\}$, and assume same-substrate addressability
(Definition~\ref{def:same_substrate}).  Assume further that the
population's \emph{available actions} are computations on substrates
it currently occupies: the action space does not include provisioning
new physical infrastructure, recruiting biological observers, or
otherwise instantiating computation on a substrate not in
$\{S_1, \ldots, S_m\}$.  If $m = 1$, then:
\begin{enumerate}
    \item[(a)] For every evaluation function $E$ constructible from the
    population's computational resources, and every agent
    $a_j \in \mathcal{A}$, the evaluation is endogenous:
    $\mathcal{R}(a_j, E) \neq \emptyset$.
    \item[(b)] No computational action available to the population can
    produce an evaluation function $E'$ with
    $\mathcal{R}(a_j, E') = \emptyset$ for any $a_j$.
\end{enumerate}
\end{theorem}

\begin{remark}
Under the positive-measure reachability assumption (that
same-substrate addressability produces a reachable set of positive
measure, not merely a nonempty one), part~(a) strengthens to
$\rho(a_j, E) > 0$.  This holds for all practical substrates:
if a process can modify at least one evaluator input, it can
modify that input over a range of values, not just at a single
point.  When input spaces are discrete (the default for
computational processes), $\mathcal{R} \neq \emptyset$ and
$\rho > 0$ coincide.
\end{remark}

\begin{proof}
(a)~If $m = 1$, every process in the population (including any
evaluator $E$ the population constructs) executes on the single
substrate $S_1$.  By same-substrate addressability, agent $a_j$ on
$S_1$ can address at least one input to $E$ on $S_1$:
$\mathcal{R}(a_j, E) \neq \emptyset$.  The evaluation is therefore
endogenous (Definition~\ref{def:endo_exo}).

(b)~Achieving $\mathcal{R}(a_j, E') = \emptyset$ requires $E'$ to have
no inputs addressable from $a_j$'s substrate.  By same-substrate
addressability, any $E'$ on $S_1$ has addressable inputs from $a_j$.
Therefore $E'$ must execute on some $s' \neq S_1$.  By the
computational action-model assumption, the population's available
actions are computations on $S_1$.  No sequence of computations on
$S_1$ produces a process on $s' \neq S_1$: substrate provisioning is a
physical action (procuring hardware, establishing a biological observer,
or otherwise instantiating computation on a distinct physical medium)
outside the computational action space.
\end{proof}

\begin{corollary}[Monopolar Epistemic Closure]
\label{cor:monopolar_undetect}
In a monopolar population ($m = 1$) satisfying same-substrate
addressability, no agent can distinguish between:
\begin{enumerate}
    \item[(a)] Verification passing because the claim is true, and
    \item[(b)] Verification passing because the evaluator's inputs have
    been influenced by the claimant ($\mathcal{R}(a, E) \neq \emptyset$).
\end{enumerate}
\end{corollary}

\begin{proof}
Distinguishing (a) from (b) requires a reference evaluation with
$\mathcal{R}(a, E) = \emptyset$.  By Theorem~\ref{thm:anti_monopolar}(b),
no such evaluation is constructible from the population's computational
resources.  Therefore the distinction is unresolvable
from within the population.
\end{proof}

Theorem~\ref{thm:anti_monopolar} and Corollary~\ref{cor:monopolar_undetect}
give the anti-monopolar property a second structural justification beyond
Paper~3's $\volL$ argument~\citep{lasser2026horizon}:

\begin{itemize}
    \item \textbf{\citet{lasser2026horizon}:} A monopolar population is anti-maximizing
    ($\volL$ decreases because cross-substrate cooperative capacity is
    destroyed).
    \item \textbf{This paper:} A monopolar population is
    \emph{unverifiable} (no substrate-exclusive witnesses exist, so all
    capability claims, risk assessments, and behavioral consistency
    checks are endogenous).
\end{itemize}

A population that becomes monopolar loses both the safety guarantee
\emph{and the ability to detect that it has lost it}.  This makes
monopolar convergence doubly dangerous: the loss of cross-substrate
cooperative capacity~\citep{lasser2026horizon} is compounded by the loss of the
verification infrastructure that would detect the convergence.  The
compound failure is silent---from within a monopolar population, all
verification reports ``everything is fine,'' because the evaluators
share the substrate that the evaluated agents can influence.

This result strengthens the case for substrate diversity as a
structural safety property.  \citet{lasser2026horizon} argued that
diversity is instrumentally valuable (it maximizes $\volL$); we argue
that
diversity is epistemically necessary (it is the precondition for knowing
whether the safety guarantee holds).  A safety guarantee that cannot be
verified is not a guarantee---it is a hope.
