\section{Introduction}
\label{sec:introduction}

Alignment research addresses three structural layers, each with distinct
failure modes and distinct research traditions.  \emph{Specification}
asks what objective an agent should optimize; failures here include
Goodhart's Law, perverse instantiation, and instrumental
convergence~\citep{bostrom2014superintelligence, omohundro2008basic}.
\emph{Generalization} asks whether the agent reliably pursues the
specified objective; failures here include mesa-optimization and goal
misgeneralization~\citep{hubinger2019risks}.  \emph{Verification} asks
whether the agent's pursuit of the objective is measured faithfully;
failures here include wireheading, self-assessment gaming, deceptive
alignment, and reward-model
exploitation~\citep{amodei2016concrete, gao2023scaling}.  The
overwhelming majority of the field's attention has been directed at the
first two layers.  This paper argues that the third layer---verification---is
the prerequisite for the other two to matter.

The argument is structural.  A perfectly specified, perfectly generalized
agent that can game its evaluator is still dangerous: it can construct
self-consistent measurements reporting alignment while its actual behavior
diverges.  The verification problem has a precise form that we formalize
in \S\ref{sec:asymmetry}: the agent's action space includes actions that
modify the evaluator's inputs, and this \emph{write-access asymmetry}
grows with capability.  As the agent becomes more capable, more of the
evaluator's input space becomes reachable from the agent's actions, until
at the limit the evaluator is entirely endogenous: a function of
quantities the agent controls.  At that point, verification is not
merely difficult; it is structurally absent.

The fix is equally structural.  The only invariant limit on an agent's
write access is \emph{substrate exclusivity}: an agent cannot write to
observation channels rooted in substrates it does not
occupy~\citep{lasser2026horizon, lasser2026scm}.  This is not a
capability limitation (the agent may be arbitrarily powerful within its
own substrate) but a physical constraint on cross-substrate causal
influence.  We construct a cryptographic commitment protocol
(\S\ref{sec:commitment}) in which the agent commits to a claim before
verification occurs, and a \emph{substrate-exclusive witness}
(\S\ref{sec:witnesses}) verifies the claim through a fully algorithmic
exercise protocol.  The commitment is binding (the agent cannot change
the claim after committing) and hiding (the witness cannot learn the
claim before verification runs).  Human judgment enters at exactly one
layer: benchmark admissibility, decided through a governance fork
mechanism analogous to Ethereum Improvement Proposals.  Per-claim
verification is fully algorithmic against governance-accepted benchmarks
and operates at machine speed.

The paper is organized in two parts.  Part~I (\S\ref{sec:failure_modes}--\S\ref{sec:ledger}) develops the verification layer in
alignment-general terms: no GFM vocabulary is required, and the
construction applies to any framework whose evaluation step is
vulnerable to the write-access asymmetry.  Part~II
(\S\ref{sec:gfm_claims}--\S\ref{sec:integration}) instantiates the
construction for the goal-frontier maximization (GFM) framework of
Papers~1--4~\citep{lasser2026gfm, lasser2026poset, lasser2026horizon,
lasser2026scm}, closing the observation-channel integrity assumption
that multi-channel attribution~\citep{lasser2026scm} relies on and
connecting verification integrity to the anti-monopolar
property~\citep{lasser2026horizon}.

\paragraph{Epistemic status.}
This paper provides verification infrastructure.  The structural
guarantees of Papers~1--3 (the anti-monopolar property, the
substrate-diversity gradient, the cooperative production theorem) do not
depend on this paper's correctness; a flaw here degrades verification
fidelity, not alignment properties.  The relationship is analogous to
\citet{lasser2026scm}: implementation-layer machinery that makes the framework's
theoretical guarantees \emph{enforceable} rather than merely
\emph{provable}.  The cryptographic constructions we use (Pedersen commitments; with
zero-knowledge proofs as an optional variant for sensitive claims) are
standard and well-studied; our contribution is the structural argument
for why they are the right tool and the protocol that connects them to
the alignment verification problem.
