\section{Ledger Architecture}
\label{sec:ledger}

The commitment protocol of \S\ref{sec:commitment} requires a persistent,
tamper-evident record of commitments, verification results, and
governance decisions.  This section specifies the minimal ledger
architecture that satisfies those requirements.

\subsection{Design Principles}
\label{sec:ledger_principles}

The ledger is motivated by verification requirements, not by analogy to
financial blockchains.  Its design is driven by three properties that
the commitment protocol needs and constrained by what the protocol does
\emph{not} need.

\paragraph{Required properties.}
\begin{enumerate}
    \item \textbf{Append-only:} Entries cannot be deleted or modified
    after recording.  This ensures that commitments, once published,
    remain binding---an agent cannot retroactively alter a commitment
    to match a different claim.
    \item \textbf{Publicly auditable:} Any ledger participant can verify
    the integrity of the entire chain from genesis to the present.
    This enables third-party audit of both individual verification
    records and aggregate patterns (e.g., an agent whose claims are
    systematically falsified across evaluation protocols).
    \item \textbf{Substrate-distributed:} The ledger is replicated
    across nodes on different substrates.  No single substrate controls
    the canonical copy.  This prevents a same-substrate coalition from
    modifying or censoring verification records.
\end{enumerate}

\paragraph{Not required.}
The verification ledger does \emph{not} require token economics,
proof-of-work or proof-of-stake consensus, smart contract execution, or
high transaction throughput.  Governance decisions are infrequent
(monthly or less); per-claim verification entries are more frequent but
do not require global consensus (see \S\ref{sec:ledger_consensus}).
The minimal construction is an authenticated append-only data
structure---a Merkle hash tree~\citep{crosby2009efficient}---replicated
across substrate-exclusive nodes.

\subsection{Verification Entries}
\label{sec:ver_entries}

\begin{definition}[Verification Ledger Entry]
\label{def:ver_entry}
A \textbf{verification entry} in $\Ledger$ is a tuple
\[
    e_{\mathrm{ver}} = (a_j,\; \lambda,\; w,\; B,\; v,\; \xi,\;
    t,\; \sigma_a,\; \sigma_w)
\]
where:
\begin{itemize}
    \item $a_j$ is the agent identifier,
    \item $\lambda = \Commit(C, r)$ is the commitment,
    \item $w$ identifies the algorithmic witness
    $\Witness_{\mathrm{alg}}$,
    \item $B$ is the governance-accepted evaluation protocol identifier,
    \item $v \in \{\texttt{pass}, \texttt{fail}\}$ is the verdict,
    \item $\xi$ is a hash of the supporting evidence (raw data is
    stored off-chain; the hash ensures integrity),
    \item $t$ is the timestamp,
    \item $\sigma_a$ is the agent's signature over $\lambda$ (from the
    Commit phase), and
    \item $\sigma_w$ is the witness's signature over
    $(a_j, \lambda, B, v, \xi, t)$
    (from the Verify phase), binding the verdict to the specific
    agent, commitment, and benchmark to prevent replay or rebinding
    attacks.
\end{itemize}
The entry is completed when either: (a)~$a_j$ opens the commitment by
publishing $(C, r)$, at which point any participant can verify
$\Commit(C, r) = \lambda$; or (b)~the reveal deadline
$t + \Delta_{\mathrm{reveal}}$ expires without a reveal, in which
case a timeout entry is appended with $v = \texttt{fail}$ and
$\xi = \texttt{reveal\_timeout}$ (no witness signature is required
for timeout entries, as they are mechanically verifiable from the
ledger timestamps).
\end{definition}

Verification entries require agreement between exactly two parties: the
committing agent (who signs the commitment) and the verifying witness
(who signs the verdict).  This two-party, cross-substrate agreement is
sufficient for per-claim records and is simpler than general
multi-party consensus.

\subsection{Governance Entries}
\label{sec:gov_entries}

\begin{definition}[Governance Ledger Entry]
\label{def:gov_entry}
A \textbf{governance entry} in $\Ledger$ is a tuple
\[
    e_{\mathrm{gov}} = (B,\; \sigma_B,\; \tau,\; p,\;
    \mathcal{D},\; \mathcal{V},\; \mathrm{outcome},\; t_{\mathrm{eff}})
\]
where:
\begin{itemize}
    \item $B$ is the evaluation protocol (benchmark) identifier,
    \item $\sigma_B$ is the protocol specification (what it tests,
    scoring criteria, passing threshold),
    \item $\tau$ is the target capability type,
    \item $p$ is the proposer identifier,
    \item $\mathcal{D}$ is the deliberation record (all analysis
    submitted during the deliberation window, with submitter
    identifiers and timestamps),
    \item $\mathcal{V}$ is the vote record (per-validator votes with
    substrate identifiers),
    \item $\mathrm{outcome} \in \{\texttt{accepted},
    \texttt{rejected}\}$, and
    \item $t_{\mathrm{eff}}$ is the effective date (for accepted
    protocols; null for rejected ones).
\end{itemize}
\end{definition}

The governance record is permanent: rejected proposals remain in the
ledger with their full deliberation records.  This provides an audit
trail for evaluation-protocol evolution and prevents re-proposal of
previously-studied-and-rejected protocols without presenting new
evidence.  The deliberation record is itself a public good: future
protocol designers can study past proposals' failure modes.

\subsection{Consensus Among Substrate-Exclusive Nodes}
\label{sec:ledger_consensus}

The ledger's consensus requirements differ between entry types,
reflecting their different trust models and frequency.

\paragraph{Verification entries.}
Appending a verification entry requires agreement between the committing
agent $a_j$ and the verifying witness $w$---a two-party cross-substrate
protocol.  No global consensus is needed for \emph{individual entry
validity}: the entry is self-verifying, since any participant can check
the signatures (including the witness's signature over the full context
$(a_j, \lambda, B, v, \xi, t)$) and, after the commitment is opened,
verify that $\Commit(C, r) = \lambda$.  Conflicts (e.g., an
agent claiming the witness misreported the verdict) are resolvable by
replaying the exercise protocol against the committed data, which is
deterministic by Definition~\ref{def:alg_witness}.

\emph{Chain consistency} across replicated nodes is maintained by the
Merkle hash tree structure: each node can verify that its local chain
is internally consistent (no entries have been modified or reordered)
and can compare root hashes with nodes on other substrates to detect
equivocation (a node presenting different chains to different
observers).  Because the chain is append-only and entries are
self-verifying, equivocation is detectable but not preventable without
global consensus---a deliberately lightweight design, since the
verification entries' integrity does not depend on ordering (each
entry is independently checkable), only on non-suppression.

\paragraph{Governance entries.}
Appending a governance entry (protocol acceptance) requires a
cross-substrate supermajority as specified in
Definition~\ref{def:gov_fork}: at least fraction $\theta$ of validators
overall, with at least one validator per represented substrate.  This is
a special-purpose consensus that is simpler than general Byzantine fault
tolerance because: (a)~governance entries are infrequent (monthly or
less), so latency tolerance is high; (b)~the vote is over a
well-defined proposal with a fixed deliberation record, not an
arbitrary state transition; and (c)~the cross-substrate requirement is a
structural constraint (ensured by the substrate partition) rather than a
probabilistic one (ensured by random committee selection).

\subsection{Relation to Application-Layer Ledgers}
\label{sec:harness_relation}

Any alignment framework that maintains an internal record of agent
properties (capabilities, trust scores, behavioral histories) can
layer the verification ledger beneath it.  The application-layer
ledger records \emph{what} is tracked; the verification ledger of
this section provides the \emph{integrity layer} that makes tracking
trustworthy.  An application-layer entry claiming ``agent $a_j$ has
property $P$ with confidence $c$'' is grounded when the verification
ledger contains a corresponding entry showing that $P$ was committed,
exercised by a substrate-exclusive witness, and confirmed.
Part~II (\S\ref{sec:integration}) instantiates this layering for
the GFM framework's capability and trust ledger.
