\section*{Author Contributions}

\textbf{Teague Lasser} identified the statistical-to-causal upgrade as
the unifying contribution, proposed the log-odds risk aggregation and
the two-gate attention/verification architecture, directed all
revisions, and made final editorial decisions. Responsible for the
paper's intellectual direction and all claims made.

\textbf{Claude Opus 4.6} (Anthropic) drafted the formal exposition:
the structural causal model for capability dynamics, the do-calculus
contraction attribution, the $L^2$ risk-trust EWMA dynamics, the
pathway-conditional structural verification residual, the SPRT-based
causal scorpion detection and evasion bandwidth bounds, and the
probabilistic dependency-risk model with the adversarial balance
condition.

\textbf{GPT 5.4} (OpenAI) served as technical reviewer, identifying
formal gaps in the risk-trust convergence claims, the causal
attribution assumptions, the log-odds aggregation's cross-paper type
consistency, and the dependency-risk quantitative analysis --- in
particular the need to replace the pointwise cascade-fraction
assumption with the adversarial balance condition that admits
contagion dynamics.

\medskip
\noindent\emph{Transparency note.} Both AI systems operated as tools under
human direction. Neither system has continuity across sessions, cannot take
responsibility for the work in the sense required by most venue authorship
policies, and cannot respond to reviewer queries independently. They are
listed as authors to accurately represent their contributions to the
intellectual content of the paper, not to claim that they meet all criteria
of traditional academic authorship. The corresponding author for all
inquiries is Teague Lasser.
