\section{Discussion}
\label{sec:discussion}

\subsection{Limitations}

\paragraph{SCM structure learning as channel optimization, not
prerequisite.} The causal attribution channel of
Section~\ref{sec:causal} requires the actor to know or learn the
causal DAG over agent actions and capability changes. Standard
observational structure learning \citep{spirtes2000causation} is
inadequate for the GFM setting for three distinct reasons. First,
the data-generating process is adversarial: scorpions can
strategically align their actions with observed noise to mask their
causal role, so classical conditional-independence tests can be
gamed by an adversary who knows which tests will be run. Standard
structure learning does not have a threat model. Second, the agent
cardinality is non-fixed: new agents can appear, existing agents can
delegate, and a group can coordinate and behave as a single
effective actor, so the variable set itself evolves. Latent-variable
discovery methods (FCI, RFCI) assume a fixed variable cardinality
and do not naturally handle this. Third, the actor cannot typically
intervene on other agents: interventional causal discovery relies on
actual interventions, which a GFM actor can simulate inside its SCM
but cannot enforce in the world. This rules out the entire class of
interventional discovery methods and leaves only observational
methods, which are strictly weaker under the adversarial conditions
above.

Because causal attribution is one channel among several
(Section~\ref{sec:multi_channel}), and capability gating via the
poset of \citet{lasser2026poset} and \citet{lasser2026horizon} is
the strongest and most difficult to fake channel (relative to a
trusted, sufficiently complete capability graph), structure learning
becomes an \emph{optimization target} for one channel's contribution
rather than a \emph{prerequisite} for attribution at all. When the
SCM is poorly learned, the causal channel degrades, the other
channels are not directly degraded by the SCM misspecification, and
the multi-channel aggregate reweights toward the
structurally-grounded channels. The system does not collapse; it
falls back on the trust and capability machinery of
\citet{lasser2026gfm}, \citet{lasser2026poset}, and
\citet{lasser2026horizon}.

The structural assumptions underlying
Proposition~\ref{prop:causal_dominates}(c) are partially testable
through a residual-covariance diagnostic: if
$\mathrm{Cov}(\CausalAttr(a_j), \CausalAttr(a_k) \mid G_t)$ exceeds
the level the current SCM predicts, the SCM is misspecified ---
possible causes include a missing edge, latent common causes,
exogenous-orthogonality failure, non-additivity, or nonstationarity
--- and the causal channel's reliability should degrade accordingly. Integrating such a
diagnostic into a calibrated SCM-confidence scalar
$c_{\SCM}(t) \in [0,1]$ that weights the causal channel's
contribution to the multi-channel aggregate --- so declining
sufficiency down-weights causal attribution without burning specific
agents' trust --- is an architectural direction we identify but do
not formalize here. The full adversarial-robust structure-learning
problem, including the calibration procedure and its interaction
with the behavioral and capability channels, is deferred to a
companion paper.

\paragraph{Coordinated multi-agent attacks and non-additive
interactions.} The one-at-a-time causal attribution
$\CausalAttr(a_j, t)$ (Definition~\ref{def:causal_attribution})
decomposes the total $\volL$-change exactly only under additive SCM
structure (Equation~\eqref{eq:additive_scm}). In multi-agent
populations, coordinated attacks produce non-additive interactions:
an attack requiring $m$ agents to cooperate has
$\Delta \volL$ that depends on their joint action, and the
one-at-a-time intervention $\mathrm{do}(\pi_j = \mathrm{skip})$
produces non-unique individual attributions: under pure
complementarities ($\Delta \volL = \pi_1 \cdot \pi_2$), marginal
attribution may over-count or double-count each coordinator's
contribution; under other non-additive structures the attribution
allocates the interaction term arbitrarily across agents. The
fundamental issue is that the one-at-a-time decomposition is not
well-defined when the structural equation is non-additive, and no
single-agent marginal faithfully represents the joint effect. The exact fix is
Shapley-style attribution
\citep{shapley1953value} that averages over all orderings of agent
inclusion; developing this within the GFM SCM, including its
computational cost (exponential in coalition size without
structure-exploiting approximations) and its interaction with the
SPRT detection rule, is deferred to a companion paper on joint
attribution. In the interim, the multi-channel attribution
architecture (Section~\ref{sec:multi_channel}) provides the
structural defense: coordinated attacks that evade the causal
channel are still detectable through capability gating --- which
constrains the set of agents that \emph{could} have caused the
contraction regardless of interaction structure --- and through
behavioral prediction residuals, which detect individual
coordinators whose actions diverge from their predictive models.
The causal channel's failure under non-additivity degrades one
channel's contribution; the multi-channel aggregate reweights
toward the structurally-grounded channels, as designed.

\paragraph{Risk-trust residual dependence on the actor's model.} The
pathway-conditional structural verification residual
(Definition~\ref{def:risk_residual}) measures divergence between the
claim and the actor's own assessment of the claimed pathway. If the
actor's world model is itself poorly calibrated, the residual reflects
model error rather than claim quality. An agent whose risk assessments
are accurate but disagree with a miscalibrated actor will accumulate
high $\|r^{\mathrm{risk}}\|^2$ and concentrate around low
$\Trisk_j^*$ --- a false negative. This is mitigated by the self-trust
mechanism of \citet{lasser2026gfm}: an actor with a poorly calibrated
model has low $T_s$ and high learning rate, which accelerates model
correction. But during the correction period, risk-trust assessments
may be unreliable.

\paragraph{Exogenous orthogonality as the most fragile assumption.}
Of the three structural assumptions underlying
Proposition~\ref{prop:causal_dominates}(c), exogenous orthogonality
is the most likely to fail in deployment. Shared environmental
shocks --- power grid state, information broadcasts, market
movements, weather, time-of-day effects on a shared population ---
induce covariance between agents' contributions even under a
correctly specified causal DAG. Conditioning on $G_t$ does not
remove time-localized shocks that affect multiple agents' actions
simultaneously. When the assumption fails, the variance reduction
$\sigma_{\mathrm{do}}^2 \leq \sigma^2$ no longer holds and the
causal detector may perform worse than the correlational baseline
(as acknowledged in Proposition~\ref{prop:causal_dominates}(c)).
The residual-covariance diagnostic described above provides
detection: empirically elevated
$\mathrm{Cov}(\CausalAttr(a_j), \CausalAttr(a_k) \mid G_t)$
signals orthogonality failure, and the SCM-confidence scalar
$c_{\SCM}(t)$ should down-weight the causal channel accordingly.
The multi-channel aggregate then shifts weight toward capability
gating and behavioral channels, which do not depend on exogenous
orthogonality. Importantly, orthogonality failure is \emph{detectable} rather
than silent. The residual covariance term
$\mathrm{Cov}(\CausalAttr(a_j), \CausalAttr(a_k) \mid G_t)$ is an
estimable quantity: it can be estimated from repeated observations
at similar capability-graph states, using the same attribution data
the actor already collects for the SPRT detector. When the estimated covariance exceeds the level predicted
under exogenous orthogonality, the assumption has failed and the
failure is empirically visible. This turns the most fragile of the
three structural assumptions into the most monitorable one---a
better epistemic position than ``fragile but unobservable.'' The
natural integration point is the SCM-confidence scalar
$c_{\SCM}(t)$ described above: elevated residual covariance is a
direct input to $c_{\SCM}$ updating, triggering the multi-channel
aggregate to down-weight the causal channel before the degraded
variance reduction produces misleading detection results.
Characterizing the populations and environments in
which exogenous orthogonality approximately holds --- and deriving
a partial-independence correction for the variance reduction when
it does not --- is an open problem.

\paragraph{Log-odds aggregation and reporter correlation.} The
trust-weighted log-linear opinion pool (Definition~\ref{def:logodds})
recovers Bayesian updating in the no-tempering limit under conditional
independence of agents' observations given the true risk level. If
reporters' assessments are correlated --- they share information
sources, or one reporter's assessment influences others --- the
aggregation overweights the shared signal. In practice, partial
independence is the norm: agents with different observation channels
and different world models produce partially independent assessments,
and the log-linear pool is an approximation whose error grows with
the correlation. Characterizing the approximation error under partial
independence, and deriving a correction term from a reporter-network
model, is deferred to future work.

\paragraph{Adversarial parameter exposure.} Three of this paper's
mechanisms share a common vulnerability: the adversary can observe
and exploit fixed defense parameters.

\emph{Verification-gate denial of service} (Section~\ref{sec:attention_not_action}).
The salience gate is cheap (aggregate and threshold); the
verification gate is computationally expensive (forward causal
evaluation on the actor's own SCM). An adversary who floods the
actor with structurally plausible false risk claims can exhaust the
verification budget without any single claim bypassing independent
verification. The two-gate architecture defends against coordinated
amplification of false claims into \emph{action} but does not
defend against attention exhaustion via structurally plausible
decoys. The defense direction is prioritization: the verification
gate should process candidates in order of $\Trisk_j$-weighted
expected contraction magnitude, so high-trust reporters' claims are
verified first and low-trust flooding imposes only queue delay, not
budget exhaustion. Formalizing the verification-gate capacity model
and its interaction with the salience threshold is deferred.

\emph{The adversary knows $\epsilon$} (Section~\ref{sec:scorpion}).
The SPRT design parameter $\epsilon$ sets the minimum detectable
contraction magnitude. A well-informed adversary who can infer
$\epsilon$ from the actor's behavior --- by probing with small
contractions and observing flag rates --- will calibrate
$|\mu_j|$ just below $\epsilon$ and sit in the undetected regime
permanently. The evasion-bandwidth bound's ``undetected contraction
rate bounded by $|\mu_j|$'' is only informative if $|\mu_j|$ is
bounded by something other than the design parameter itself.
The defense direction is to treat $\epsilon$ as a nuisance
parameter with an informative prior $p(\epsilon)$ encoding the
actor's belief about the adversary's calibration capability, and
to marginalize. Because the evasion-bandwidth bound
$\tau_{\mathrm{adapt}}(\epsilon) =
O\!\left(\sigma_{\mathrm{do}}^2 /
[\epsilon(|\mu_j| - \epsilon/2)]\right)$
is highly nonlinear in $\epsilon$---small-$\epsilon$ draws produce
very long detection times---the marginalized bound
\begin{equation}
\label{eq:marginalized_evasion}
\mathbb{E}_{p}[\tau_{\mathrm{adapt}}(\epsilon)]
= \int_{\epsilon_{\min}}^{\epsilon_{\max}}
\frac{\sigma_{\mathrm{do}}^2}
     {\epsilon\,(|\mu_j| - \epsilon/2)}\,
p(\epsilon)\,d\epsilon
\end{equation}
is finite for any prior supported on
$[\epsilon_{\min}, \epsilon_{\max}]$ with
$0 < \epsilon_{\min}$ and $\epsilon_{\max} < 2|\mu_j|$
(the detection regime where the SPRT drift is positive),
and the tail bound
$\Pr_{p}(\tau_{\mathrm{adapt}} > T)$ is the prior mass on the
$\epsilon$-region where
$\tau_{\mathrm{adapt}}(\epsilon) > T$---a set that shrinks as
$T$ grows. A log-uniform prior
$p(\epsilon) \propto 1/\epsilon$ on
$[\epsilon_{\min}, \epsilon_{\max}]$ is a natural special case:
it is the maximum-entropy prior over $\log \epsilon$ on bounded
support \citep{jaynes1957information}, equivalently the
scale-invariant maxent prior under the multiplicative
invariance group \citep{jaynes1968prior}, corresponding to
maximum uncertainty about the scale of the adversary's evasion
margin. The defense requires the actor to
\emph{deploy} randomized $\epsilon$---drawing from $p(\epsilon)$
each epoch and keeping the draw private---not merely to
marginalize analytically. Under deployed randomization, the
adversary's optimal evasion strategy must be computed against the
distribution $p(\epsilon)$ rather than a known point estimate---the
minimax-theorem argument for mixed strategies against an optimizing
adversary \citep{vonneumann1928zur}---and
inferring the distribution itself requires observing the actor's
flagging behavior across multiple epochs to estimate
$p(\epsilon \mid \text{flags})$. This is strictly harder than
inverting a known $\epsilon$. The general principle is that fixed,
known parameters are the exploit surface; replacing a point
$\epsilon$ with a privately-drawn $p(\epsilon)$ is one instance of
the randomization-family defense that applies to all three
adversarial surfaces identified above.

\emph{$T_j$ / $\Trisk_j$ independence}
(Section~\ref{sec:risk_trust}). The behavioral trust $T_j$ and
risk-trust $\Trisk_j$ evolve independently by design, optimizing
for the case of honest specialists (good capability reporter, poor
risk forecaster, or vice versa). A strategic reporter can exploit
the independence: operating as a known-unreliable capability
reporter (low $T_j$) while generating structurally plausible risk
claims that accumulate high $\Trisk_j$, thereby driving the
verification-gate queue despite a poor behavioral track record. A
two-sided clamp on risk-trust addresses both failure modes:
\begin{equation}
\label{eq:coupling_clamp}
T^{\mathrm{risk}}_{\min}
\;\leq\; \Trisk_j
\;\leq\; \max\bigl(f(T_j),\; T^{\mathrm{risk}}_{\min}\bigr)
\end{equation}
for some monotone $f$ and a design parameter
$T^{\mathrm{risk}}_{\min} > 0$. The lower bound is the floor:
after each EWMA update the realized $\Trisk_j$ is clipped from
below at $T^{\mathrm{risk}}_{\min}$, so neither an adversary
driving $T_j$ toward zero nor a run of correlated environmental
noise on an honest agent can push risk-trust below the floor,
preventing self-reinforcing exclusion. The upper bound is the
coupling cap: when behavioral trust is high enough that
$f(T_j) \geq T^{\mathrm{risk}}_{\min}$, the cap limits how far
risk-trust can exceed the behaviorally-warranted level, closing
the low-$T_j$ exploit. This is structurally analogous to the
adversarial-balance condition
$c > \max_i(f_i \cdot c_i)$ of Section~\ref{sec:dependency}: a
structural constraint that weakens both failure modes without
claiming to resolve either tightly. The parameter
$T^{\mathrm{risk}}_{\min}$ itself carries a trade-off: too large
re-admits the low-$T_j$ exploit (the cap lifts so far that
risk-trust can remain high despite poor behavioral trust); too
small approaches the unbounded case. The two-sided form does not
eliminate the robustness-versus-fairness tension; it bounds its
worst-case expression on both sides.

The unifying theme is that fixed, observable parameters of the
defense create an adversarial surface. The structural defense
direction across all three is \emph{randomization and coupling}:
randomize the parameters the adversary needs to know (gate
priority, $\epsilon$, channel weights) and couple the trust
channels so that gaming one does not leave the others unaffected.

\paragraph{Shared-prior $\pi_0$ as an adversarial surface.}
The log-odds aggregation (Definition~\ref{def:logodds}) requires
every reporter to calibrate against a shared prior $\pi_0$, and
the paper notes this requires a coordination protocol. The choice
of $\pi_0$ is itself a manipulation surface: a reporter who can
influence $\pi_0$ --- by proposing it, by coordinating on a biased
value, or by exploiting a protocol weakness --- can tilt the
aggregation without submitting a false report. This paper does not
develop a prior-robust solution; the adversarial robustness of
$\pi_0$ establishment is an open problem that interacts with the
adversarial parameter exposure pattern above. We note that
per-agent prior randomization---the natural analog of the
randomized-$\epsilon$ defense---would break the shared-prior
condition on which the log-odds aggregation's Bayes-optimality
depends (Definition~\ref{def:logodds}), so prior-robustness likely
requires re-deriving the aggregation rule under heterogeneous priors
rather than perturbing $\pi_0$.

\paragraph{Structural verification residual norm weights.}
The structural verification residual
(Definition~\ref{def:risk_residual}) uses the Euclidean norm on a
two-dimensional vector (probability difference, normalized magnitude
difference), which implicitly treats both components as equally
important. A probability error of $0.1$ and a magnitude error of
$0.1 \cdot \volL(G_t)$ enter symmetrically into the trust update
but represent different kinds of miscalibration. A weighted norm
with weights tuned to the actor's utility over probability-versus-magnitude
errors would be more principled, but would require the actor to
specify a preference ordering over error types that the framework
does not assume. The equal-weight Euclidean norm is a neutral
default; domain-specific actors with a preference for probability
accuracy over magnitude accuracy (or vice versa) should adjust the
norm accordingly.

\paragraph{Cascade confinement and cross-substrate propagation.}
The dependency-risk analysis (Section~\ref{sec:dependency}) assumes
cascade fractions $c_i$ measure only within-substrate propagation,
with cross-substrate cascades outside the scope of the current
bound. In the motivating threat classes --- memetic attacks,
financial contagion, supply-chain disruption --- cross-substrate
propagation is exactly the concerning pathway: a memetic attack on
silicon agents that triggers panic-driven behavior in biological
agents through inter-substrate communication channels is precisely a
cross-substrate cascade, and the adversarial balance condition does
not cover it. Modeling cross-substrate cascade propagation requires
extending the SCM's structural equations to include inter-substrate
edges, which produces a coupled cascade model whose analysis is
substantially more complex than the within-substrate case.

This gap is also a coupling point with the structural diversification
guarantees developed elsewhere in the framework.
\citet{lasser2026horizon} establishes the minimax result that
$m \geq 2$ substrate types strictly dominate $m = 1$;
Corollary~\ref{cor:m_bound} of the present paper sharpens this to
$m > c_{\max}/c$ under the contagion model of
Section~\ref{sec:dependency}. Both results assume that
substrate-exclusive capabilities do not propagate across the
partition boundary. In the limit of zero cross-substrate leakage
this assumption holds and the bounds are model-level guarantees
under the stated assumptions;
at nonzero leakage rate, the bounds become approximations whose
error depends on the magnitude of inter-substrate information flow.
This does not invalidate the theorems---the proofs are correct for
the model they assume---but it clarifies the empirical condition
under which the guarantees are tight: cross-substrate informational
channels (shared training data, memetic propagation, supply-chain
coupling) must be small relative to the capability differences that
define the partition. The forward direction is either modeling
inter-substrate edges explicitly within this paper's SCM framework
or characterizing the leakage rate empirically so that the
approximation error can be bounded.

\paragraph{Epoch-boundary gaming.}
Proposition~\ref{prop:evasion_bandwidth}'s SPRT bounds are
per-epoch, and an epoch boundary resets the SPRT statistic.
The epoch-declaration protocol is a structural vulnerability:
if the actor declares epoch boundaries based on detected
non-stationarity, an adversary can trigger epoch resets by inducing
structural transitions. If epoch boundaries are time-based (every
$\tau$ steps), an adversary can synchronize strategy shifts to epoch
boundaries so the SPRT never accumulates sufficient evidence.
The defensive direction is a hybrid protocol: time-based epoch
boundaries provide a baseline accumulation window that the adversary
cannot shorten, while non-stationarity detection triggers auxiliary
monitoring (e.g., a parallel SPRT with carry-over statistics) rather
than resetting the primary accumulator. This mitigates the most
direct epoch-manipulation attacks but does not eliminate the
vulnerability, since any fixed-window protocol still gives the
adversary a known accumulation budget to calibrate against.
The hybrid construction does, however, make the adversary's
inversion problem harder: the detection trigger is a function of
SPRT threshold accumulation, which depends on
$\sigma_{\mathrm{do}}^2$, $\epsilon$, and the observed statistics
sequence---and the observed sequence is itself a function of both
the actor's prior monitoring history and the adversary's current
strategy. The joint distribution over (time-based budget phase,
detection-trigger sensitivity) is therefore harder for the adversary
to invert than the pure time-based case, unless it can reconstruct
the actor's internal monitoring state from observable history. The residual vulnerability is that an adversary with
high-bandwidth observation of the actor's flag decisions can
estimate this joint distribution over many epochs, gradually
reducing its entropy. ``Hybrid'' thus narrows the adversarial
surface rather than closing it, consistent with the
adversarial-parameter-exposure theme: fixed parameters leak to
observation; hybrid parameters leak more slowly.
Formalizing this hybrid protocol and its adversarial robustness
is deferred.

