\section{Causal Scorpion Detection}
\label{sec:scorpion}

The foundational paper's Proposition~2 \citep{lasser2026gfm} detects
scorpions through two channels: contraction attribution (correlational) and
deception detection (behavioral trust decay). This section upgrades
channel~(a) from correlational to causal, using the SCM of
Section~\ref{sec:causal}, and provides bounds on non-stationary scorpion
evasion.

\subsection{Causal Detection}

\begin{proposition}[Causal Scorpion Detection]
\label{prop:causal_scorpion}
Under the SCM $\SCM$ (Definition~\ref{def:scm}), a GFM actor can detect
    scorpion agents through two upgraded channels:
\begin{enumerate}
\item[\textbf{(a)}] \textbf{Causal contraction attribution.}
    $\CausalAttr(a_j, t)$ (Definition~\ref{def:causal_attribution}) replaces
    the correlational $\Deltahat$ signal from Proposition~2(a) of
    \citet{lasser2026gfm}. Within each stationary epoch of $a_j$'s
    strategy, the causal-attribution sequence
    $\CausalAttr(a_j, 1), \CausalAttr(a_j, 2), \ldots$ is i.i.d.\
    $\mathcal{N}(\mu_j, \sigma_{\mathrm{do}}^2)$ with unknown mean
    $\mu_j$ and known variance $\sigma_{\mathrm{do}}^2$. (The Gaussian
    assumption is a modeling choice; for non-Gaussian attributions the
    SPRT likelihoods below are replaced by their distributional
    counterparts and the $O(\cdot)$ rate constants change, but the
    structural comparison with the correlational detector is
    preserved.) The
    actor maintains two auxiliary EWMA statistics --- a
    \emph{signed-mean} estimator and a \emph{variance} estimator ---
    for monitoring:
\begin{align}
\label{eq:causal_trust_mean}
\mu_j^{\mathrm{causal}}(t) &= \alpha \cdot \mu_j^{\mathrm{causal}}(t-1)
    + (1 - \alpha) \cdot \CausalAttr(a_j, t) \\
\label{eq:causal_trust_var}
\sigma_j^{\mathrm{causal},2}(t) &= \alpha \cdot
    \sigma_j^{\mathrm{causal},2}(t-1)
    + (1 - \alpha) \cdot
    \bigl(\CausalAttr(a_j, t) - \mu_j^{\mathrm{causal}}(t)\bigr)^2
\end{align}
    These statistics satisfy the same $L^2$ convergence guarantees as
    Proposition~\ref{prop:risk_convergence}: asymptotically unbiased
    with stationary variance bounded by
    $\tfrac{1 - \alpha}{1 + \alpha} \sigma_{\mathrm{do}}^2$ for the
    signed mean. They do not converge almost surely to a deterministic
    limit, and we deliberately do not use them as the flag condition.

    The flag condition is instead a sequential probability ratio test
    (SPRT) \citep{wald1945sequential} on the causal-attribution
    sequence, testing the least-favorable simple hypotheses
    $H_0: \mu_j = 0$ (non-scorpion) versus
    $H_1: \mu_j = -\epsilon$ (scorpion with minimum detectable
    contraction magnitude $\epsilon > 0$). The composite problem
    ($\mu_j \geq 0$ vs.\ $\mu_j \leq -\epsilon$) is reduced to
    this simple pair by the standard least-favorable-distribution
    argument: Type~I error is maximized under the null distribution
    closest to the alternative ($\mu_j = 0$), and Type~II error is
    maximized under the alternative closest to the null
    ($\mu_j = -\epsilon$), so error guarantees for the simple pair
    bound the composite problem from above.
    The actor accumulates the
    log-likelihood ratio
    $\Lambda_j(t) = \sum_{s=1}^{t} \log\bigl(p_{H_1}(\CausalAttr(a_j,
    s)) / p_{H_0}(\CausalAttr(a_j, s))\bigr)$ and applies a
    two-boundary decision rule: flag $a_j$ as a scorpion when
    $\Lambda_j(t) \geq \log((1 - \beta_{\mathrm{err}}) /
    \alpha_{\mathrm{err}})$ (upper boundary), or provisionally clear
    $a_j$ for the current epoch when
    $\Lambda_j(t) \leq \log(\beta_{\mathrm{err}} /
    (1 - \alpha_{\mathrm{err}}))$ (lower boundary). Here
    $\alpha_{\mathrm{err}}$ and $\beta_{\mathrm{err}}$ are the target
    Type~I and Type~II error rates for the simple pair. A provisional
    clearance is epoch-local: the SPRT resets at the next epoch
    boundary, reflecting the possibility that a non-scorpion agent
    may become one. Under this decision rule with
    Gaussian likelihoods, the per-observation expected
    log-likelihood ratio under the true mean $\mu_j$ is
    $\mathbb{E}_{\mu_j}[\log(p_{H_1}/p_{H_0})] =
    \epsilon(|\mu_j| - \epsilon/2) / \sigma_{\mathrm{do}}^2$,
    giving expected sample size
    $\mathbb{E}[\tau_{\mathrm{detect}}] =
    O\bigl(\sigma_{\mathrm{do}}^2 /
    [\epsilon(|\mu_j| - \epsilon/2)]\bigr)$
    by Wald's identity. At the design point $|\mu_j| = \epsilon$
    this reduces to $O(\sigma_{\mathrm{do}}^2 / \epsilon^2)$; for
    $|\mu_j| \gg \epsilon$ it improves to
    $O(\sigma_{\mathrm{do}}^2 / (\epsilon |\mu_j|))$.
    By Markov's inequality the detection probability within
    $k \cdot \mathbb{E}[\tau_{\mathrm{detect}}]$ observations is at
    least $1 - 1/k$ for any $k > 1$. The structural advantage over
    the correlational detector
    (Proposition~\ref{prop:causal_dominates}) is preserved: the
    causal detector operates on variance $\sigma_{\mathrm{do}}^2$
    while the correlational detector operates on $\sigma^2 \geq
    \sigma_{\mathrm{do}}^2$. The signed-mean
    EWMA is retained as a human-interpretable monitor of the
    estimator state; the SPRT statistic $\Lambda_j(t)$ is the
    load-bearing decision variable.
\item[\textbf{(b)}] \textbf{Deception detection (unchanged).} The
    behavioral trust factor $T_j$ detects agents whose reported capability
    changes diverge from observations, through the same prediction-residual
    dynamics as the foundational paper (Equations~25, 33, 38 of
    \citet{lasser2026gfm}). Causal attribution does not affect this channel:
    a deceptive agent accumulates prediction residuals regardless of whether
    the contraction is causally attributed.
\end{enumerate}
\end{proposition}

\begin{proof}
Channel~(a): By the SCM, $\CausalAttr(a_j, t)$ isolates $a_j$'s causal
contribution from confounders
(Definition~\ref{def:causal_attribution}). Within a stationary epoch
of $a_j$'s strategy, $\{\CausalAttr(a_j, t)\}$ is i.i.d.\ with mean
$\mu_j$ and variance $\sigma_{\mathrm{do}}^2$; for a scorpion
$\mu_j \leq -\epsilon < 0$. The auxiliary EWMAs
$\mu_j^{\mathrm{causal}}(t)$ and $\sigma_j^{\mathrm{causal},2}(t)$
inherit the $L^2$ guarantees of
Proposition~\ref{prop:risk_convergence}: asymptotically unbiased
estimators of $\mu_j$ and $\sigma_{\mathrm{do}}^2$ with bounded
stationary variance. They are not used for the detection decision,
so we do not require almost-sure convergence.

The SPRT is applied to the least-favorable simple pair
$H_0: \mu_j = 0$ versus $H_1: \mu_j = -\epsilon$
(see channel~(a) above for the reduction from composite to simple
hypotheses). Under the Gaussian model, the per-observation expected
log-likelihood ratio under the true mean $\mu_j$ is
$\mathbb{E}_{\mu_j}[\log(p_{H_1}/p_{H_0})] =
\epsilon(|\mu_j| - \epsilon/2) / \sigma_{\mathrm{do}}^2$. By Wald's
identity, the expected detection time under true $\mu_j$ is
$\mathbb{E}[\tau_{\mathrm{detect}}] \approx
\log((1 - \beta_{\mathrm{err}}) / \alpha_{\mathrm{err}}) \cdot
\sigma_{\mathrm{do}}^2 /
[\epsilon(|\mu_j| - \epsilon/2)]$, which at the design point
$|\mu_j| = \epsilon$ reduces to $O(\sigma_{\mathrm{do}}^2 / \epsilon^2)$
and for $|\mu_j| \gg \epsilon$ improves to
$O(\sigma_{\mathrm{do}}^2 / (\epsilon |\mu_j|))$.
The SPRT Type~I error for the simple pair is bounded by
$\alpha_{\mathrm{err}}$ and Type~II by $\beta_{\mathrm{err}}$; the
least-favorable argument ensures these bounds hold for the
composite problem.

Channel~(b): The behavioral trust dynamics (Appendix~B of
\citet{lasser2026gfm}) operate on behavioral prediction residuals
$r_j(t) = R_j(t) - \hat{R}_j(t)$, which are independent of causal
attribution. A deceptive scorpion that misreports capability changes
accumulates $r_j(t) \neq 0$ regardless of confounders, driving
$T_j \to T_j^* < 1$.
\end{proof}

\subsection{Non-Stationary Scorpion Bounds}

The foundational paper notes that ``a non-stationary scorpion that adapts its strategy in
response to detection pressure can evade both channels indefinitely''
(Proposition~2 of \citet{lasser2026gfm}). Under the SCM, we can bound the
adaptation rate required for evasion.

\begin{proposition}[Evasion Bandwidth]
\label{prop:evasion_bandwidth}
A scorpion $a_j$ with per-epoch mean causal contraction $|\mu_j|$
    that changes strategy every $\tau_{\mathrm{adapt}}$ steps
    produces at most $\tau_{\mathrm{adapt}}$ observations per strategy
    epoch. Let
    $\bar{\tau}(\mu_j) = O\bigl(\sigma_{\mathrm{do}}^2 /
    [\epsilon(|\mu_j| - \epsilon/2)]\bigr)$ be the expected detection
    time per epoch from the SPRT under the Gaussian model with
    design parameter $\epsilon$
    (Proposition~\ref{prop:causal_scorpion}). Then:
\begin{enumerate}
\item[\textbf{(a)}] If
    $\tau_{\mathrm{adapt}} \geq k \cdot \bar{\tau}(\mu_j)$ for some
    $k > 1$, the scorpion is detected within each epoch with
    probability at least $1 - 1/k$ (by Markov's inequality on the
    SPRT stopping time).
\item[\textbf{(b)}] If $\tau_{\mathrm{adapt}} < \bar{\tau}(\mu_j)$,
    the expected accumulated evidence at the epoch boundary is below
    the SPRT threshold. While this does not by itself guarantee
    evasion with any specific probability, it establishes that the
    expected number of observations is insufficient for detection,
    so the SPRT is unlikely to trigger within the epoch.
\end{enumerate}
\end{proposition}

\begin{proof}
Within each epoch of length $\tau_{\mathrm{adapt}}$, the scorpion's
strategy is stationary and the causal-attribution sequence is i.i.d.\
$\mathcal{N}(\mu_j, \sigma_{\mathrm{do}}^2)$. The SPRT
statistic accumulates log-likelihood-ratio evidence at expected rate
$\epsilon(|\mu_j| - \epsilon/2) / \sigma_{\mathrm{do}}^2$ per
observation under the Gaussian model
(Proposition~\ref{prop:causal_scorpion}), giving expected detection time
$\bar{\tau}(\mu_j) =
\sigma_{\mathrm{do}}^2 \cdot
\log((1 - \beta_{\mathrm{err}}) / \alpha_{\mathrm{err}}) /
[\epsilon(|\mu_j| - \epsilon/2)]$.

(a) By Markov's inequality,
$P(\tau_{\mathrm{detect}} > k \bar{\tau}) \leq 1/k$ for any $k > 1$.
When $\tau_{\mathrm{adapt}} \geq k \bar{\tau}(\mu_j)$, the epoch
provides enough observations that the SPRT crosses the detection
threshold with probability at least $1 - 1/k$.

(b) An epoch boundary resets the SPRT statistic (the hypotheses are
epoch-local). If $\tau_{\mathrm{adapt}} < \bar{\tau}(\mu_j)$, the
expected accumulated evidence at the epoch boundary is below the
SPRT threshold. Since the number of observations is less than the
expected detection time, the SPRT is unlikely to trigger within the
epoch: the evidence accumulation rate is insufficient for reliable
detection.
\end{proof}

\paragraph{Interpretation.} The evasion bandwidth
$\tau_{\mathrm{adapt}}^* = O\bigl(\sigma_{\mathrm{do}}^2 /
[\epsilon(|\mu_j| - \epsilon/2)]\bigr)$ is the
characteristic timescale: a scorpion that adapts faster than
$\tau_{\mathrm{adapt}}^*$ evades with non-negligible probability per
epoch, while one that adapts slower than $k \cdot
\tau_{\mathrm{adapt}}^*$ is detected with probability at least
$1 - 1/k$.
A fast-adapting scorpion ($\tau_{\mathrm{adapt}} \ll
\tau_{\mathrm{adapt}}^*$) evades but is constrained to small contractions
per epoch (otherwise $\mu_j$ is large and $\tau_{\mathrm{adapt}}^*$ shrinks).
A slow-adapting scorpion ($\tau_{\mathrm{adapt}} \gg
\tau_{\mathrm{adapt}}^*$) is detected within each epoch with high
probability. The product $\mu_j \cdot \tau_{\mathrm{adapt}}$ (total
contraction per epoch) is bounded: a scorpion can either cause large
damage slowly (and be detected) or small damage quickly (and evade),
but not large damage quickly.

The time-averaged undetected contraction rate for an evading scorpion
is bounded by $|\mu_j|$: within each epoch the per-step contraction
is $|\mu_j|$, and the evasion constraint
$\tau_{\mathrm{adapt}} < \bar{\tau}(\mu_j)$ bounds the epoch length.
The appendix proof (§\ref{app:proof_evasion}) establishes this via
the SPRT accumulation rate. The key trade-off is controlled by the
SPRT design parameter $\epsilon$: a scorpion whose per-step
contraction satisfies $|\mu_j| \leq \epsilon$ requires at least
$\bar{\tau}(\epsilon) = O(\sigma_{\mathrm{do}}^2 / \epsilon^2)$
observations to detect (the worst case at the design point), while
a scorpion with $|\mu_j| \gg \epsilon$ is detected faster in
proportion to $|\mu_j| / \epsilon$. The SPRT design parameter
$\epsilon$ therefore sets the resolution: reducing $\epsilon$ detects
smaller contractions but requires more observations per epoch.
The bound tightens as the causal model improves ($\sigma_{\mathrm{do}}$
decreases) and as the scorpion's adaptation budget
$\tau_{\mathrm{adapt}}$ grows: a scorpion that adapts frequently
(small $\tau_{\mathrm{adapt}}$) can only afford very small per-step
contractions, while a scorpion that adapts slowly can sustain larger
contractions per step but for correspondingly fewer steps before
detection.
