\section{Discussion}
\label{sec:discussion}

\subsection{What this paper accomplishes}

\paragraph{Breaks the Wamura problem.}
Theorem~\ref{thm:convergence} shows that controlled relaxation
drives risk estimates to truth at exponential rate, converting
``permanent overcaution'' into ``temporary overcaution with a
characterized resolution time.''  This closes the most important
named open problem of~\citep{lasser2026horizon}.

\paragraph{Substitutes experimental evidence for catastrophe-as-teacher.}
The classical solution to long-timeline-risk uncertainty is to
wait for the first catastrophe to write the historical markers:
the tsunami stones of Fudai, the building-code revisions that
followed Galveston 1900, the pharmacovigilance regimes that
followed thalidomide.  That mechanism is available only if the
first catastrophe has already been paid for and the markers still
survive in a legible form.  Neither condition holds for
capabilities that did not exist in previous generations---which
is precisely the regime the framework's $\volL$-maximising agents
generate.  Controlled relaxation provides experimental evidence
with the evidential content historical markers would have
recorded---did the pathway activate?  how severe was the
contraction?  which keystones blocked it?---but acquires the
evidence within the damage budget $\damagebound$ rather than the
cost-of-life that produced historical markers.  This is the
practical import of Theorem~\ref{thm:damage_bound}: the cost of
learning is bounded and chosen by the framework, not paid by the
next generation.

\paragraph{Provides a damage-bounded mechanism.}
Theorem~\ref{thm:damage_bound} bounds the maximum damage from a
test by the user-controllable parameter $\damagebound$.
Controlled relaxation is therefore safe to deploy: the cost of
learning is bounded even when the risk turns out to be real, and
the cooperative-closure condition of
Definition~\ref{def:test_scope}(ii) prevents cross-boundary
cascades through shared cooperative capabilities.

\paragraph{Integrates with the existing machinery.}
Propositions~\ref{prop:preserve_balance},
\ref{prop:empirical_arbitration} and Lemma~\ref{lem:sixth_channel}
show that the protocol preserves the self-balancing property of
$\volL$, resolves inter-agent risk disagreements, and extends
the multi-channel detection architecture.  The protocol is
additive to the existing framework, not a replacement.

\paragraph{Resolves the Loop 3 phase-boundary degeneracy.}
The degenerate phase boundary $\rS = 0$ for Loop~3
in~\citep[Proposition~4]{lasser2026phase} arises precisely
because the Wamura asymmetry eliminates the exogenous
correction rate on the restricted dimension.  Controlled
relaxation makes $\rS > 0$ for Loop~3 (via the exercise-regime
observations of Section~\ref{sec:relaxation_protocol}),
restoring the phase boundary to the non-degenerate form that
the design criterion
of~\citep[Theorem~2]{lasser2026phase} requires for
self-correction.  The two papers are complementary: one
characterizes when self-correction works in aggregate; this one
supplies the micro-mechanism that makes self-correction work
for the restriction-specific feedback loop.

\subsection{Limitations}

\paragraph{B-to-C gap inheritance.}
All results are stated in terms of $\volL$.
Section~\ref{sec:framing_precondition} records that $\damagebound$
bounds contraction of $\volL$, not of the experiential target
directly; if the gap between them is large, even a
controlled-relaxation-clean restriction landscape need not
correspond to experiential optimality.  The gap-closure programme
is a separate research direction, noted
in~\citep{lasser2026gfm} and outside the scope of this paper.

\paragraph{Honest-channel assumption.}
We assume honest observation channels throughout.  Adversarial
manipulation of the detection signal---a compromised channel
reporting fabricated contraction or concealing genuine
contraction---falls outside the scope of the proof and is the
domain of the commitment protocols of~\citep{lasser2026exo}.
Controlled relaxation composes cleanly with those
protocols---the ledger of~\citep[Section~5]{lasser2026exo} can
record test outcomes as committed observations---but the
composition itself is not formalized here.

\paragraph{Identifiability under misspecification.}
Assumption~\ref{ass:T2} requires the Bayesian observation model
to match the true data-generating process.  If $\alphafp$ or
$\beta$ are misspecified, Theorem~\ref{thm:convergence}'s
log-likelihood-ratio argument still converges, but it converges
to the wrong hypothesis: under the misspecified model, the
posterior concentrates on whichever hypothesis is closer in
KL-divergence to the true distribution, not on the true
hypothesis.  A concrete sign-preservation bound delineates the
safe region: the convergent hypothesis is correct iff the
modeled likelihoods $\Lzero$ and $\Lone$ have the correct sign
relative to the true distribution---formally,
$E_{H_0^{\mathrm{true}}}[\log L] < 0$ and
$E_{H_1^{\mathrm{true}}}[\log L] > 0$.  In the binary model,
this reduces to: if Assumption~\ref{ass:T1} holds for the
modeled $(\beta, \alphafp)$ and the true $(\beta^\star,
\alphafp^\star)$ jointly satisfy
$(1 - \beta^\star)/\alphafp^\star > 1$ and
$\beta^\star/(1 - \alphafp^\star) < 1$ (i.e., the true detector
also has nontrivial power), then the convergent hypothesis is
correct, with rate degraded by the KL gap between modeled and
true distributions.  Misspecification flips the convergent
hypothesis only when the modeled detector is qualitatively
miscalibrated (modeled $\Lone$ on the wrong side of $1$);
quantitative misestimation only slows convergence.  Concretely,
the safe region in $(\beta^{\star}, \alphafp^{\star})$ space is
characterised by:
\begin{center}
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{True hypothesis} & \textbf{Safe-region boundary}
  & \textbf{Worked-example value} \\
\midrule
$H_0$ true & $\alphafp^{\star} <
  \frac{|\log \Lzero|}{|\log \Lzero| + \log \Lone}$
  & $\alphafp^{\star} < 0.34$ \\[0.5em]
$H_1$ true & $\beta^{\star} <
  \frac{\log \Lone}{|\log \Lzero| + \log \Lone}$
  & $\beta^{\star} < 0.66$ \\
\bottomrule
\end{tabular}
\end{center}
\noindent (Worked-example values use $\Lzero = 0.101$,
$\Lone = 90$ from $\beta = 0.10$, $\alphafp = 0.01$,
$\thetaone = 1$.)  The protocol tolerates a true
$\alphafp^{\star}$ up to ${\sim}34\times$ the modelled value and
a true $\beta^{\star}$ up to ${\sim}6.6\times$ the modelled
value before the convergent hypothesis flips.  Detector
calibration is therefore a prerequisite for the convergence
guarantee to carry its safety interpretation, but the bound is
robust to plausible-magnitude calibration error.

The deeper problem is that calibration itself reintroduces a
verification asymmetry one level up.  For first-of-kind
capabilities---which is precisely the regime where the Wamura
problem is most acute---there is no historical contraction
record on which to estimate $\beta$ and $\alphafp$ for the
specific pathway under test, and the restricted regime by
construction generates none.  Calling this a bootstrap problem
is more honest than treating it as a side condition: the
detector that probes a novel pathway must itself be calibrated,
and the calibration cannot come from observing the very
restricted regime CR is intended to break.  Two partial remedies
are available.  (i)~Cross-pathway calibration: the same detector
architecture, applied to non-restricted but structurally similar
pathways, yields $\beta$ and $\alphafp$ estimates that transfer
under the assumption that detection physics is pathway-class
invariant rather than pathway-specific.  (ii)~Cross-substrate
commitment: the integrity machinery of~\citep{lasser2026exo}
permits committed predictions about what the detector should
observe under benign and adversarial conditions, with the
commitment itself constituting a calibration witness independent
of the restricted-regime data the bootstrap lacks.  Both are
partial; neither closes the bootstrap completely.  We name the
gap rather than dissolve it.

The asymmetry between the two misspecification failure modes is
worth contrasting explicitly.  Upward $p_{\mathrm{ex}}$
misspecification is \emph{detectable from within the protocol}:
observed exercise frequency $\hat{p}_{\mathrm{ex}}$ falls
visibly below the modelled value, and
Criterion~\ref{def:test_initiation}(v) catches it.  Detector
miscalibration of $\alphafp$ or $\beta$, by contrast, is
\emph{undetectable from within}: under a false $H_1$ claim with
a miscalibrated detector, the protocol observes exactly the run
of $O_n = 0$ it expects, and nothing internal to the test
sequence can distinguish ``correctly clearing a false claim''
from ``converging to the wrong hypothesis.''
$p_{\mathrm{ex}}$ misspecification slows convergence; detector
miscalibration flips it; the former is operationally handleable
by the meta-policy, the latter must be ruled out externally.

\paragraph{Scope-existence for densely cooperative capabilities.}
Remark~\ref{rem:scope_existence} names scope existence as a
scheduling precondition that must be enforced by the meta-policy
before a CR test is initiated.  For application domains where
cooperative-capability structure is dense, this precondition is
more than a technical caveat.  AI capability spaces in
particular have the property that almost every capability
cooperates with almost every other through the shared
computational substrate; in such spaces, $\maxcon(P)$ on the full
poset may dominate any acceptable $\damagebound$, and any scope
satisfying Definition~\ref{def:test_scope} may be similarly
infeasible.  We do not have a quantitative estimate for how
often feasible scopes exist in AI capability domains.  For some
risk classes---capabilities with strongly localised cooperative
structure (e.g., specific tool-use patterns confined to a
domain)---feasible scopes plausibly exist; for others
(capabilities with broad cooperative reach, e.g., general
reasoning) they may not.  When feasible scopes do not exist, the
alternative mechanisms named in
Remark~\ref{rem:scope_existence}---structural discovery
\citep[Proposition~2]{lasser2026horizon} or scope-refinement
that trades coverage for feasibility---must be deployed instead.
Controlled relaxation is not a universal remedy for the Wamura
problem; it is a remedy for the subset of risks where bounded
scopes exist.

\subsection{Open questions}

\paragraph{Damage-tolerance calibration.}
The tolerance $\damagebound$ is a free parameter that trades off
learning speed against risk exposure.  Options include
\begin{itemize}
\item a fixed fraction of $\volL$ (e.g.,
  $\damagebound = 0.01 \cdot \volL$);
\item proportional to restriction cost
  ($\damagebound = c \cdot |\Delta \volL(\Xirestrict)|$);
\item proportional to information value
  ($\damagebound = c' \cdot \IV(T)$);
\item user-set per test (maximum safety, maximum user burden).
\end{itemize}
The worked example (Section~\ref{sec:worked_example}) shows that
even small $\damagebound$ produces rapid convergence when
detection sensitivity is high, so the answer may depend more on
the detection architecture than on $\damagebound$ itself.

\paragraph{Transient pathways with unknown duration.}
The composition rule of
Proposition~\ref{prop:transient_design} addresses the case where
the pathway duration $k$ is known or bounded above.  When $k$ is
itself unknown---the detector must infer pathway duration from
limited test data---the design constraint becomes circular: a
detector tuned for assumed $k$ will under-perform if the true
duration is shorter.  Conservative practice uses the smallest
plausible $k$ (typically $k = 1$) for the design rule, accepting
the corresponding strong $\deltaagg$ requirement.  A principled
joint inference of $k$ and $\thetaone$ from test data is an open
question.

\paragraph{Adversarial risk claimants.}
Assumption~\ref{ass:T3}(a) rules out adaptation by the tested
agent.  What happens when an agent strategically inflates risk
claims to lock competitors' capabilities?  The controlled
relaxation protocol still works (it generates evidence
regardless of the claimant's intent), but the meta-policy for
test initiation may need to be more aggressive when strategic
inflation is suspected.  Detection of strategic inflation is a
variant of the scorpion-detection
problem~\citep[Propositions~3--4]{lasser2026scm}: an agent
whose risk claims are systematically disconfirmed by
controlled-relaxation tests should see its risk-trust
$\Trisk_j$ decrease, eventually down-weighting its future
claims in aggregation.  Whether the meta-policy should
explicitly model this adversarial case, or whether the
trust-dynamics response is sufficient, is an open design
question.

The composition is operationally fragile during the
$\testdur$-window itself.  The Wamura problem is a structural
invitation for an adversary---an incumbent, a regulator, or a
rent-seeker---to manufacture risk claims that lock competitor
capabilities for the duration of test scheduling and execution,
even if the asymptotic trust dynamics eventually correct.  The
peak instantaneous $\volL$ locked by concurrently-active
adversarial claims is bounded above by
$|\Delta \volL(\Xirestrict)| \cdot N_{\mathrm{adv}}$, and the
total \emph{capability-loss exposure} ($\volL$ integrated over
time) attributable to the adversarial claims before trust-weight
downweighting takes effect is bounded above by
$|\Delta \volL(\Xirestrict)| \cdot N_{\mathrm{adv}} \cdot
(T_{\mathrm{stale}} + \testdur)$, where $N_{\mathrm{adv}}$ is
the number of adversarial claims the trust dynamics absorb
before downweighting and $T_{\mathrm{stale}} + \testdur$ is the
per-claim duration each ties up the targeted capability.  In
contested commercial or strategic settings where adversarial
claim manufacture is plausible, this bound is the correct
deployment guard: the protocol should not be deployed without
either (i) an integrity layer (cryptographic commitment via
\citep{lasser2026exo}, institutional review, or jurisdictional
accountability) that bounds $N_{\mathrm{adv}}$ at the claim
generation step, or (ii) a meta-policy with an explicit
adversarial-claim detector before trust dynamics take over.
The current protocol carries no such guard internally;
deployment in contested settings requires the integrity layer be
specified externally and is named here as a precondition for
operational use.  The natural analogue of Criterion~(v)'s online
check for $p_{\mathrm{ex}}$ misspecification would be an online
check for the adversarial-claim rate: track the observed rate of
claims that are subsequently disconfirmed by CR tests, and raise
an alarm when it exceeds the integrity layer's assumed
$N_{\mathrm{adv}}$ bound.  This a~posteriori guard is future
work; we note the symmetry with Criterion~(v) as a design
pattern.

\subsection{Relationship to prior and subsequent papers}

\paragraph{Prior papers.}
The protocol sits downstream of and draws on:
\begin{itemize}
\item \citet{lasser2026poset} for axiomatic properties of
  $\volL$ preserved under relaxation;
\item \citet{lasser2026horizon} for the risk machinery (exercise
  pathways, concentration risk, restriction criterion) the
  protocol operates within;
\item \citet{lasser2026scm} for the multi-channel detection
  architecture that supplies $\deltaagg$ and the log-odds
  aggregation into which test outcomes feed;
\item \citet{lasser2026exo} for the verification-asymmetry
  framing and the cross-substrate isolation that justifies the
  conservative detection bound
  (Proposition~\ref{prop:detection_bounds}(c));
\item \citet{lasser2026phase} for the dynamical-systems context:
  Loop~3's degenerate phase boundary is the problem this paper
  solves.
\end{itemize}

\paragraph{Forward connections.}
Controlled relaxation generates exercise events that increase
realized-capability volume $\vol_{\mathrm{R}}$: when a capability
is locked by a false risk claim, $\vol_{\mathrm{R}}$ is
artificially suppressed, and lifting the lock lets the
capability contribute to $\vol_{\mathrm{R}}$.  The convergence
of the restriction landscape
(Proposition~\ref{prop:preserve_balance}(c)) implies that
$\vol_{\mathrm{R}} / \volL$ approaches its true value as false
restrictions clear, partially closing the B-to-C gap for the
class of capabilities locked by the Wamura problem.  A full
treatment of this partial closure is beyond the scope of this
paper.

The protocol also provides a clean experimental signal for
structure learning: the SCM
of~\citep[Definition~1]{lasser2026scm} can use designed-experiment
observations from controlled relaxation (where the intervention
is known and the outcome is observed) as ground-truth data
points for calibrating structural confidence.  This is strictly
more informative than passive observational data and could
accelerate SCM calibration beyond the rates achievable under
passive observation alone.
