% =============================================================================
% APPENDIX D: AGENT SIMILARITY AND GOAL ESTIMATION
%
% Adapted from predraft Sections XI-XII with notation rebuilt
% from this paper's definitions. Provides the metrics and mechanisms
% by which a GFM actor estimates other agents' goals, determines
% agent identity, and groups agents by goal profile.
% =============================================================================

\section{Observation Channel: Agent Identity and Goal Estimation}
\label{app:agent_similarity}

The trust factor $T_k$ (Definition~\ref{def:observable_goal_model}) and the
aggregate signal $\Deltahat(\pi)$ (Equation~\eqref{eq:aggregate_signal})
presuppose that the GFM actor can attach persistent identity to the agents it
observes: the actor must know that the $a_k$ reporting today is the same $a_k$
whose prediction residuals it has been tracking. Without a mechanism for
re-identification, $T_k$ cannot accumulate across interactions, and the trust
model collapses. This appendix sketches the observation-channel infrastructure
that makes persistent identity and goal estimation possible. The material is an
implementation sketch rather than a formal extension of the measure-theoretic
framework; it addresses the engineering question of \emph{how} the observation
set $\mathcal{O}$ is maintained, not the formal question of what $\vol(G)$
measures.

The appendix develops four components: a goal similarity metric that determines
when two goals are close in capability space, an agent similarity metric that
determines when two observations correspond to the same agent, a
zero-knowledge proof mechanism for the strongest form of identity verification,
and a clustering procedure that groups agents into modal categories by goal
profile. Together, these provide the substrate on which the trust model
(Appendix~\ref{app:trust_model}) and the local volume estimator
(Definition~\ref{def:local_volume_estimator}) operate.

% ---------------------------------------------------------------------------
% D.1: Goal Similarity Metric S_G
% ---------------------------------------------------------------------------

\subsection{Goal Similarity Metric \texorpdfstring{$S_G$}{S\_G}}
\label{app:goal_similarity}

A GFM actor proposing actions to expand $\vol(G)$ must determine which goals are similar---close in the capability space---and which are distant. Without a similarity metric, the actor cannot generalize from observed goals to unobserved ones, cannot identify redundant capabilities, and cannot efficiently allocate effort across the population.

\begin{definition}[Goal Similarity]
\label{def:goal_similarity}
For two goals $g_i, g_j \in \G$, the \emph{goal similarity} $S_G(g_i, g_j)$ is defined as the overlap in their projected capability contributions, conditioned on the GFM actor's current world model $M$:
\begin{equation}
\label{eq:goal_similarity}
S_G(g_i, g_j \mid M) = \frac{\vol\bigl(\Phi_M(g_i) \cap \Phi_M(g_j)\bigr)}{\vol\bigl(\Phi_M(g_i) \cup \Phi_M(g_j)\bigr)}
\end{equation}
where $\Phi_M(g)$ is the \emph{projected capability region}---the set of capabilities in $\G$ that the world model $M$ associates with goal $g$, including both the direct capability and the instrumental capabilities required to realize it.
\end{definition}

The metric has the structure of a Jaccard index over projected capability regions: $S_G = 1$ when two goals project to identical capability sets (they are the same goal under different descriptions), and $S_G = 0$ when their projections are disjoint (they share no instrumental or terminal capabilities). The conditioning on $M$ is essential---two goals that appear similar under one world model may be dissimilar under another, since the instrumental capabilities connecting them depend on the model's beliefs about the environment.

\paragraph{Properties.} $S_G$ satisfies:
\begin{itemize}
\item \emph{Symmetry:} $S_G(g_i, g_j \mid M) = S_G(g_j, g_i \mid M)$.
\item \emph{Boundedness:} $S_G \in [0, 1]$ by construction.
\item \emph{Model-dependence:} $S_G$ is a property of goals-as-projected-through-a-model, not an intrinsic property of goals. This dependence is constructive: the similarity metric improves as the world model improves, and two GFM actors with different world models may assign different similarities to the same goal pair.
\end{itemize}

\paragraph{Relation to $\vol(G)$.} Goal similarity connects to the optimization objective through redundancy. When $S_G(g_i, g_j \mid M)$ is high, the two goals contribute overlapping capability volume to $G$. Expanding both adds less to $\vol(G)$ than expanding two dissimilar goals, because the intersection is counted only once. A GFM actor seeking to maximize $\Delta \vol(G)$ per unit effort should preferentially expand goals with low pairwise similarity---diversifying the capability space rather than deepening existing niches. This provides a formal basis for the intuition that diversity is instrumentally valuable: a population with diverse goals has higher $\vol(G)$ than one with uniform goals, all else equal, because the union of dissimilar sets has greater measure than the union of similar ones.

% ---------------------------------------------------------------------------
% D.2: Agent Similarity Metric S_A
% ---------------------------------------------------------------------------

\subsection{Agent Similarity Metric \texorpdfstring{$S_A$}{S\_A}}
\label{app:agent_similarity_metric}

Goal similarity measures the relationship between goals; agent similarity measures the relationship between the entities pursuing them. The GFM actor needs agent similarity for a different purpose: to determine whether two observations correspond to the same agent (re-identification), whether an agent's behavior is consistent over time (stability), and how much weight to assign an agent's reports in the estimator (trust).

\begin{definition}[Agent Similarity]
\label{def:agent_similarity}
For two agent observations $a_i(t_1)$ and $a_j(t_2)$ (potentially the same agent at different times), the \emph{agent similarity} $S_A$ is defined as:
\begin{equation}
\label{eq:agent_similarity}
S_A\bigl(a_i(t_1), a_j(t_2)\bigr) = \max\Bigl(\underbrace{S_{\mathrm{beh}}\bigl(a_i(t_1), a_j(t_2)\bigr)}_{\text{behavioral consistency}},\; \underbrace{S_{\mathrm{zkp}}\bigl(a_i(t_1), a_j(t_2)\bigr)}_{\text{identity verification}}\Bigr)
\end{equation}
where $S_{\mathrm{beh}}$ measures behavioral consistency and $S_{\mathrm{zkp}}$ measures cryptographic identity verification. The $\max$ operator reflects that either channel alone is sufficient for identification: a cryptographic proof of identity overrides behavioral evidence, and strong behavioral consistency can establish practical identity in the absence of cryptographic proof.
\end{definition}

The behavioral component $S_{\mathrm{beh}}$ aggregates multiple observable features of the agent:
\begin{equation}
\label{eq:behavioral_similarity}
S_{\mathrm{beh}}\bigl(a_i(t_1), a_j(t_2)\bigr) = \sum_{\ell} w_\ell \cdot s_\ell\bigl(a_i(t_1), a_j(t_2)\bigr)
\end{equation}
where each $s_\ell \in [0, 1]$ measures consistency along a specific dimension and $w_\ell$ are learned weights summing to 1. The dimensions include:
\begin{enumerate}
\item \emph{Goal consistency:} $S_G\bigl(\Gind{i}(t_1), \Gind{j}(t_2) \mid M\bigr)$---do the two observations project to similar capability sets?
\item \emph{Behavioral patterns:} Consistency of action selection, communication style, and response timing.
\item \emph{Attentional focus:} Consistency of which aspects of the environment the agent attends to and which it ignores.
\item \emph{Environmental context:} Consistency of spatial, temporal, and social context---an agent observed in incompatible locations simultaneously cannot be the same entity.
\end{enumerate}

The weights $w_\ell$ are learned from the GFM actor's interaction history: dimensions that have historically been more predictive of identity receive higher weight. This learning is itself subject to adversarial manipulation (an agent could mimic another's behavioral patterns), which is why the cryptographic component $S_{\mathrm{zkp}}$ exists as an independent verification channel.

\paragraph{Relation to trust.} Agent similarity connects to the trust model through the observable goal model. When $S_A\bigl(a_k(t), a_k(t-1)\bigr)$ is high---the agent appears consistent over time---the prediction residual $r_k(t)$ (Equation~\ref{eq:prediction_residual}) will be small, and trust $T_k$ will be maintained or increased. When $S_A$ drops (the agent appears inconsistent, as though it has been replaced or is behaving erratically), the prediction residual increases and trust decreases. The agent similarity metric thus provides the input signal that drives the trust update dynamics of Appendix~\ref{app:trust_model}.

% ---------------------------------------------------------------------------
% D.3: Zero-Knowledge Proof for Re-identification
% ---------------------------------------------------------------------------

\subsection{Zero-Knowledge Proof for Agent Re-identification}
\label{app:zkp}

The strongest form of agent identity verification is cryptographic: an agent proves it is the same entity the GFM actor interacted with previously, without revealing any private information beyond the fact of identity.

\begin{definition}[Identity Proof]
\label{def:identity_proof}
An agent $a_k$ can establish identity with the GFM actor through a zero-knowledge proof $P_{\mathrm{zk}}$ if the following conditions hold:
\begin{enumerate}
\item \emph{Prior interaction:} The GFM actor and $a_k$ have established a shared secret or commitment during a previous interaction at time $t_0$.
\item \emph{Challenge-response:} The GFM actor issues a challenge $c$ derived from the shared commitment. The agent produces a response $r$ such that $\mathrm{Verify}(c, r, \mathrm{commitment}) = 1$.
\item \emph{Zero-knowledge:} The response $r$ reveals no information about the shared secret beyond the fact that the responder possesses it.
\end{enumerate}
When the proof succeeds, the identity verification component is set to $S_{\mathrm{zkp}} = 1$, and the GFM actor can directly assign the prior relationship's accumulated trust and goal model to the current observation:
\begin{equation}
\label{eq:zkp_identity}
P_{\mathrm{zk}}\bigl(a_k(t), a_k(t_0)\bigr) = 1 \;\implies\; S_A\bigl(a_k(t), a_k(t_0)\bigr) = 1
\end{equation}
\end{definition}

The zero-knowledge proof serves as an \emph{override} for the behavioral similarity metric. An agent whose behavior has changed significantly (perhaps because its capabilities have expanded, its environment has shifted, or it has undergone internal restructuring) may have low $S_{\mathrm{beh}}$ despite being the same entity. The cryptographic proof resolves this ambiguity definitively: if the proof succeeds, the agent is the same entity regardless of behavioral divergence.

\paragraph{Applications.} The primary use of zero-knowledge identity proofs in the GFM framework is \emph{trust persistence}. When an agent $a_k$ goes unobserved for an extended period and reappears, the GFM actor faces a choice: treat $a_k$ as a new agent (low initial trust, cooling period) or as a returning agent (prior trust level, accumulated model). Without cryptographic verification, the actor must rely on behavioral similarity, which degrades over time as the agent's behavior evolves. With a zero-knowledge proof, the agent can immediately recover its prior trust level, bypassing the cooling period.

A second application is defense against impersonation. An adversary $a_s$ that mimics agent $a_k$'s behavioral patterns could achieve high $S_{\mathrm{beh}}$ and inherit $a_k$'s trust level---a spoofing attack. The zero-knowledge proof prevents this: an impersonator cannot produce a valid response to the challenge without possessing the shared secret, regardless of how accurately it mimics $a_k$'s observable behavior.

\paragraph{Limitations.} The zero-knowledge proof requires a prior interaction during which the shared commitment is established. It cannot verify the identity of a truly novel agent. It also assumes that the agent's private key (or equivalent secret) has not been compromised---if an adversary obtains the shared secret, it can pass the identity proof and inherit the victim's trust. The proof establishes that the current entity \emph{possesses the same secret} as the prior interactant, not that it \emph{is} the same entity in any deeper sense. For practical purposes, this distinction matters only when secrets can be stolen, which is an implementation security question rather than a framework question.

% ---------------------------------------------------------------------------
% D.4: Modal Groupings
% ---------------------------------------------------------------------------

\subsection{Modal Groupings}
\label{app:modal_groupings}

As a GFM actor accumulates goal models $\{M_k(G)\}$ across many agents, structure emerges: clusters of agents with similar goal profiles, pursuing overlapping capabilities through similar strategies. These clusters are \emph{modal groupings}: the modes of the goal distribution across the observed population.

\begin{definition}[Modal Grouping]
\label{def:modal_grouping}
A \emph{modal grouping} $\mathcal{M}_j$ is a subset of observed agents whose pairwise goal similarity exceeds a threshold $\eta$:
\begin{equation}
\label{eq:modal_grouping}
\mathcal{M}_j = \bigl\{a_k \in \mathcal{O} : S_G(\Gind{k}, \mu_j \mid M) > \eta\bigr\}
\end{equation}
where $\mu_j$ is the centroid of the grouping in projected capability space and $\mathcal{O}$ is the GFM actor's observation set. The set of all modal groupings $\{\mathcal{M}_1, \ldots, \mathcal{M}_m\}$ partitions the observed population into goal-coherent clusters, with agents below $\eta$ similarity to all centroids assigned to a residual category.
\end{definition}

The modal groupings serve three functions in the GFM framework.

\paragraph{Goal proposal.} When the GFM actor proposes new capabilities to expand $\vol(G)$, the modal groupings identify which proposals are likely to receive positive responses from which agents. A proposal aligned with $\mu_j$'s centroid is likely to be accepted by agents in $\mathcal{M}_j$ and ignored or rejected by agents in other groupings. The actor can target proposals to the groupings where they will produce the largest $\Deltahat(\pi)$ signal, improving the efficiency of the optimization loop (Section~\ref{subsec:optimization_loop}).

\paragraph{Diversity monitoring.} The number and spread of modal groupings provides a coarse measure of the goal diversity in the population. A population with many well-separated groupings has higher potential $\vol(G)$ than one with few tightly clustered groupings, because the union of dissimilar capability sets has greater measure. If the GFM actor observes modal groupings collapsing (agents converging on fewer, more similar goal profiles), this is a signal of $\vol(G)$ stagnation or impending contraction, even if no individual agent's capabilities have changed.

\paragraph{Scorpion contextualization.} The modal groupings provide context for interpreting scorpion behavior. An agent that contracts capabilities within its own modal grouping is likely a cornered animal or a defector from internal conflict. An agent that contracts capabilities across multiple groupings is likely pursuing an exogenous objective that crosses group boundaries. An agent whose behavior is inconsistent with \emph{all} modal groupings may be a blue-and-orange type (Section~\ref{app:blue_orange}) operating under axioms orthogonal to the entire observed population.

\paragraph{Updating groupings.} The modal structure is not static. As the GFM actor's world model improves and agents' goals evolve, groupings merge, split, and shift. The update procedure follows the same exponentially weighted scheme as the trust model: new observations are weighted more heavily than old ones (controlled by the decay parameter $\alpha$ from Equation~\ref{eq:cumulative_consistency}), so the grouping structure tracks the current population rather than reflecting historical patterns. The centroid $\mu_j$ is updated at each observation as an exponentially weighted moving average of the member agents' projected capability sets.

\paragraph{Relation to $\vol(G)$.} Modal groupings approximate the structure of the joint goal space $G$ at the population level. The union of all groupings' projected capability sets provides a coarse estimate of $\bigcup_k \Gind{k}$, and the gaps between groupings identify regions of $\G$ where no observed agent has capabilities, marking potential expansion targets. A GFM actor can use the grouping structure to identify underserved regions of capability space and prioritize actions that expand $\vol(G)$ into these gaps, rather than deepening capabilities in already-dense regions. This connects modal groupings to the diversity argument of Section~\ref{app:goal_similarity}: the groupings operationalize the intuition that frontier expansion is most efficient at the boundaries between known and unknown capabilities.
