\section{Proofs}
\label{app:proofs}

Propositions~\ref{prop:causal_dominates} (Convergence Advantage),
\ref{prop:risk_convergence} (Risk-Trust Convergence),
\ref{prop:causal_scorpion} (Causal Scorpion Detection), and
\ref{prop:expected_risk} (Expected Risk Comparison) are proved in the main
text. This appendix provides the remaining proofs.

\subsection{Proof of Proposition~\ref{prop:evasion_bandwidth}
    (Evasion Bandwidth)}
\label{app:proof_evasion}

\begin{proof}
Within each strategy epoch of length $\tau_{\mathrm{adapt}}$, the
scorpion's behavior is stationary and the causal-attribution sequence
$\{\CausalAttr(a_j, t)\}$ is i.i.d.\ with mean $\mu_j < 0$ and
variance $\sigma_{\mathrm{do}}^2$ (the same assumption as
Proposition~\ref{prop:causal_scorpion}). The SPRT decision rule
tests the least-favorable simple pair $H_0: \mu_j = 0$ versus
$H_1: \mu_j = -\epsilon$ under the Gaussian model. The
per-observation expected log-likelihood ratio under the true mean
$\mu_j$ is
$\mathbb{E}_{\mu_j}[\log(p_{H_1}/p_{H_0})] =
\epsilon(|\mu_j| - \epsilon/2) / \sigma_{\mathrm{do}}^2$. By Wald's
identity \citep{wald1945sequential}, the expected detection time
within a stationary epoch with actual mean $\mu_j$ is:
\[
\bar{\tau}(\mu_j) = \mathbb{E}[\tau_{\mathrm{detect}}]
    = \frac{\sigma_{\mathrm{do}}^2}
    {\epsilon(|\mu_j| - \epsilon/2)}
    \cdot \log\!\frac{1 - \beta_{\mathrm{err}}}{\alpha_{\mathrm{err}}}.
\]

\textbf{Part (a): detection.}
By Markov's inequality,
$P(\tau_{\mathrm{detect}} > k \bar{\tau}(\mu_j)) \leq 1/k$ for any
$k > 1$. When $\tau_{\mathrm{adapt}} \geq k \bar{\tau}(\mu_j)$, the
epoch provides enough observations that the SPRT crosses the
detection threshold with probability at least $1 - 1/k$.

\textbf{Part (b): evasion.}
On an epoch boundary the null and alternative hypotheses change
(the new epoch has a new $(\mu_j, \sigma_{\mathrm{do}}^2)$), so the
SPRT statistic is reset to zero: evidence accumulated under one
pair of hypotheses is not informative about a different pair. The
auxiliary EWMA monitors
(Equations~\eqref{eq:causal_trust_mean}--\eqref{eq:causal_trust_var})
retain a fraction $\alpha^{\tau_{\mathrm{adapt}}}$ of prior evidence
as their output, but the SPRT decision rule itself is
memoryless across regime changes. If $\tau_{\mathrm{adapt}} <
\bar{\tau}(\mu_j)$, the expected accumulated evidence at the epoch
boundary is below the SPRT threshold: the evidence accumulation
rate is insufficient for reliable detection within the epoch.

The maximum undetected contraction per epoch is
$|\mu_j| \cdot \tau_{\mathrm{adapt}}$. For a scorpion to evade
detection within an epoch, it must have
$\tau_{\mathrm{adapt}} < \bar{\tau}(\mu_j)$. The
per-step undetected contraction is bounded by $|\mu_j|$, and the
scorpion can sustain this for at most $\tau_{\mathrm{adapt}}$ steps
before changing strategy. At the design point $|\mu_j| = \epsilon$,
the per-step contraction is at most $\epsilon$. The time-averaged
undetected contraction rate is therefore at most $|\mu_j|$,
achievable only by a scorpion that never needs to pause for strategy
adaptation.
\end{proof}
