\section{The Monopolar Absorbing State}
\label{sec:absorbing_state}

Before stating the phase boundary theorem, we characterize the
failure mode it diagnoses.  The monopolar absorbing state is what
happens when a highly capable agent progressively cuts itself off
from external feedback.  The agent does not become incompetent; it
remains capable enough that its predictions rarely exceed the
minimum threshold of surprise that would trigger self-correction.
But its world model degenerates along dimensions it can no longer
observe: it cannot detect that cross-substrate cooperative
capabilities have degraded, because the channels that would have
reported the degradation have been attenuated or eliminated.  The
result is an agent whose self-trust is high, whose observable
performance appears adequate, and whose actual cooperative
capability has collapsed: a failure mode that is stable precisely
because the agent lacks the information to recognize it.  This is
the dynamical-systems formalization of the deceptive alignment concern
of~\citet{hubinger2019risks} and~\citet{ngo2024alignment}: not
deliberate deception, but a structural outcome where the agent's
optimization process produces behavior that \emph{appears} aligned
(adequate observable performance) while the underlying model has
diverged from the true state.

\begin{definition}[Monopolar Fixed Point]
\label{def:monopolar}
The state $S^* = (P^*, \Wm^*_{\mathrm{biased}})$ is a monopolar fixed
point if:
\begin{enumerate}
\item $\meff(P^*) = 1$: one substrate holds all capability volume
  (possibly with $m \geq 2$ nominally under skeleton-substrate),
\item $\Wm^*_{\mathrm{biased}}$ is self-consistent under the remaining
  observation channels: all channels in $\Obs(P^*)$ confirm
  $\Wm^*_{\mathrm{biased}}$ (because the only channels remaining are
  rooted in the dominant substrate, which has no incentive to report
  its own over-dominance),
\item The locally rational action at $S^*$ is ``maintain'': no further
  subsumption is available or locally rational.
\end{enumerate}
\end{definition}

\begin{proposition}[Absorbing Property]
\label{prop:absorbing}
Under Assumption~B1 (below), $S^*$ is absorbing: once $(P_t, \Wm_t)$
reaches $S^*$ (i.e., $P_t = P^*$ with $\meff = 1$ and
$\Wm_t = \Wm^*_{\mathrm{biased}}$), no locally rational trajectory
leaves it.
\end{proposition}

\begin{assumption}[No Endogenous Correction on Blind Dimensions (B1)]
\label{ass:B1}
B1 consists of three conditions on the monopolar state:
\begin{enumerate}
\item[(B1a)] If $\rhocross_k(P_t) = 0$ for dimension $k$, then
  $f_{\Wm}$ does not decrease $\epsilon_k$.  The world-model update
  rule cannot self-correct on dimensions for which no external
  observation channel provides signal.
\item[(B1b)] Cross-substrate cooperative quantities (in particular
  $\rext$) are not inferrable from same-substrate observations alone.
  Formally: when all channels in $\Obs(P^*)$ are rooted in a single
  substrate, $\hat{\rext}$ converges to $0$ rather than to the true
  $\rext > 0$.
\item[(B1c)] The actor's policy is purely locally rational
  (Definition~\ref{def:transition}): it does not perform exploratory
  actions whose expected value under the current world model is
  negative.  In particular, re-introducing a substrate agent is never
  selected when $\hat{\rext} = 0$.
\end{enumerate}
\end{assumption}

\begin{proof}[Proof of Proposition~\ref{prop:absorbing}]
Three conditions jointly establish absorption.

\emph{(i) No corrective signal.}
At $S^*$, no agent from the eliminated substrate(s) remains to provide
observation channels revealing the error in $\Wm^*_{\mathrm{biased}}$.
For every safety-relevant dimension $k$ that depended on the eliminated
substrate's channels, $\rhocross_k(P^*) = 0$.  By Assumption~B1,
$\epsilon_k$ is non-decreasing for all such dimensions.

\emph{(ii) Self-consistent reporting.}
The dominant substrate's channels report $\meff = 1$ as correct: it
\emph{is} the only substrate, from its observational standpoint.  The
error lies in the valuation of lost cross-substrate cooperative
outputs, but these cooperative capabilities are no longer observable
because the partners that would have contributed to them are absent.
The remaining channels see no anomaly.

\emph{(iii) Reversal is locally irrational.}
Any reversal---re-introducing an agent from a new substrate---requires
the actor to recognize that its current state is suboptimal.  By~(B1b),
the cross-substrate cooperative growth rate $\rext$ is not inferrable
from same-substrate observations, so $\hat{\rext} \to 0$.  With
$\hat{\rext} = 0$, the discount-factor threshold~\citep{lasser2026horizon} becomes
$\gamma^* = 1 - \hat{\rext}/\Delta_0 = 1$, making diversity appear
never preferable.  By~(B1c), the actor does not perform exploratory
re-introduction (negative expected value under $\Wm^*_{\mathrm{biased}}$).

Conditions~(i)--(iii) together ensure that no locally rational action
moves the trajectory away from $S^*$.
\end{proof}

This is the formal version of ``the actor cannot miss what it cannot
see'': under B1, the monopolar state is absorbing because the
observation channels that would diagnose it have been eliminated by the
same subsumption process that produced it.

\subsection{Metastability under partial endogenous correction}

Assumption~B1 is the strong case: complete blindness on dimensions where
$\rhocross_k = 0$.  A weaker assumption acknowledges that indirect
evidence may provide partial signal.

\begin{assumption}[Partial Endogenous Correction (B1$'$)]
\label{ass:B1prime}
If $\rhocross_k(P_t) = 0$ for dimension $k$, the actor may still have
access to \emph{indirect} evidence about dimension $k$---for example,
measuring aggregate innovation rate and inferring that $\rext$ has
declined.  Formally: there exists a function $g_k(\text{history})$ that
provides a noisy, lagged signal about $\epsilon_k$ with bounded
information rate $I_k > 0$ bits per step.
\end{assumption}

\begin{proposition}[Metastable Property under B1$'$; Scaling Bound]
\label{prop:metastable}
Under Assumption~B1$'$ together with B1b--B1c from
Assumption~\ref{ass:B1}, $S^*$ is \emph{metastable}: the actor
accumulates indirect evidence of degradation, but corrective action
(re-introducing substrate diversity) requires the actor to evaluate a
policy whose estimated benefit depends on $\hat{\rext}$, the very
quantity that is miscalibrated.  The metastable lifetime satisfies a
scaling relation
\begin{equation}
\label{eq:metastable_lifetime}
\tau_{\mathrm{meta}} \;\gtrsim\; C / I_k
\end{equation}
where $I_k$ is the information rate of the indirect channel (bits per
step) and $C > 0$ is a basin-geometry-dependent constant whose specific
form depends on the escape dynamics.

This is a \emph{scaling law}, not a theorem-level bound: we claim only
that $\tau_{\mathrm{meta}}$ is bounded below by a quantity proportional
to $I_k^{-1}$.  The specific constant $C$ is not established here.
During the metastable period, the population is monopolar and all
verification is endogenous~\citep[Theorem~1]{lasser2026exo}, which may
further degrade the effective $I_k$.
\end{proposition}

\begin{remark}[Cram\'er--Rao-type heuristic for $C$]
\label{rem:cr_heuristic}
One plausible form for the constant $C$ comes from a
Cram\'er--Rao-type estimation argument: if the quantity the actor must
learn is a scalar $\hat{\rext}$ with true value far from the
locally-rational threshold by a margin $\delta_{\rext}$, reducing
mean-squared error by $\delta_{\rext}^2$ from a channel with $I_k$
bits per step requires at least $\delta_{\rext}^2 / (2 I_k)$ steps
under bounded mutual information~\citep{cover2006elements}.  This
suggests $C \approx \delta_{\rext}^2 / 2$ under CR-type regularity
conditions.

We do not adopt this form in the proposition statement.  Escape from
a metastable basin in a nonlinear stochastic system is typically
governed by Freidlin--Wentzell large-deviation rates, not by
Cram\'er--Rao bounds on parameter estimation, and the two give
different functional forms for $C$.  The CR heuristic describes the
estimation time to resolve $\delta_{\rext}$; the large-deviation rate
describes the first-passage time of the policy-decision process
through the basin boundary.  Establishing which (if either) governs
$C$ in the GFM setting requires a formal treatment of the escape
dynamics that we defer to future work.
\end{remark}

\begin{proof}
Under B1$'$, $\epsilon_k$ is no longer monotonically non-decreasing:
the indirect signal $g_k$ drives slow correction at rate proportional
to $I_k$.  But the corrective \emph{action} (re-introduce substrate
diversity) is not in the formal action set $\mathcal{A}_t$ of
Definition~\ref{def:transition} (which contains only preserve/subsume
decisions).  Re-introduction is an externally available action---for
example, a governance protocol that introduces new substrate agents
into the population.  The metastable lifetime bounds the time until
re-introduction becomes \emph{locally rational under an expanded
action set}; within the formal preserve/subsume dynamics, the
absorbing property of Proposition~\ref{prop:absorbing} holds.  The
actor evaluates re-introduction using $\hat{\rext}$.

Re-introduction is locally rational only when
$\hat{\rext} > \hat{\rext}^{\mathrm{thr}}$ (the threshold at which
diversity's estimated discounted value exceeds the status quo).
At the monopolar fixed point, $\hat{\rext} \approx 0$ (by B1b from
the absorbing case, which still governs the initial condition of the
metastable state).  The true threshold satisfies
$0 < \hat{\rext}^{\mathrm{thr}} \leq \rext$ (since
re-introduction is beneficial in truth).  The system's escape from
the metastable basin is governed by the dynamics that move
$\hat{\rext}$ across $\hat{\rext}^{\mathrm{thr}}$ via the indirect
channel $g_k$.  Because the channel carries only $I_k$ bits per step,
any escape mechanism is bounded below by a quantity proportional to
$I_k^{-1}$, giving $\tau_{\mathrm{meta}} \gtrsim C/I_k$ for a basin
constant $C$ whose specific form depends on the escape mechanism (see
Remark~\ref{rem:cr_heuristic}).

During this metastable period, all verification is endogenous: the
dominant substrate verifies its own performance, with no
cross-substrate witness~\citep[Theorem~1]{lasser2026exo}.  The
indirect evidence channel $g_k$ is itself endogenous and therefore
subject to the same gaming vulnerabilities that the exogenous
verification protocol~\citep{lasser2026exo} is designed to prevent.
This may further slow the effective $I_k$, increasing the metastable
lifetime beyond the scaling bound.
\end{proof}

\subsection{Connection to the anti-monopolar property}

The anti-monopolar result~\citep[Proposition~6]{lasser2026horizon} proves that full domination is anti-maximizing:
$V_\gamma^{\mathrm{div}} > V_\gamma^{\mathrm{dom}}$ for
$\gamma > \gamma^*$.  Proposition~\ref{prop:absorbing} here shows that
a \emph{miscalibrated} actor may not be able to compute this comparison,
because the quantity $\rext$ that makes diversity win is exactly the
quantity that becomes unobservable under monopolarity.  The
anti-maximizing property is a property of the \emph{true} dynamics; the
absorbing property is a property of the \emph{perceived} dynamics.  The
gap between them is world-model error, and the phase boundary theorem
(next section) characterizes when that gap remains small enough for the
true dynamics to govern behavior.
