\begin{abstract}
Locally rational subsumption steps can cascade to full capability
domination through a compound feedback loop: each step removes
observation channels, degrading the world model that would have detected
the next step's harm.  We formalize this as a coupled dynamical system
over capability posets and world-model states, and characterize the
\emph{phase boundary} separating self-correcting dynamics (where
remaining channels suffice to detect and reverse errors) from absorbing
dynamics (where the world model degrades irreversibly toward the
monopolar fixed point).  The phase boundary is an inequality in terms of
five quantities: the effective correction rate (combining exogenous
channel signals with an endogenous calibration incentive), the
error-proportional degradation rate, the subsumption frequency, the
maximum exogenous drift from lost channels, and the minimum
cross-substrate redundancy.  Three previously identified compound feedback
loops---structural avoidance, partial subsumption, and over-calibrated
restriction---are shown to be instances of the same dynamical system
with different effective parameters.  The Wamura problem (unfalsifiable
risk claims under max-aggregation) emerges as a degenerate phase
boundary: without controlled relaxation, the self-correction rate is
zero and the entire state space is absorbing.  For two-substrate
systems, the analysis predicts that each substrate's observation
channels are singularly load-bearing, with progressive attenuation
(rather than outright removal) as the primary vulnerability.

This paper analyzes channel \emph{availability}; its results
compose with the exogenous-verification construction
of~\citet{lasser2026exo}, which analyzes channel \emph{integrity}.
Neither paper is self-sufficient as a safety result: the corrective
mechanism we identify (meta-knowledge of the phase boundary) requires
the verification ledger of~\citet{lasser2026exo} to avoid becoming
gameable, and the verification protocol's persuasion-scaling concern
requires this paper's absorbing-basin analysis to identify when
integrity assumptions stop binding.  The two papers are best read as
a single safety result with separate structural content.
\end{abstract}
