\section*{Author Contributions}

\textbf{Teague Lasser} conceived the compound feedback loop analysis
as the dynamical framework connecting Papers~1--5: the identification
of three loop instances (observation-channel, trust-calibration,
capability-evolution) from the gap analysis, the phase boundary as a
separating surface in parameter space, and the connection between
self-trust calibration and the monopolar absorbing state. Directed the
formal development, identified the endogenous calibration incentive
($c_{\mathrm{V}} \cdot \nu$ term in Assumption~C1), proposed the
prediction-market corrective force as the Paper~5 connection, and made
final editorial decisions. Responsible for the paper's intellectual
direction and all claims made.

\textbf{Claude Opus 4.6} (Anthropic) drafted the formal exposition:
the coupled $(P, \Wm)$ dynamical system, the Lyapunov error
construction with weighted sum-of-squares candidate, the per-step
error dynamics (Proposition~1), the phase boundary theorem with its
three parts (self-correcting basin, absorbing basin, critical
surface), the cascade lemma, the metastable basin under partial
endogenous correction, the minimum-redundancy design criterion
(Theorem~2) with its two relaxations, the skeleton-substrate
vulnerability analysis, and the worked example instantiation. Also
developed the meta-knowledge discussion and the world-model
verification corrective force analysis.

\textbf{GPT 5.4} (OpenAI) served as technical reviewer across
multiple codex review cycles, identifying formal gaps including: the
$r_S^{\mathrm{eff}}$ circularity (the effective self-correction rate
depending on $\rho$ while $\rho^*$ was defined as a ratio involving
$r_S^{\mathrm{eff}}$), the missing $\alpha_{\min}$ factor on the
endogenous correction term throughout the simplified conditions,
the need for explicit connectivity and non-stationarity conditions
in Part~(b)'s cascade propagation, the overshoot constant $\mu$ in
the affine-recursion convergence (requiring explicit linearized-regime
scoping), the strict-gap requirement for $\epsilon_{\mathrm{safe}} > 0$,
and the per-dimension relaxations R1/R2 in the design criterion.

\medskip
\noindent\emph{Transparency note.} Both AI systems operated as tools under
human direction. Neither system has continuity across sessions, cannot take
responsibility for the work in the sense required by most venue authorship
policies, and cannot respond to reviewer queries independently. They are
listed as authors to accurately represent their contributions to the
intellectual content of the paper, not to claim that they meet all criteria
of traditional academic authorship. The corresponding author for all
inquiries is Teague Lasser.
