\section{Design Criterion: Minimum Channel Redundancy}
\label{sec:design_criterion}

The phase boundary theorem converts a dynamical-systems result into a
design criterion: what is the minimum cross-substrate redundancy per
safety-relevant dimension needed to guarantee the system stays in the
self-correcting basin?

\subsection{The redundancy criterion}

\begin{theorem}[Minimum Redundancy for Self-Correcting Dynamics]
\label{thm:redundancy}
The following per-dimension criterion is a necessary condition
derived from the aggregate simplified sufficient condition
(Equation~\ref{eq:self_correcting}); it is stated per-dimension for
design guidance but inherits two relaxations from the derivation
(see proof and qualifications below):
\begin{equation}
\label{eq:kmin}
\forall\; k \in \{1, \ldots, K\}: \quad
\rhocross_k(P) \geq \Kmin(k, \Lyap_{\mathrm{avg}})
\end{equation}
where
\begin{equation}
\label{eq:kmin_formula}
\Kmin(k, \Lyap_{\mathrm{avg}}) =
  \frac{\rW \cdot \rsub^{\max}
    + \dmax \cdot \Delta\rho_{\mathrm{avg}}(k) /
    \Lyap_{\mathrm{avg}}
    - \alphamin \cdot c_{\mathrm{V}} \cdot \underline{\nu}}
  {\rS}
\end{equation}
$\rsub^{\max}$ is the maximum subsumption rate under
locally rational dynamics over the relevant trajectory window, and $\Delta\rho_{\mathrm{avg}}(k)$ is
the time-averaged redundancy loss on dimension $k$ per subsumption
step.  The exogenous rates $\rS$, $\rW$, $\dmax$ are the global
bounds from Theorem~\ref{thm:phase_boundary}; the endogenous
correction $\alphamin \cdot c_{\mathrm{V}} \cdot \underline{\nu}$
appears as a
subtraction in the numerator (it reduces the channel redundancy
needed).  When $\alphamin \cdot c_{\mathrm{V}} \cdot \underline{\nu}$
alone exceeds
the degradation terms, $\Kmin \leq 0$ and endogenous correction
suffices; setting $c_{\mathrm{V}} = 0$ recovers the purely exogenous
formula.  The per-dimension dependence enters
only through $\Delta\rho_{\mathrm{avg}}(k)$, which varies across
dimensions because different dimensions lose different amounts of
channel coverage per subsumption event.  Note that $\Kmin$ is
real-valued, since $\rhocross_k$ is defined as a sensitivity-weighted
sum (Definition~\ref{def:cross_redundancy}), not an integer count.

\emph{Qualification:} even at initialization, the per-dimension
$\Kmin$ may underestimate the true required redundancy due to
relaxations R1 and R2 (see proof).  Along trajectories, the
redundancy $\rhocross_k(P_t)$ may further decrease under the dynamics
being analyzed (Corollary~\ref{cor:redundancy_monotone}).  A stronger
result would bound the worst-case $\rhocross_k$ along all trajectories
starting from the basin, which requires coupling the redundancy
dynamics to the phase boundary analysis (see proof for discussion).
The aggregate condition (Equation~\ref{eq:self_correcting}) does not
have these per-dimension relaxations and is the primary guarantee;
the per-dimension $\Kmin$ serves as design guidance for allocating
redundancy across dimensions.
\end{theorem}

\begin{proof}
The aggregate correction signal from C1 is
$S(t) \geq (c_{\mathrm{S}} \cdot \rhomincross + c_{\mathrm{V}}
\cdot \nu) \cdot \Lyap$.  For the per-dimension criterion, we
attribute the exogenous correction proportionally: dimension $k$
receives correction at rate $\rS \cdot \rhocross_k(P_t)$ from its
covering channels.  The endogenous term $\alphamin \cdot
c_{\mathrm{V}} \cdot \underline{\nu}$ is aggregate (it arises from
the actor's overall $\volL$-gradient, not from any single
dimension's channels).  We assign the full aggregate endogenous
correction to each dimension's budget.  This makes $\Kmin$ a
\emph{lower bound} on the true per-dimension requirement: the
endogenous correction is a shared resource whose per-dimension
allocation is bounded by the aggregate total $\alphamin \cdot
c_{\mathrm{V}} \cdot \underline{\nu}$, so crediting each dimension
with the full amount may underestimate $\Kmin$.  For a conservative
per-dimension bound, set $c_{\mathrm{V}} = 0$ in
Equation~\ref{eq:kmin_formula}; the aggregate condition
(Equation~\ref{eq:self_correcting}) does not have this decomposition
issue.
Each dimension has its own drift exposure
$\dmax \cdot \Delta\rho_k(t)$, where $\Delta\rho_k$ is the
redundancy loss on dimension $k$ specifically.  Requiring each
dimension individually to satisfy the self-correcting condition
yields the per-dimension criterion: $\rS \cdot \rhocross_k +
\alphamin \cdot c_{\mathrm{V}} \cdot \underline{\nu} >
\rW \cdot \rsub + \dmax \cdot \Delta\rho_{\mathrm{avg}}(k) /
\Lyap_{\mathrm{avg}}$.  Solving for $\rhocross_k$ yields
Equation~\ref{eq:kmin_formula}.  Two relaxations enter:
\begin{enumerate}
\item[(R1)] The denominator uses $\Lyap_{\mathrm{avg}}$ (the
  aggregate time-averaged error across all dimensions), not a
  per-dimension error.  If dimension~$k$ is well-calibrated but
  other dimensions carry large error, $\Lyap_{\mathrm{avg}}$ is
  inflated, making the drift term $\dmax \cdot
  \Delta\rho_{\mathrm{avg}}(k) / \Lyap_{\mathrm{avg}}$ artificially
  small and $\Kmin(k)$ artificially low.  For conservative use,
  substitute $\Lyap_{\mathrm{avg}} = w_k \cdot
  \epsilon_{k,\mathrm{avg}}^2$ (the per-dimension contribution).
\item[(R2)] The endogenous term is credited in full to each dimension
  (see caveat above).  For conservative use, set $c_{\mathrm{V}} = 0$.
\end{enumerate}
Note that $\Kmin$ increases as
$\Lyap_{\mathrm{avg}}$ decreases: well-calibrated dimensions require
\emph{more} redundancy to protect against channel removal, because
drift exposure dominates the error-proportional term when current
error is small.  (The global condition
Equation~\ref{eq:self_correcting} is recovered by taking
$\rhomincross = \min_k \rhocross_k$ and
$\Delta\rho_{\mathrm{avg}} = \max_k \Delta\rho_{\mathrm{avg}}(k)$;
the aggregate condition does not have relaxations R1 or R2.)

The $\Kmin$ formula has two regimes: when $\Lyap_{\mathrm{avg}}$ is
large relative to $\Delta\rho_{\mathrm{avg}}(k)$, the
error-proportional term dominates and $\Kmin$ is moderate; when
$\Lyap_{\mathrm{avg}}$ is small, the drift-exposure term
$\dmax / \Lyap_{\mathrm{avg}}$ dominates and $\Kmin$ grows.  The
second regime captures the ``well-calibrated channel removal'' failure
mode: $\Kmin$ can be substantially larger than the formula without the
drift term would predict.

The design criterion requires knowing $\Lyap_{\mathrm{avg}}$, a
trajectory-dependent quantity.  In practice one uses a representative
operating-regime estimate; for a system with error bounded by
$L_{\max}$, substituting $\Lyap_{\mathrm{avg}} = L_{\max}$ gives a
\emph{lower bound} on $\Kmin$ (anti-conservative; the true $\Kmin$
may be larger if the system operates at lower error).  The criterion
is therefore a design guideline for the expected operating regime, not
a worst-case safety guarantee.  A closed guarantee would require
either a lower bound on $\Lyap_{\mathrm{avg}}$ (bounding the system
away from perfect calibration) or coupling the redundancy dynamics to
the phase boundary analysis.
\end{proof}

\subsection{Practical implications}

\begin{corollary}[Two-Substrate Minimum]
\label{cor:two_substrate}
For a two-substrate partition ($m = 2$, e.g., biology $+$ silicon):
\begin{enumerate}
\item $\rhocross_k \leq 2$ for each dimension $k$ (at most one channel
  from each substrate).
\item The full criterion (Equation~\ref{eq:kmin_formula}) with
  $\rhocross_k = 2$ requires $2 \rS + \alphamin \cdot c_{\mathrm{V}}
  \cdot \underline{\nu} \geq \rW \cdot \rsub^{\max} + \dmax \cdot
  \Delta\rho_{\mathrm{avg}}(k) / \Lyap_{\mathrm{avg}}$ for every
  dimension $k$.
\item Dimensions where one substrate provides the only cross-substrate
  signal have $\rhocross_k = 1$, requiring
  $\rS + \alphamin \cdot c_{\mathrm{V}} \cdot \underline{\nu} \geq
  \rW \cdot \rsub^{\max} + \dmax \cdot
  \Delta\rho_{\mathrm{avg}}(k) / \Lyap_{\mathrm{avg}}$: exogenous
  correction plus endogenous incentive must together exceed
  degradation and drift exposure on that dimension alone.
\end{enumerate}
\end{corollary}

This means that in a two-substrate world, each substrate's observation
channels are singularly load-bearing for every dimension they cover.
Loss of one substrate's channel on a dimension reduces $\rhocross_k$
from at most $2$ to at most $1$; the redundancy criterion is
immediately violated on any dimension where
$\Kmin(k, \Lyap_{\mathrm{avg}}) > 1$, and provides zero safety margin
on dimensions where $\Kmin(k, \Lyap_{\mathrm{avg}}) \leq 1$.

\begin{remark}[Progressive attenuation as the primary threat]
\label{rem:attenuation}
The $m = 2$ vulnerability is not that one substrate is \emph{removed}
(an overt act that would itself be detectable) but that it is
\emph{attenuated}: the human stops reviewing AI capability claims
because the AI's self-reports ``seem fine,'' or human judgment is
progressively replaced by automated proxies that share the silicon
substrate's failure modes.

This attenuation reduces effective sensitivity without changing the
nominal channel count, causing $\rhocross_k$ to drop below the
threshold while appearing unchanged to a naive audit.  The
skeleton-substrate scenario (Corollary~\ref{cor:skeleton}) is
precisely this failure mode.

The framing should be read as emphasizing $m = 2$ fragility and the
value of $m \geq 3$, not the specialness of any particular substrate.
\end{remark}

\begin{corollary}[Skeleton-Substrate Vulnerability]
\label{cor:skeleton}
Under a skeleton-substrate strategy where $m \geq 2$ nominally but the
non-dominant substrate's channels have sensitivity $\to 0$:
\begin{equation}
\label{eq:skeleton}
\rhocross_k(P) \to \rhocross_{k,\mathrm{dominant}}(P)
\end{equation}
because the sensitivity-weighted definitions
(Definitions~\ref{def:redundancy} and~\ref{def:cross_redundancy})
already discount low-sensitivity channels.
\end{corollary}

\begin{proof}
From Definition~\ref{def:cross_redundancy}, the contribution of
substrate $s$ to $\rhocross_k$ is
$\min(1, \max_{c \in s} \sigma(c,k) / \sigma_{\mathrm{thr}})$.
As $\sigma(c,k) \to 0$ for all channels on the non-dominant substrate,
this contribution goes to zero.  The sum reduces to the dominant
substrate's contribution alone.
\end{proof}

The skeleton-substrate strategy does not evade the redundancy criterion
by construction: it directly reduces $\rhocross_k$ by reducing the
marginalized substrate's contribution to near-zero.  No separate
``effective'' redundancy is needed; the phase boundary formalism
already captures this vulnerability through the sensitivity weighting.

\subsection{Self-diagnosability}

The phase boundary is partially self-diagnosable by the actor.  The
actor can observe:
\begin{itemize}
\item $\rhocross_k$ (it knows which channels it has and their
  approximate sensitivities),
\item $\rS$ (from its own learning rate and historical correction
  behavior).
\end{itemize}

The actor cannot directly observe:
\begin{itemize}
\item $\rW$ (the degradation rate depends on the true dynamics, not the
  perceived dynamics),
\item $\dmax$ (the exogenous drift rate on blinded dimensions is by
  definition unobservable once the dimension is blind).
\end{itemize}

This means the actor can check ``am I above the threshold for the $\rW$
I estimate?'' but not ``is my estimate of $\rW$ correct?''  The actor
can monitor the observable ingredients and flag when they approach the
boundary, even if it cannot guarantee the boundary is computed
correctly.  This is an honest limitation: the phase boundary is a
sufficient condition for self-correction, not a decision procedure for
whether self-correction will actually obtain.
