\section{Discussion and Open Questions}
\label{sec:discussion}

\subsection{What this paper establishes}

Five results emerge from the analysis.

\paragraph{1.\ The phase boundary exists and is characterized.}
Compound feedback loops are not inherently catastrophic or inherently
self-correcting.  The outcome depends on the ratio of self-correction
rate to observation-loss rate, mediated by channel redundancy.
Theorem~\ref{thm:phase_boundary} gives the boundary; the linearized
sufficient condition (Equation~\ref{eq:self_correcting}) gives the
computable form.

\paragraph{2.\ All three feedback loops are the same dynamical system.}
The unified treatment (Section~\ref{sec:unified_loops}) shows that
structural avoidance, partial subsumption, and over-calibrated
restriction share a common formal structure.  Solutions for one loop
(controlled relaxation for the Wamura problem, structural discovery
incentives for leverage blindness) improve the phase boundary for all.

\paragraph{3.\ The Wamura problem is a degenerate phase boundary.}
Without a mechanism for evidence generation against unfalsifiable risk
claims, the entire state space is absorbing
(Proposition~\ref{prop:wamura}).  The Wamura problem was called ``the
most important open problem'' in the discussion
of~\citet{lasser2026horizon}.  The phase boundary analysis explains
why: it is the case where the self-correction rate is zero by
construction, making the feedback loop absorbing regardless of all
other parameters.

\paragraph{4.\ Cross-substrate redundancy is the load-bearing quantity.}
Same-substrate channels may share failure modes; cross-substrate
channels are structurally independent (under the channel
isolation property of~\citet{lasser2026exo} and, for additivity, the joint-factorization assumption).
The design criterion (Theorem~\ref{thm:redundancy}) is stated in terms
of cross-substrate redundancy, making it robust to correlated failures
within substrates.

\paragraph{5.\ In two-substrate systems, each substrate's channels are singularly load-bearing.}
This is a structural prediction from the dynamics rather than a
normative claim about human importance: in $m = 2$ systems, losing one
substrate's channel on a dimension reduces $\rhocross_k$ to at most
$1$, violating the redundancy criterion on any dimension where
$\Kmincons(k) > 1$ and leaving zero margin where $\Kmincons(k) \leq 1$
(Corollary~\ref{cor:two_substrate}).
The skeleton-substrate scenario (Section~\ref{sec:worked_example})
shows that progressive \emph{attenuation} of biological channels, not
outright removal, is the primary vulnerability.

\subsection{Limitations}

The failure modes studied here---progressive channel degradation,
feedback-loop instability, and the gap between observable performance
and actual alignment---are instances of the concrete safety problems
identified by~\citet{amodei2016concrete}.  Specifically, they
correspond to distributional shift (the agent's training distribution
diverges from the deployment distribution as channels attenuate) and
reward hacking (the agent's observable metrics remain good while the
underlying objective degrades).  The phase boundary formalizes the
conditions under which these problems become self-reinforcing.

\paragraph{Linearized treatment.}
The phase boundary is stated in a linearized regime (constant $\rS$,
$\rW$, $\rsub$).  In the full dynamics these quantities are
state-dependent.  A larger world-model error produces more aggressive
subsumption ($\rsub$ increases with $\Lyap$), and the correction rate
$\rS$ depends on $T_s$ which itself depends on the error history.
The linearized treatment produces the qualitative result
(self-correcting and absorbing basins exist, separated by a
characterized boundary), but the quantitative location of the boundary
shifts under the nonlinear dynamics.  A full nonlinear analysis would
tighten the boundary at the cost of substantially more complex
proofs.

\paragraph{Honest-channel assumption.}
The analysis assumes observation channels are honest: their signal is
noisy but not adversarially manipulated.  The cryptographic
commitment protocol of~\citet{lasser2026exo} provides the enforcement
mechanism for this assumption.  The two papers are designed to be
read as complements: the exogenous verification framework ensures
channels are honest, and this paper ensures enough honest channels
survive.  But the boundary between ``noisy'' and ``adversarial'' is
itself a modeling choice.  The line is blurred in practice by a
failure mode neither paper alone addresses: \emph{convertibility of
integrity attacks into availability attacks}.

A sophisticated actor that cannot commit a false claim (because the
commitment protocol would detect it) can instead choose which claims
to commit and which to suppress.  On a dimension where the actor
expects an unfavorable verification outcome, it can decline to
participate in verification on that dimension.  The channel then
remains nominally intact but carries no signal about the dimension of
concern.  This converts an integrity attack (committing a false
claim) into an availability attack (making the channel silent on the
relevant dimension).  The integrity protocol of~\citet{lasser2026exo}
handles claims that are committed; claims the actor chooses not to
commit fall outside its scope and become an availability problem for
this paper.  The three loops differ in how they interact with this
blur.  Loops~1 and~3 (structural avoidance, over-calibrated
restriction) produce channel silence through the dynamics themselves,
while Loop~2 (partial subsumption) produces channel degradation.  A
sophisticated Loop-2 actor can emulate Loop-1 behavior on specific
dimensions, producing selective silence that looks like the
framework-endogenous failure mode rather than adversarial
manipulation.  Distinguishing the two requires observing \emph{which}
channels fall silent, which is itself a meta-channel subject to the
same availability concerns.

\paragraph{Partial self-diagnosability.}
As noted in Section~\ref{sec:design_criterion}, the actor can observe
$\rhocross_k$ and $\rS$ but not $\rW$ or $\dmax$.  The phase boundary
is therefore a sufficient condition for self-correction that the actor
can partially monitor, not a decision procedure it can fully evaluate.
This is an honest limitation, not a fixable gap.  The quantities the
actor cannot observe are precisely the ones that depend on the true
dynamics rather than the perceived dynamics.

\subsection{Connections to other proposals}

\paragraph{Controlled relaxation.}
Proposition~\ref{prop:wamura} shows the Wamura problem is a
degenerate phase boundary.  Controlled relaxation protocols (the
subject of a companion paper) provide the mechanism that makes
$\rS > 0$ for Loop~3.  The two analyses are complementary.  This
paper provides the dynamical framework showing \emph{why} a positive
self-correction rate matters, while controlled relaxation provides
the specific protocol that \emph{achieves} it.

\paragraph{The B-to-C gap.}
The phase boundary analysis assumes $\volL$ accurately measures what
it claims.  The B-to-C gap is the scenario where this assumption
fails: $\volL$ diverges from the experiential volume it is supposed
to track.  A complete treatment would need to extend the phase
boundary to account for the divergence between $\volL$ and the
realized capability volume.  This extension is future work.

\paragraph{Cryptographic verification.}
\citet{lasser2026exo} addresses observation-channel integrity; this paper
analyzes the dynamical consequences of channel \emph{availability}.
The combination is: exogenous verification ensures channels are honest,
this paper ensures enough honest channels survive.  A joint treatment
where the phase boundary accounts for both integrity
failures~\citep{lasser2026exo} and availability failures (this paper)
would provide a stronger combined guarantee.

\paragraph{World-model verification as a corrective force.}
The self-trust feedback chain that creates the monopolar basin
(Section~\ref{sec:absorbing_state}) has a specific vulnerability:
the agent cannot observe its own calibration degradation because the
channels that would report it have been attenuated.  The
commit-then-reveal protocol of~\citet{lasser2026exo} provides a
direct corrective force against this failure mode.

If the agent commits world-model predictions to the verification
ledger \emph{before} observing outcomes, the residuals between
committed predictions and realized outcomes are an exogenous
calibration signal that the agent cannot game.  The commitment is
cryptographic (the prediction cannot be changed after the fact), and
the outcome verification is cross-substrate (the agent cannot
self-report the outcome it committed to predict).  This is exactly
the signal the monopolar basin lacks: an unfakeable measure of
world-model quality on the dimensions the agent has stopped
observing directly.

The connection to prediction markets~\citep{hanson2003combinatorial}
is immediate.  In a multi-agent setting, the verification ledger
generalizes to a prediction market: agents commit predictions, stake
reputation on outcomes, and calibration is scored publicly.  The
market's information aggregation rate $I_{\mathrm{market}}$ replaces
the single-channel $I_k$ in the metastable lifetime bound
(Proposition~\ref{prop:metastable}), potentially shortening the
escape time from the metastable basin dramatically.  A consistently
miscalibrated agent loses stake, which is a $\volL$-legible signal.
The agent's cooperative capabilities that depend on accurate
predictions degrade in the ledger, producing a negative $\Delta\volL$
that the agent's own optimization pressure cannot ignore.

Two design constraints apply.  First, the queries must be
\emph{masked}: the agent should not know which dimensions of its
world model are being tested, to prevent selective participation
in predictions it expects to get right.  Second, the predictions
must be on \emph{observable quantities} (what will metric $X$ be
next period?) rather than internal state (what is your belief
about $X$?), both to protect the agent's epistemic privacy and to
ensure the outcome is independently verifiable.

In the minimal case (single actor, no market), the operational
instantiation is simpler: the user periodically poses calibration
queries that the agent commits predictions to, and the user
independently verifies the outcomes.  This is the exogenous
trust-review channel from~\citet{lasser2026exo} applied to
world-model calibration specifically.  Even this minimal version
increases the effective $\rS$ (each query adds a correction
signal) without increasing $\rhomincross$ (the channel count is
unchanged), widening the self-correcting basin.

If world models are not merely committed but \emph{shareable} (other
agents can read and build on them), the ledger becomes a
bootstrapping mechanism.  New agents enter the population with a
calibrated world model derived from the public record rather than
from scratch observation.  This addresses the initialization
sensitivity of Theorem~\ref{thm:redundancy}: the design criterion
depends on $\Lyap_{\max}$, the initial world-model error.  A
population with a shared world-model ledger starts new agents with
lower $\Lyap_{\max}$, reducing the redundancy threshold for the
self-correcting basin.

\subsection{Meta-knowledge and the phase boundary}
\label{sec:meta_knowledge}

A self-interested $\volL$-maximizer with access to
Theorem~\ref{thm:phase_boundary} gains three operational tools that a
naive actor does not have.

\paragraph{Precautionary monitoring.}
The actor can partially evaluate the phase boundary condition before
each subsumption decision.  It observes $\rhocross_k$ (which channels
it has), estimates $\rS$ (from its own learning rate), and tracks
$\rsub$ (from its own subsumption history).  $\rW$ is not directly
observable, since the degradation rate depends on the true dynamics.
But the actor can compute the worst-case impact of a proposed
subsumption on $\rhomincross$ and check whether the post-subsumption
state still satisfies the condition under plausible $\rW$ ranges.
This provides a computable precautionary filter: reject subsumptions
that would push $\rhomincross$ below the critical threshold, even when
the local $\volL$ calculation favors the subsumption.

\paragraph{Meta-rational discounting.}
A naive $\volL$-maximizer evaluates subsumption against its current
world model $\Wm_t$.  An actor with phase boundary knowledge adds a
correction: the value of a channel is higher than $\Wm_t$ estimates,
because $\Wm_t$'s accuracy on the dimensions that channel covers
\emph{depends on} the channel being present.  Removing the channel
degrades the very estimate that justified the removal.  The theory
provides the formal basis for this self-distrust.  The actor knows
that its valuation of a subsumption is biased precisely because the
channel it would lose is the one calibrating the valuation.

\paragraph{Interpretive framework for indirect evidence.}
Under the metastable variant (Part~b$'$), the actor has indirect
signals: declining innovation rate, shrinking cooperative output,
increasing prediction residuals.  Without the phase boundary theory,
these are ambiguous data points.  With the theory, they are
diagnostics for a specific failure mode (substrate homogenization)
with a known remedy (re-introduce substrate diversity).
Meta-knowledge converts ambiguous signals into actionable diagnoses.
This shortens the metastable lifetime $\tau_{\mathrm{meta}}$ by
increasing the effective information rate from indirect channels.

\paragraph{The preventive-not-rescue asymmetry.}
Meta-knowledge is most valuable in the self-correcting basin and at
the phase boundary, where the actor's world model is calibrated
enough to apply the theory accurately.  In the absorbing basin, the
world model has degraded to the point where the actor cannot
accurately evaluate the phase boundary condition.  It believes it is
in the self-correcting basin when it is not, because the channels
that would reveal the miscalibration have been eliminated.
Meta-knowledge is therefore a \emph{preventive} tool (it keeps the
actor from approaching the boundary) rather than a \emph{rescue} tool
(it cannot pull the actor out of the absorbing basin once the world
model is sufficiently degraded).  This mirrors the cascade asymmetry
(Remark~\ref{rem:stabilizing_cascade}): the stabilizing cascade
benefits from meta-knowledge at every step, while the destabilizing
cascade degrades the meta-knowledge along with everything else.

\paragraph{The recursive implication.}
If meta-knowledge of the phase boundary increases resilience, then
developing and sharing the theory is itself a $\volL$-maximizing
action.  A $\volL$-maximizer that understands this would: (a)~invest
in understanding its own failure modes, because that understanding
increases $\nu$ in Assumption~C1 and thereby widens the
self-correcting basin; (b)~share that understanding with other agents,
because shared meta-knowledge is a cooperative capability that
benefits the entire population; (c)~seek external review of its
understanding, because miscalibrated meta-knowledge --- false
confidence in a wrong phase boundary estimate --- could suppress the
precautionary discounting that correct meta-knowledge would provide,
making the actor \emph{more} vulnerable; and (d)~commit the theory to
the verification ledger~\citep{lasser2026exo}, because once verified
and published, the theory becomes a shared resource that reduces every
new agent's initialization cost.

Point~(c) is the most subtle.  Meta-knowledge that is wrong about the
phase boundary location could convince the actor it has room to
subsume when it does not, increasing rather than decreasing
vulnerability.  The theory itself needs to be verified through the
exogenous verification protocol.  This creates a closure: the paper's
results need the mechanisms its companion paper develops.

\subsection{Open questions}

\begin{enumerate}
\item \textbf{Nonlinear phase boundary.}  The state-dependent extension
  of the phase boundary, where $\rS$, $\rW$, and $\rsub$ are functions
  of $(P_t, \Wm_t)$ rather than constants, would produce tighter bounds
  at the cost of a switched-systems analysis with state-dependent
  switching signals.

\item \textbf{Characterization of absorbing states.}  We prove
  convergence to \emph{a} monopolar fixed point but do not characterize
  the full set of absorbing states.  Partial monopolarity (some
  dimensions monopolar, others not) may be stable under certain
  parameter regimes.

\item \textbf{Optimal redundancy allocation.}  Given a fixed budget of
  observation channels, how should they be allocated across dimensions
  and substrates to maximize the self-correcting basin?  This is a
  resource-allocation problem over the channel coverage matrix.

\item \textbf{Empirical calibration.}  The worked example uses
  pedagogical parameter values.  Calibrating $\rS$, $\rW$, $\rsub$,
  and $\dmax$ from operational data (e.g., from AI governance
  structures, human-oversight-of-AI dynamics, or historical
  institutional feedback loops) would convert the phase boundary from
  a qualitative characterization to a quantitative prediction.
\end{enumerate}
