\section{World-Model Quality and the Lyapunov Function}
\label{sec:lyapunov}

The phase boundary requires a scalar measure of world-model quality
whose dynamics can be bounded.  We define this as a weighted
sum-of-squares error and establish the per-step bounds that drive the
rest of the analysis.

\subsection{Measuring world-model quality}

\begin{definition}[World-Model Error]
\label{def:wm_error}
The error of $\Wm_t$ on dimension $k$ is
\begin{equation}
\label{eq:dim_error}
\epsilon_k(t) = |\Wm_t[k] - \Wm^*[k]|
\end{equation}
where $\Wm^*[k]$ is the true value of dimension $k$ (the true leverage
of capability $d$, the true risk of tuple $S$, the true effective
substrate diversity, etc.).  The total error is:
\begin{equation}
\label{eq:lyapunov}
\Lyap(\Wm_t) = \sum_{k=1}^{K} w_k \, \epsilon_k(t)^2
\end{equation}
where $w_k > 0$ is the safety-relevance weight of dimension $k$
(higher for dimensions whose miscalibration produces larger $\volL$
errors).  $\Lyap$ serves as the Lyapunov candidate.
\end{definition}

\begin{remark}[Weight choice]
\label{rem:weights}
Two natural choices for $w_k$ present themselves.  The first is
$\volL$-sensitivity: $w_k = |\partial \volL / \partial \Wm[k]|$,
making $\Lyap$ measure $\volL$-relevant error directly.  The second is
uniform weights ($w_k = 1$ for all $k$), which simplifies the analysis.
Both produce valid phase boundaries; the choice affects the constants
$c_{\mathrm{S}}$ and $c_{\mathrm{D}}$ in the comparison-function
assumptions below but not the structural form of the phase boundary
theorem.  We leave $w_k$ as a free parameter and note that the
$\volL$-sensitivity choice is the natural default for a framework whose
objective is $\volL$ maximization.
\end{remark}

\subsection{Lyapunov dynamics}

The per-step behavior of $\Lyap$ depends on two comparison-function
assumptions that bound the correction and degradation signals.

\begin{assumption}[Coercivity of Self-Correction (C1)]
\label{ass:C1}
There exist constants $c_{\mathrm{S}} > 0$ and $c_{\mathrm{V}} \geq 0$
such that the aggregate correction signal satisfies
\begin{equation}
\label{eq:C1}
S(t) \geq \bigl(c_{\mathrm{S}} \cdot \rhomincross(P_t)
  + c_{\mathrm{V}} \cdot \nu(P_t, \Wm_t)\bigr) \cdot \Lyap(\Wm_t)
\end{equation}
whenever $\Lyap(\Wm_t) > 0$ and either $\rhomincross(P_t) > 0$ or
$\nu(P_t, \Wm_t) > 0$.  Here $\nu(P_t, \Wm_t) \geq 0$ is the
$\volL$-sensitivity to calibration improvement: the marginal $\volL$
gain from a unit reduction in $\Lyap$, evaluated at the current state
along the decision chain (better calibration $\to$ better decisions
$\to$ higher $\volL$).
\end{assumption}

The correction signal has two sources.  The first term,
$c_{\mathrm{S}} \cdot \rhomincross$, is \emph{exogenous}: observation
channels push the world model toward truth at a rate proportional to
channel strength and current error.  This follows from the EWMA learning
rule structure in the risk-trust model of~\citet{lasser2026scm}, with
additivity across cross-substrate channels under the channel
independence property of~\citet{lasser2026exo}.

The second term, $c_{\mathrm{V}} \cdot \nu$, is \emph{endogenous}: the
actor is rewarded for improving its own calibration because better
calibration produces better decisions, which preserve more cooperative
capabilities, which increase $\volL$.  The sensitivity $\nu$ measures
how much $\volL$ improves per unit of calibration improvement.  When
$\nu > 0$, the actor has a $\volL$-gradient incentive to seek
correction even beyond what the observation channels provide.  This
endogenous pressure is what drives the actor to seek exogenous
feedback: the actor \emph{wants} to be better calibrated because it can
estimate (through $\nu$) that better calibration produces higher
$\volL$.

Setting $c_{\mathrm{V}} = 0$ recovers the purely exogenous correction
model.  When $c_{\mathrm{V}} > 0$, the effective self-correction rate
is $\rSeff = c_{\mathrm{S}} \cdot \rhomincross +
c_{\mathrm{V}} \cdot \nu$, which is strictly larger than the exogenous
rate alone.  The consequences for the phase boundary are developed in
Section~\ref{sec:phase_boundary}
(Remark~\ref{rem:stabilizing_cascade}).

\begin{assumption}[Bounded Degradation (C2$'$)]
\label{ass:C2prime}
There exist constants $c_{\mathrm{D}} > 0$ and $\dmax > 0$ such that
for any step $t$ (subsumption or otherwise):
\begin{equation}
\label{eq:C2prime}
D(t) \leq c_{\mathrm{D}} \cdot \nsub(t) \cdot \Lyap(\Wm_t)
  + \dmax \cdot \Delta\rho(t)
\end{equation}
where $\nsub(t) \in \{0,1\}$ indicates whether a subsumption occurred
at step $t$, and
\begin{equation}
\label{eq:delta_rho}
\Delta\rho(t) = \sum_{k=1}^{K} w_k \cdot
  \max\!\left(0,\; \rhocross_k(P_t) - \rhocross_k(P_{t+1})\right)
\end{equation}
is the weighted redundancy lost at step $t$.
\end{assumption}

Assumption~C2$'$ contains two terms.  The first, $c_{\mathrm{D}} \cdot
\Lyap(\Wm_t)$, captures error-proportional growth: existing errors
compound faster without correction.  The second, $\dmax \cdot
\Delta\rho(t)$, captures the \emph{new} drift exposure from lost
channels.  This second term is the critical addition relative to the
simpler ``C2'' one might write without it.  A dimension with
$\epsilon_k \approx 0$ (well-calibrated \emph{because} a channel was
actively correcting it) starts drifting at rate $\dmax$ when its
channel is removed, even though $\Lyap(\Wm_t) \approx 0$ at the moment
of removal.  Without the $\Delta\rho$ term, the bound would predict
near-zero degradation precisely when subsumption is most
dangerous---removing a channel that was doing all the correction work
on a dimension that would otherwise drift.

\begin{proposition}[Error Dynamics]
\label{prop:error_dynamics}
Under the coupled $(P, \Wm)$ dynamics with locally rational transitions
(Definition~\ref{def:transition}), Assumptions~C1 and~C2$'$, and
$\alpha(t)$, $\beta(t)$ denoting the learning rate and degradation
coefficient respectively, the Lyapunov candidate satisfies two per-step
bounds.

\medskip
\noindent\textbf{Non-subsumption steps} ($\nsub(t) = 0$):
\begin{equation}
\label{eq:contraction}
\Lyap(\Wm_{t+1}) \leq \Lyap(\Wm_t) \cdot
  \left(1 - \rS \cdot \rhomincross(P_t)\right)
\end{equation}
where $\rS = \alpha(t) \cdot c_{\mathrm{S}}$ is the self-correction
rate.  This is a strict contraction provided $0 < \rS \cdot
\rhomincross(P_t) < 1$ (i.e., the step size is small enough that
correction does not overshoot).  We assume throughout that $\alpha(t)$
is chosen small enough to satisfy this bound; this is the standard
step-size condition for EWMA-style learning rules and holds whenever
$\alpha(t) \leq 1 / (c_{\mathrm{S}} \cdot \rhomincross(P_t))$.

\medskip
\noindent\textbf{Subsumption steps} ($\nsub(t) = 1$):
\begin{equation}
\label{eq:expansion}
\Lyap(\Wm_{t+1}) \leq \Lyap(\Wm_t) \cdot
  \left(1 - \rS \cdot \rhomincross(P_t) + \rW\right)
  + \dmax \cdot \Delta\rho(t)
\end{equation}
where $\rW = \beta(t) \cdot c_{\mathrm{D}}$ is the error-proportional
degradation rate.  On subsumption steps, the multiplicative factor may
exceed $1$ \emph{and} the additive kick may be nonzero.
\end{proposition}

\begin{proof}
The proof rests on the additive decomposition of the Lyapunov step:
\begin{equation}
\label{eq:lyap_decomposition}
\Lyap(\Wm_{t+1}) = \Lyap(\Wm_t) - \alpha(t) \cdot S(t)
  + \beta(t) \cdot D(t)
\end{equation}
where $S(t) \geq 0$ is the correction signal and $D(t) \geq 0$ is the
degradation signal.  This decomposition holds when the world-model
update separates into a correction component (driven by observation
channels) and a degradation component (driven by channel loss and
exogenous drift), with $\alpha(t)$ and $\beta(t)$ as their respective
coefficients.  The EWMA learning rules from the risk-trust model~\citep{lasser2026scm}
have exactly this structure: the update is linear in the residual (the
correction component) with a coefficient proportional to the learning
rate, and the drift component is separable by Lipschitz continuity of
$f_{\Wm}$ (Assumption~C2$'$).

For non-subsumption steps ($\nsub(t) = 0$), Assumption~C2$'$ gives
$D(t) \leq \dmax \cdot \Delta\rho(t) = 0$ (since $P_{t+1} = P_t$
when no subsumption occurs, $\Delta\rho(t) = 0$).  Applying
Assumption~C1:
\begin{align*}
\Lyap(\Wm_{t+1})
&\leq \Lyap(\Wm_t) - \alpha(t) \cdot c_{\mathrm{S}} \cdot
  \rhomincross(P_t) \cdot \Lyap(\Wm_t) \\
&= \Lyap(\Wm_t) \cdot \left(1 - \rS \cdot \rhomincross(P_t)\right).
\end{align*}

For subsumption steps ($\nsub(t) = 1$), apply both assumptions:
\begin{align*}
\Lyap(\Wm_{t+1})
&\leq \Lyap(\Wm_t)
  - \alpha(t) \cdot c_{\mathrm{S}} \cdot \rhomincross(P_t) \cdot
    \Lyap(\Wm_t)
  + \beta(t) \cdot c_{\mathrm{D}} \cdot \Lyap(\Wm_t)
  + \beta(t) \cdot \dmax \cdot \Delta\rho(t) \\
&= \Lyap(\Wm_t) \cdot
  \left(1 - \rS \cdot \rhomincross(P_t) + \rW\right)
  + \dmax \cdot \Delta\rho(t)
\end{align*}
where the last line uses the fact that C2$'$ bounds
$D(t) \leq c_{\mathrm{D}} \cdot \Lyap(\Wm_t) + \dmax \cdot
\Delta\rho(t)$ (with $\dmax$ already incorporating any environmental
degradation effects as a free constant).  The $\beta(t)$ coefficient
from the Lyapunov decomposition~\eqref{eq:lyap_decomposition} is
absorbed: $\beta(t) \cdot \dmax$ is bounded by $\dmax$ when
$\beta(t) \leq 1$ (which holds when the degradation coefficient is
normalized into $c_{\mathrm{D}}$ and $\dmax$ in C2$'$).  All
downstream uses of $\dmax$ refer to the constant from C2$'$ without
redefinition.
\end{proof}

\subsection{Cumulative error budget}

The per-step bound is no longer purely multiplicative (the additive
$\dmax \cdot \Delta\rho$ term prevents clean products of factors).
We therefore track the cumulative error budget over a window
$[t_0, t_0 + T]$:
\begin{equation}
\label{eq:error_budget}
B(t_0, T) =
  \underbrace{\sum_{t \in \mathrm{sub}} \dmax \cdot \Delta\rho(t)}
    _{\text{drift exposure}}
  + \underbrace{\sum_{t \in \mathrm{sub}} \rW \cdot \Lyap(\Wm_t)}
    _{\text{error-proportional degradation}}
  - \underbrace{\sum_{t=t_0}^{t_0+T-1} \rS \cdot \rhomincross(P_t) \cdot
    \Lyap(\Wm_t)}_{\text{cumulative correction}}
\end{equation}
where $\sum_{t \in \mathrm{sub}}$ ranges over the subsumption steps in
the window.

The first sum is total drift exposure from channel loss.  The second is
error-proportional degradation on subsumption steps.  The third is total
correction across all steps.  The system is net-degrading over
$[t_0, t_0 + T]$ when $B(t_0, T) > 0$.

\begin{remark}[Linearized approximation]
\label{rem:linearized}
When $\Lyap(\Wm_t)$ is approximately constant over the window and
$\Delta\rho$ is spread across $T_{\mathrm{sub}}$ subsumption steps, the
cumulative condition simplifies to
\begin{equation}
\label{eq:linearized}
\rS \cdot \rhomincross > \rW \cdot \rsub
  + \dmax \cdot \frac{\Delta\rho_{\mathrm{avg}}}{\Lyap_{\mathrm{avg}}}
\end{equation}
where $\rsub = T_{\mathrm{sub}} / T$ is the subsumption frequency and
$\Delta\rho_{\mathrm{avg}}$, $\Lyap_{\mathrm{avg}}$ are time-averaged
quantities.  The $\dmax$ term makes the condition harder to satisfy
when the rate of channel loss on well-calibrated dimensions is high
relative to current error---correctly diagnosing the failure mode where
removing a channel that was doing all the correction work appears safe
because current error is low.
\end{remark}

\begin{remark}[Why $\rhomincross$, not $\rho_{\min}$]
\label{rem:why_cross}
Proposition~\ref{prop:error_dynamics} uses cross-substrate redundancy
$\rhomincross$ rather than all-channel redundancy $\rho_{\min}$ because
the coercivity bound C1 requires correction signals to be additive
across channels.  Same-substrate channels may have correlated errors,
breaking additivity.  Cross-substrate channels satisfy two
distinct properties from~\citet{lasser2026exo}:
\begin{enumerate}
\item[(a)] \emph{Channel isolation} (unconditional): corruption on one
  substrate cannot propagate to channels on another substrate.
\item[(b)] \emph{Channel independence} (conditional on the
  joint-factorization assumption): channel outputs are statistically
  independent given the true state.
\end{enumerate}
Isolation prevents corruption propagation; independence makes correction
signals additive.  The Lyapunov bound uses additivity, which requires
independence, which requires joint factorization (termed ``joint
sufficiency'' in~\citet{lasser2026exo}).  If joint factorization fails (shared
environmental noise affects both substrates), correction signals may
be sub-additive, and the effective $\rhocross_k$ should be discounted
accordingly.  Using $\rhomincross$ makes the bound valid without
assuming same-substrate independence while still requiring
cross-substrate independence via the joint-factorization assumption.
\end{remark}

\begin{remark}
The cumulative-budget structure is the core of the paper.  The phase
boundary (Section~\ref{sec:phase_boundary}) is a condition on the
\emph{frequency and severity} of subsumption relative to the
inter-shock correction capacity, not a condition that must hold at every
individual step.  Subsumption events are discrete shocks; the system can
tolerate occasional shocks as long as the inter-shock correction is
sufficient.
\end{remark}
