\section{Unified Treatment of Three Feedback Loops}
\label{sec:unified_loops}

The three compound feedback loops identified in the gap analysis across
the prior papers in this sequence~\citep{lasser2026gfm,lasser2026poset,lasser2026horizon,lasser2026scm,lasser2026exo} are instances of the same $(P, \Wm)$ dynamical system with
different effective parameters.  This section maps each loop to the
formalism and shows how the phase boundary specializes.

\subsection{Loop 1: Structural avoidance $\to$ leverage blindness}

\paragraph{Mapping.}
The ``subsumption'' here is not agent elimination but \emph{structural
discovery avoidance}: the actor avoids learning poset structure
(subsumption edges) because discovery can contract $\volL$ in the short
term.

\begin{itemize}
\item $P_t$: the actor's \emph{known} poset (a subset of the true
  poset).
\item $\Wm_t$: leverage estimates $\hat{\lambda}(d, t)$, which are
  accurate only for the known poset.
\item ``Observation channels'' lost: each avoided structural discovery
  removes a potential correction to $\hat{\lambda}$.
\item $\rW$: the rate at which leverage estimates drift due to
  un-learned structure.
\item $\rS$: the rate at which the structural discovery
  value~\citep[Proposition~2, Definition~5]{lasser2026horizon} drives the actor toward
  learning structure.
\end{itemize}

\paragraph{Phase boundary specialization.}
Self-correction dominates when the information value $I(d_1 \leq d_2)$
\citep[Definition~5]{lasser2026horizon}) exceeds the immediate $\volL$ contraction from
discovery for a sufficient fraction of unknown edges.  The critical
ratio becomes:
\begin{equation}
\label{eq:loop1_boundary}
\min_{\text{undiscovered edges}}
  \frac{I(\text{edge})}{|\Delta\volL(\text{discover edge})|}
  > \text{threshold}
\end{equation}
where the threshold is determined by the degradation and drift
parameters from the general phase boundary.  This formalizes the
discussion of structural avoidance in~\citet{lasser2026poset} as a phase-boundary condition
rather than an informal concern.

\subsection{Loop 2: Partial subsumption $\to$ world-model homogenization}

This is the primary case, directly formalized in
Sections~\ref{sec:coupled_system}--\ref{sec:phase_boundary}.  The
parameters are as defined throughout: $P_t$ is the capability poset,
$\Wm_t$ is the full world-model accuracy vector, observation channels
are rooted in agents, and the phase boundary is stated in terms of
$(\rS, \rW, \rsub, \dmax, \rhomincross)$.

\subsection{Loop 3: Over-calibrated restriction $\to$ discovery avoidance}

\paragraph{Mapping.}
The ``subsumption'' is the actor \emph{restricting} capabilities rather
than eliminating agents.  But the dynamical structure is identical:
restriction prevents exercise, which prevents observation, which
prevents calibration of the risk estimates that justified the
restriction.

\begin{itemize}
\item $P_t$: the set of unrestricted capabilities (restriction reduces
  the exercisable poset).
\item $\Wm_t$: risk estimates $\hat{\Risk}(S, t)$, which are accurate
  only for exercised capabilities.
\item ``Observation channels'' lost: each restriction prevents the
  exercise events that would have generated Bayesian evidence about
  the risk claim (the verification asymmetry from
  \citet{lasser2026horizon}, Remark~2).
\item $\rW$: the rate at which risk estimates drift due to
  un-exercised pathways (this is exactly the Wamura problem from
  \citet{lasser2026horizon}).
\item $\rS$: the rate at which controlled relaxation protocols or
  natural exercise events provide counter-evidence.
\end{itemize}

\paragraph{Phase boundary specialization.}
Self-correction dominates when the controlled relaxation protocol
generates evidence faster than restriction accumulates unfalsifiable
risk claims.  This connects the present paper to the Wamura problem: a
controlled relaxation protocol is the \emph{mechanism} that makes $\rS$
positive for Loop~3, without which $\rS = 0$ and the absorbing basin is
the entire state space.

\begin{proposition}[Wamura Problem as Phase Boundary Degeneracy]
\label{prop:wamura}
In the absence of controlled relaxation ($\rS = 0$ for Loop~3) and
without endogenous calibration incentive on the risk dimension
($\alphamin \cdot c_{\mathrm{V}} \cdot \nu = 0$; see derivation
below), the self-correcting sufficient condition
(Equation~\ref{eq:self_correcting}) cannot be satisfied: the
correction term is identically zero.  Under the additional conditions
of Theorem~\ref{thm:phase_boundary}, Part~(b)---full blindness on the
risk dimension ($\rhocross_k \to 0$), environmental non-stationarity,
and Assumption~B1---every non-trivial risk restriction drives the
system toward absorption and maximal restriction.  Under B1$'$ (partial
indirect evidence), the outcome is metastable rather than absorbing
(Theorem~\ref{thm:phase_boundary}, Part~b$'$).
\end{proposition}

\begin{proof}
With $\rS = 0$ and $\alphamin \cdot c_{\mathrm{V}} \cdot \nu = 0$,
the cumulative error budget $B(t_0, T)$ becomes:

\emph{Why $c_{\mathrm{V}} \cdot \nu = 0$ on the risk dimension.}
The endogenous sensitivity $\nu(P_t, \Wm_t)$ measures the marginal
$\volL$ gain from a unit reduction in $\Lyap$, evaluated along the
decision chain: better calibration $\to$ better decisions $\to$ higher
$\volL$.  On the risk dimension of Loop~3, exercise is prohibited by
the restriction.  The actor therefore has no operational data from
which to estimate how improving its risk calibration would change its
decisions; the counterfactual ``what would happen if this capability
were exercised?'' is precisely what the restriction prevents observing.
Without an estimate of $\nu$, the actor cannot form the
$\volL$-gradient that drives endogenous correction.  Formally:
$\nu = \partial \volL / \partial \Lyap$ along the decision chain,
but the chain is severed at the restriction step (the capability is
never exercised, so the downstream $\volL$ consequences are
unobservable), giving $\nu = 0$ on the restricted dimension.
(As noted in Assumption~\ref{ass:C1}, $\nu$ is an objective
marginal sensitivity, but the endogenous correction operates through
the actor's gradient computation; when the chain is severed on a
dimension, $c_{\mathrm{V}} \cdot \nu = 0$ \emph{effectively} on that
dimension regardless of the objective value.)
\begin{equation*}
B(t_0, T) = \sum_{t \in \mathrm{sub}} \dmax \cdot \Delta\rho(t)
  + \sum_{t \in \mathrm{sub}} \rW \cdot \Lyap(\Wm_t)
\end{equation*}
Both sums contain only non-negative terms ($\dmax, \Delta\rho, \rW,
\Lyap \geq 0$), and the correction term vanishes entirely.
Therefore $B(t_0, T) \geq 0$ for every window, with equality only in
the trivial case where no restriction occurs.

For any non-trivial restriction step ($\nsub(t) = 1$ with
$\Delta\rho(t) > 0$ or $\Lyap(\Wm_t) > 0$), $B > 0$, satisfying
condition~(i) of Theorem~\ref{thm:phase_boundary}, Part~(b).
Conditions~(ii) and~(iii) follow from the restriction-specific
dynamics: each restriction step removes the exercise that would
generate evidence (no exercise $\to$ no evidence), driving
$\rhocross_k \to 0$ on the risk dimension (satisfying~(ii)).  With
$\rS = 0$ and the risk dimension environmentally non-stationary (the
true risk changes as the restricted capability's context evolves
without observation), the error $\epsilon_k$ grows, eventually
exceeding $\epsilon_{\mathrm{crit},k}$ (satisfying~(iii)).  Note that
$\dmax > 0$ bounds the drift rate from above (C2$'$); the cascade
requires positive drift, which is a structural consequence of
non-stationarity on the restricted dimension.  All three conditions of Part~(b) are
therefore met, and the cascade drives the system toward maximal
restriction.

This is the formal statement of why the Wamura problem was called
``the most important open problem'' in the discussion of~\citet{lasser2026horizon}: without a
mechanism to generate evidence against unfalsifiable risk claims, the
feedback loop between restriction and evidence suppression is
absorbing by default.
\end{proof}

\subsection{Summary}

Table~\ref{tab:loops} summarizes the three loops as parameter
instantiations of the unified system.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lp{4.5cm}p{4.5cm}@{}}
\toprule
\textbf{Loop} & \textbf{Correction ($\rS$)} & \textbf{Degradation ($\rW$)} \\
\midrule
1.\ Structural avoidance &
  Discovery value~\citep{lasser2026horizon} drives learning of unknown poset edges &
  Leverage estimates drift from un-learned structure \\[0.5em]
2.\ Partial subsumption &
  Self-trust $T_s$~\citep{lasser2026gfm} corrects world model via prediction residuals &
  Subsumption removes observation channels \\[0.5em]
3.\ Over-calibrated restriction &
  Controlled relaxation generates counter-evidence &
  Unfalsifiable risk claims accumulate under restriction \\
\bottomrule
\end{tabular}
\caption{Three feedback loops as instances of the $(P, \Wm)$ system.
  Each loop maps a correction source ($\rS$) and degradation source
  ($\rW$) onto the phase boundary of Theorem~\ref{thm:phase_boundary}.}
\label{tab:loops}
\end{table}

The unified treatment reveals that solutions for one loop improve the
phase boundary for all three.  Controlled relaxation (which makes
$\rS > 0$ for Loop~3) has no direct effect on Loop~2's parameters, but
the structural insight transfers: any mechanism that generates
observation-channel signal on dimensions that would otherwise go blind
is a contribution to $\rS$ in the relevant loop's phase boundary.
