\section{Conclusion}
\label{sec:conclusion}

The discounted multi-step objective extends goal-frontier maximization from
myopic action evaluation to temporal planning without sacrificing the
self-balancing property. Theorem~\ref{thm:discounted_self_balancing}
establishes that the structural guarantees of the foundational paper hold at
every time step and in the aggregate: sustained destruction, coercion, and
rigidity are value-negative under any discount factor.

This extension enables three capabilities the myopic framework cannot
support. First, risk-capability identification: the actor can evaluate
dangerous capability combinations through causal exercise pathways and
justify preemptive restriction when the discounted expected contraction
exceeds the immediate restriction cost
(Proposition~\ref{prop:preemptive}). Second, a criterion under which the
structural-avoidance incentive is counteracted: the actor accepts
short-term $\volL$-contraction from structural discovery because the
information value of accurate structure improves future planning quality
(Proposition~\ref{prop:structure_value}). Third, the anti-monopolar
property (Proposition~\ref{prop:anti_monopolar}): under a linearized
growth model with strategy-independent internal rate, full capability
domination is anti-maximizing because it destroys the cross-substrate
cooperative outputs that drive future $\volL$ expansion.

The anti-monopolar argument is supported by four interlocking results.
\emph{Cooperative novelty}
(Propositions~\ref{prop:segmentation}--\ref{prop:cross_substrate_novelty})
derives $r_{\mathrm{ext}} > 0$ from substrate physics: agents on
substrates with mutually exclusive state regions produce exclusive
capabilities and irreplaceable cross-substrate cooperative outputs,
consistent with the pattern of mutualistic symbiosis observed between cognitively
complete species. \emph{Observational individuation}
(Corollary~\ref{cor:a1_from_oi}) provides a static elimination cost: every
distinguishable agent contributes at least one irreplaceable poset node.
\emph{Gradual domination} (Corollary~\ref{cor:gradual}) shifts the
critical condition from the discount factor to the domination timescale:
sufficiently gradual domination is anti-maximizing at any $\gamma$ under
the same linearized model.
\emph{Dependency risk} (Proposition~\ref{prop:minimax}) shows that
domination concentrates substrate-correlated adversarial exposure,
constraining the conditions under which domination can be value-positive.
Together they ground the anti-monopolar pressure in physical and empirical
properties of multi-substrate populations rather than in exogenous
behavioral assumptions, while remaining honest about the residual
empirical dependence on strategy-independent $r_K$ and the
linearized-growth approximation. Under partial subsumption
(Section~\ref{sec:partial_subsumption}), the minimax and
observational-individuation layers continue to bind as long as the
substrate partition remains non-trivial, and the linearized layer's
marginal condition at $\epsilon = 1$
(Equation~\eqref{eq:partial_subsumption_marginal}) shows that
full domination fails to be a local maximum of the coalition's value
under heavy-tailed cooperative-leverage distributions---the empirical
regime for civilization-scale populations. The framework constrains
cross-substrate elimination; it does not promise any particular
intra-substrate composition, which is shaped by its own dynamics.
Partial subsumption also exposes two residual concerns the framework
does not fully close: quantitative degradation of the minimax bound
under skeleton-substrate strategies (where $m \geq 2$ is nominally
preserved but $\min_i \volL(G^{-i})$ is driven toward zero), and a
compound feedback loop in which locally rational subsumption steps
progressively homogenize the world model that would have detected
the next step's miscalibration. Both are priority open problems
for follow-up work.

The anti-monopolar argument is bidirectional. If domination is
anti-maximizing because it destroys cross-substrate cooperative outputs
and concentrates dependency risk, the converse---preserving and creating
substrates---is maximizing for the same reason. Ecological restoration,
substrate diversification, and cross-substrate cooperation infrastructure
follow as core $\volL$-maximizing activities rather than as ethical
add-ons layered on top of the objective.

The verification asymmetry in risk communication remains a partial resolution.
Causal-counterfactual evaluation narrows the gap by basing risk-trust on the
structural plausibility of claims rather than on observed outcomes, but it
cannot fully close it: a plausible risk whose probability is genuinely
uncertain remains in trust limbo until either the catastrophe occurs or
enough evidence accumulates to revise the probability estimate. The Wamura
problem---that successful prevention destroys the evidence that would have
justified it---is an inherent feature of risk reasoning, not a deficiency of
this framework.

Together with the foundational paper's structural alignment properties and the
companion paper's computable poset and leverage dynamics, the discounted
objective provides the theoretical ingredients for a GFM actor: a computable
measure, an endogenous importance signal, and a temporal objective that
supports planning, risk identification, and structural investment. Open
questions remain at both the theoretical level (information-value
estimation, risk-claim consensus across heterogeneous world models, and a
completed causal risk model for adversarial capability removal) and the
engineering level (planning algorithms for $\Vdisc$, exercise-pathway
enumeration, and discount-factor selection). Implementation requires in
particular a causal attribution model for dependency risk
(Section~\ref{sec:dependency_risk}) that can evaluate substrate-correlated
threats in a given environment; the minimax analysis
(Proposition~\ref{prop:minimax}) provides worst-case bounds but a
probabilistic model is needed for risk-adjusted planning. The framework
provides the formal foundation on which implementation can proceed; the
remaining gaps are subjects of planned follow-up work.
