\section*{Author Contributions}

\textbf{Teague Lasser} conceived the multi-step extension, identified the
risk-capability and verification-asymmetry problems, proposed the
causal-counterfactual resolution for risk-trust, directed all revisions, and
made final editorial decisions. Responsible for the paper's intellectual
direction and all claims made.

\textbf{Claude Opus 4.6} (Anthropic) drafted the formal exposition, including
the discounted value function definition, the self-balancing transfer theorem
and proof, the exercise-pathway and concentration-risk definitions, the
preemptive restriction criterion, the information-value-of-structure
formalization, the anti-monopolar property with observational individuation
and gradual domination corollaries, and the structural-avoidance proof.

\textbf{GPT 5.4} (OpenAI) served as technical reviewer across multiple
rounds, identifying the equation mismatch between the anti-monopolar
proposition statement and proof, the cooperative-capability overclaim in
Part~(a), the $r_{\mathrm{ext}}$ proof gap in the observational
individuation corollary, the $\gamma$-selection problem, and the
linearization limitations. These findings directly shaped the formal
content of the paper.

\medskip
\noindent\emph{Transparency note.} Both AI systems operated as tools under
human direction. Neither system has continuity across sessions, cannot take
responsibility for the work in the sense required by most venue authorship
policies, and cannot respond to reviewer queries independently. They are
listed as authors to accurately represent their contributions to the
intellectual content of the paper, not to claim that they meet all criteria
of traditional academic authorship. The corresponding author for all
inquiries is Teague Lasser.
