\section{Introduction}
\label{sec:introduction}

Goal-frontier maximization (GFM) proposes alignment through a single
objective: maximize $\volL(G)$, the volume of the jointly achievable
capability space \citep{lasser2026gfm}. The foundational paper proves that
this objective penalizes destruction, coercion, and rigidity from the same
measure-monotonicity argument. Its companion \citep{lasser2026poset}
constructs a computable capability poset, defines leverage as an endogenous
importance signal, and argues that the objective produces a gradient toward
civilization-scale infrastructure.

Both papers optimize a \emph{myopic} objective: the actor evaluates each
action by its immediate effect on $\volL$ and selects the one with the
highest estimated expansion. This one-step lookahead is sufficient for
detecting direct harm, coercion, and capability expansion, but it cannot
handle two classes of problems that require reasoning about the future.

\paragraph{Risk capabilities.} Some capability combinations are individually
$\volL$-expanding but collectively create catastrophic contraction potential
\citep{bostrom2013existential, ord2020precipice}.
``Nuclear physics'' ($w = 10$ bits), ``uranium enrichment'' ($w = 8$ bits),
and ``precision machining'' ($w = 9$ bits) each expand the capability space
when acquired. But their co-occurrence in a single agent with low trust
creates an exercise pathway to massive $\volL$-contraction. The myopic
estimator $\Deltahat(\pi)$ evaluates each acquisition as positive; it cannot
see the future contraction that the combination enables.

\paragraph{Structural avoidance.} The companion paper identifies a perverse
incentive: discovering subsumption structure can decrease $\volL$
(Proposition~5(c) of \citet{lasser2026poset}), so a myopic $\volL$-maximizer
is rewarded for staying structurally ignorant. A multi-step optimizer can
accept the short-term contraction because accurate structure improves the
quality of future decisions.

This paper extends GFM to multi-step planning. We introduce a discounted
value function $\Vdisc(\pi) = \mathbb{E}[\sum_t \gamma^t \Delta \volL(G_t)
\mid \pi]$ and prove that the self-balancing property transfers to the
discounted sum. On this foundation, we develop risk-capability identification
through causal counterfactual models and analyze the verification asymmetry
in risk communication: successful prevention destroys the evidence that would
have justified it.

\paragraph{Contributions.}
\begin{enumerate}
\item A discounted multi-step objective for GFM that preserves the
  self-balancing property at every time step and in the aggregate
  (Section~\ref{sec:discounted}).
\item Risk-capability identification via causal exercise pathways, with
  a formal criterion for when preemptive restriction is value-positive
  (Section~\ref{sec:risk}).
\item Analysis of the verification asymmetry in risk communication, with
  a causal-counterfactual resolution for risk-trust convergence
  (Section~\ref{sec:communication}).
\item A criterion under which the discounted objective counteracts the
  structural-avoidance incentive from the companion paper
  (Section~\ref{sec:avoidance}).
\item The anti-monopolar property: under a diversity condition, full
  capability domination is anti-maximizing under the discounted objective,
  creating endogenous pressure against the singleton scenario
  (Section~\ref{sec:substitution}).
\end{enumerate}
